---
title: "Creating New Explanatory Variables"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
set.seed(12)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(1)
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
```

# Learning Objectives

- Create new explanatory variables to handle curvature (polynomial regression),
  categorical predictors (indicator / one-hot variables), and non-constant
  slopes (interaction terms).
- Chapter 9.

---

# Adding Curvature

```{r}
#| label: corn-scatter
#| message: false
ex0915 <- read_csv("https://dcgerard.github.io/stat_302/data/ex0915.csv")
ggplot(ex0915, aes(x = Rainfall, y = Yield)) +
  geom_point()
```

## Quadratic Regression **is** Multiple Linear Regression

- From last chapter, we said that we should fit
$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + \epsilon_i$$
- Relabel $X_{1i} = X_i$ and $X_{2i} = X_i^2$.
- Then this is equivalent to fitting
$$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \epsilon_i$$

## Why Is It Called *Linear* Regression?

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + \epsilon_i$$

- Multiple linear regression represents the mean of the $Y$'s as a linear combination of the $\beta$'s.
- Even though the mean is a quadratic function of the $X_i$'s, it is still a linear function of the $\beta_j$'s.

To fit a polynomial, create new variables that are powers of existing variables, then include those in the multiple regression model.

```{r}
#| label: quad-fit
ex0915 <- ex0915 |>
  mutate(Rainfall2 = Rainfall ^ 2)
lmout_quad <- lm(Yield ~ Rainfall + Rainfall2, data = ex0915)
lmout_quad
```

| `(Intercept)` | `Rainfall` | `Rainfall2` |
|:-:|:-:|:-:|
| $\hat\beta_0$ | $\hat\beta_1$ | $\hat\beta_2$ |

Estimated model:
$$\hat\mu(Y \mid \text{Rainfall}) = -5.0 + 6.0\,\text{Rainfall} - 0.2\,\text{Rainfall}^2$$

-   I think it works better to wrap a transformation within `I()` (for "As**i**s"). 
    - This allows `predict()` to work with just `Rainfall`
  
    ```{r}
    lmout_quad_2 <- lm(Yield ~ Rainfall + I(Rainfall^2), data = ex0915)
    lmout_quad_2
    ```

-   Look at how we encode `predict()`:
    ```{r}
    predict(object = lmout_quad, newdata = data.frame(Rainfall = 10, Rainfall2 = 100))
    predict(object = lmout_quad_2, newdata = data.frame(Rainfall = 10))
    ```

---

# Indicator Variables

- An **indicator variable** only takes on the values 0 or 1.
- If you have a binary explanatory variable, you can represent it as an indicator variable and include it directly in the regression model.
- The result is effectively **two lines that have the same slope but different intercepts** (parallel lines).

Researchers studied the effect of timing and light intensity on flower yield:

- Response: Flower yield (average number of flowers per plant).
- Explanatory variables: Timing of light (early/late, binary), intensity of light (quantitative).

```{r}
#| label: flowers-head
#| message: false
case0901 <- read_csv("https://dcgerard.github.io/stat_302/data/case0901.csv")
glimpse(case0901)
```

Model: $\mu(\text{Flowers} \mid \text{Time},\,\text{Intensity}) = \beta_0 + \beta_1\,\text{Time} + \beta_2\,\text{Intensity}$

- `Time` can be made into an indicator variable (because it only has two levels).
- The model at $\text{Time} = 0$: $\beta_0 + \beta_2\,\text{Intensity}$
- The model at $\text{Time} = 1$: $(\beta_0 + \beta_1) + \beta_2\,\text{Intensity}$
- Slope is $\beta_2$ both times, but the lines have different intercepts (parallel lines).

```{r}
#| label: flowers-scatter
#| echo: false
case0901 <- case0901 |>
  mutate(Time = as.factor(Time))
ggplot(case0901, aes(x = Intensity, y = Flowers, color = Time)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

---

# One-Hot Transformation

- You can represent any categorical variable with $k$ levels using $k - 1$ indicator variables.
- This representation is called a **one-hot transformation** in the machine learning community.
- Let $X_{\ell i} = 1$ if unit $i$ belongs to level $\ell$, and $X_{\ell i} = 0$ otherwise.

Let $Z$ be a categorical variable with levels "Bob", "Cindy", "Doug":

- $X_{1i} = 1$ if Cindy and 0 otherwise.
- $X_{2i} = 1$ if Doug and 0 otherwise.
- Whenever "Bob": $X_{1i} = 0$ and $X_{2i} = 0$.
- Whenever "Cindy": $X_{1i} = 1$ and $X_{2i} = 0$.
- Whenever "Doug": $X_{1i} = 0$ and $X_{2i} = 1$.

If we have a quantitative response and a categorical explanatory variable, **we can apply a one-hot transformation and use multiple linear regression.**

$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \epsilon_i$

- Mean if "Bob":   $\beta_0 + \beta_1 \cdot 0 + \beta_2 \cdot 0 = \beta_0$
- Mean if "Cindy": $\beta_0 + \beta_1 \cdot 1 + \beta_2 \cdot 0 = \beta_0 + \beta_1$
- Mean if "Doug":  $\beta_0 + \beta_1 \cdot 0 + \beta_2 \cdot 1 = \beta_0 + \beta_2$

This is equivalent to one-way ANOVA:

- Multiple regression: $Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \epsilon_i$
- One-way ANOVA: $Y_{ij} = \mu + \alpha_i + \epsilon_{ij}$
- $\mu = \beta_0$, $\alpha_2 = \beta_1$, $\alpha_3 = \beta_2$
- In linear regression, $X_{1i}$ and $X_{2i}$ index the group status of unit $i$.
- In ANOVA, $i$ indexes the group, $j$ indexes the unit within the group.

If a variable has only 2 levels, it can be represented by 1 indicator variable. For example, `Time` has levels `Late` and `Early`:

- Let $X_{1i} = 0$ if Late and $X_{1i} = 1$ if Early.

## Including Categorical Variables in R

- If the variable is a `factor`, R will automatically apply a one-hot transformation.
- Check with `class()`:

```{r}
#| label: class-check
class(case0901$Time)
```

- If it is not a `factor`, convert it with `as.factor()`:

```{r}
#| label: as-factor
case0901 <- case0901 |>
  mutate(Time = as.factor(Time))
```

```{r}
#| label: lm-flowers
lmout <- lm(Flowers ~ Time + Intensity, data = case0901)
coef(lmout)
```

Model: $\mu(Y_i \mid \text{Time},\,\text{Intensity}) = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i}$, where $X_{1i} = 1$ if `Time` is `Late` and 0 otherwise, and $X_{2i}$ is light intensity:

| `(Intercept)` | `Time2` | `Intensity` |
|:-:|:-:|:-:|
| $\hat\beta_0$ | $\hat\beta_1$ | $\hat\beta_2$ |

---

# Interactions

- An **interaction** between two variables means the slope with respect to one variable **changes** with the value of the other.
- $\mu(Y_i \mid \text{Time},\,\text{Intensity}) = \beta_0 + \beta_1\,\text{Time} + \beta_2\,\text{Intensity} + \beta_3\,\text{Time} \times \text{Intensity}$

When $\text{Time} = 0$:

$$\mu(Y_i \mid \text{Time} = 0,\,\text{Intensity}) = \beta_0 + \beta_2\,\text{Intensity}$$

When $\text{Time} = 1$:

$$\mu(Y_i \mid \text{Time} = 1,\,\text{Intensity}) = (\beta_0 + \beta_1) + (\beta_2 + \beta_3)\,\text{Intensity}$$

- Slope when $\text{Time} = 0$: $\beta_2$
- Slope when $\text{Time} = 1$: $\beta_2 + \beta_3$
- Intercept when $\text{Time} = 0$: $\beta_0$
- Intercept when $\text{Time} = 1$: $\beta_0 + \beta_1$
- Each level of `Time` has its own line.

**No interaction** — parallel lines, same slope:

```{r}
#| label: no-interaction
lmadd <- lm(Flowers ~ Time + Intensity, data = case0901)
case0901 <- case0901 |>
  mutate(fit = fitted(lmadd))
ggplot(case0901, aes(x = Intensity, y = Flowers, color = Time)) +
  geom_line(aes(y = fit), linewidth = 1)
```

**Interaction** — non-parallel lines, slopes differ by group:

```{r}
#| label: interaction-plot
case0901 <- case0901 |>
  mutate(fit1 = fit + Intensity * (Time == 2) * 0.02)
ggplot(case0901, aes(x = Intensity, y = Flowers, color = Time)) +
  geom_line(aes(y = fit1), linewidth = 1)
```

## Fitting Interactions in R

$\mu(Y_i \mid \text{Time},\,\text{Intensity}) = \beta_0 + \beta_1\,\text{Time} + \beta_2\,\text{Intensity} + \beta_3\,\text{Time} \times \text{Intensity}$

```{r}
#| label: lm-interaction
lmint <- lm(Flowers ~ Time * Intensity, data = case0901)
coef(lmint)
```

| `(Intercept)` | `Time2` | `Intensity` | `Time2:Intensity` |
|:-:|:-:|:-:|:-:|
| $\hat\beta_0$ | $\hat\beta_1$ | $\hat\beta_2$ | $\hat\beta_3$ |

- Using `*` fits interactions along with all lower-order terms.
- Using `:` just fits interactions.

Reconsider the brain weight data:

$\mu(\text{Brain} \mid \text{Body},\,\text{Litter}) = \beta_0 + \beta_1\,\text{Body} + \beta_2\,\text{Litter} + \beta_3\,\text{Body} \times \text{Litter}$

- What is the slope for `Body` at a given `Litter` size?
- What is the intercept for `Body` at a given `Litter` size?

Interpreting models with interactions is difficult. Include them only when
there is a substantive reason to believe the slope changes across levels of the
moderating variable.
