---
title: "Multiple Regression"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
set.seed(14)
library(patchwork)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(1)
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(broom)
```

# Learning Objectives

- Introduce Multiple Linear Regression.
- Interpret the coefficients of a multiple regression model in randomized
  experiments and observational studies.
- Fit multiple linear regression models in R using `lm()`.
- Conduct inference ($t$-tests and confidence intervals) for individual
  coefficients.
- Understand how $p$-value interpretation depends on what other variables are
  in the model.
- Chapters 9 and 10 in the book.

---

# Case Study: Brain Size

- What variables are associated with brain weight?
- Collected information on 96 different species.
- We know that body weight is already associated with brain weight,
    - So what variables are associated with brain weight **after controlling for body weight**?
- Possible variables: Body weight (kg), gestation period (days), litter size.

```{r}
#| label: brain-data
#| message: false
case0902 <- read_csv("https://dcgerard.github.io/stat_302/data/case0902.csv")
glimpse(case0902)
```

---

# The Model

Simple linear regression:

- One quantitative response variable ($Y$).
- One quantitative explanatory variable ($X$).
- Model the conditional distribution of $Y$ given $X$.
- $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$

```{r}
#| label: slr-scatter
#| message: false
ggplot(case0902, aes(x = log(Body), y = log(Brain))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Multiple linear regression:

- One quantitative response ($Y$).
- **Multiple** quantitative explanatory variables ($X_1, X_2, \ldots, X_p$).
- $Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_p X_{pi} + \epsilon_i$
- E.g., $X_{2i}$ is the value of the second explanatory variable for unit $i$.
- With two explanatory variables:
$$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \epsilon_i$$
- $\epsilon_i$ is still ideally normally distributed with mean 0 and constant variance $\sigma^2$.

---

# Interpreting Coefficients

Consider the model $Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \epsilon_i$.

To interpret $\beta_1$ in a **randomized experiment**:

- Conceptually fix $X_{2i}$ at a value.
- Add one to $X_{1i}$.
- $Y_i$ changes by $\beta_1$.
- $\beta_1$ is how much $Y_i$ increases when we add one to $X_{1i}$ but keep $X_{2i}$ fixed.
- "A one-unit increase in light intensity causes the mean number of flowers to increase by $\beta_1$."

To interpret $\beta_1$ in an **observational study**:

- The $X_i$'s cannot be fixed independently of one another.
- Consider a subpopulation that has the same values of the $X_j$'s, where $j \neq 1$. Then the expected difference in means between species that differ in $X_1$ only by one unit is $\beta_1$.
- $\beta_1$ is the expected difference in $Y$'s when we compare units with $X_1$ and $X_1 + 1$.
- "For any subpopulation of mammal species with the same body weight, species with a one-day longer gestation length have a mean brain weight $\beta_1$ grams larger."

## A Useful Notation

- $\mu(\text{brain} \mid \text{gestation}) = \beta_0 + \beta_1\,\text{gestation}$
    - The mean of brain weight equals $\beta_0$ plus $\beta_1$ times the gestation time.
- $\mu(\text{brain} \mid \text{gestation},\,\text{body}) = \beta_0 + \beta_1\,\text{gestation} + \beta_2\,\text{body}$
    - The mean of brain weight equals $\beta_0$ plus $\beta_1$ times the gestation time plus $\beta_2$ times the body weight.

## Interpretation Changes with the Model

The interpretation of $\beta_1$ **changes when the model changes**, because the comparison group changes:

- $\mu(\text{brain} \mid \text{gestation}) = \beta_0 + \beta_1\,\text{gestation}$
    - $\beta_1$ is the mean difference in brain weight comparing gestation periods 1 day apart **in the population of all mammal species**.
- $\mu(\text{brain} \mid \text{gestation},\,\text{body}) = \beta_0 + \beta_1\,\text{gestation} + \beta_2\,\text{body}$
    - $\beta_1$ is the mean difference comparing gestation periods 1 day apart **in subpopulations with the same body weight**.

## Visualizing the Coefficient Interpretation

Consider artificial data with $\beta_0 = 2$, $\beta_1 = 1$ (positive $X_1$ effect), $\beta_2 = -3$ (large negative $X_2$ effect), where $X_1$ and $X_2$ are positively correlated.

```{r}
#| label: fake-data
beta0 <- 2; beta1 <- 1; beta2 <- -3
x2 <- rep(1:10, 5)
x1 <- rep(0:9, 5) + rep(1:5, each = 10)
y  <- beta0 + beta1 * x1 + beta2 * x2
df_fake <- data.frame(x1 = x1, y = y, x2 = as.factor(x2))
```

**Without $X_2$ in the model** — the marginal slope looks **negative** even though $\beta_1 = 1$:

```{r}
#| label: fake-no-x2
#| message: false
ggplot(df_fake, aes(x = x1, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

**With $X_2$ color-coded** — the slope within each level of $X_2$ is positive ($= \beta_1 = 1$), and the parallel lines sit at different heights:

```{r}
#| label: fake-with-x2
#| message: false
ggplot(df_fake, aes(x = x1, y = y, color = x2)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  guides(color = guide_legend(keywidth = 0.1, keyheight = 0.1,
                               default.unit = "inch"))
```

The figure below shows the same idea using the real brain-size data (`case0902`):

- **Left:** a single marginal fit line for all 96 species ($\hat\beta_1 \approx 2.23$).
- **Middle:** points colored by log(Body) tertile — large-bodied species (green) drive much of the steep marginal slope because they have both long gestations and large brains.
- **Right:** parallel fit lines from the additive model — every body-weight group has the same gestation slope ($\hat\beta_1 \approx 0.67$) but sits at a different height, reflecting the body-size effect $\beta_2$.

```{r}
#| label: parallel-slopes
#| echo: false
#| fig-width: 11
#| fig-height: 3.5
#| message: false
df <- case0902 |>
  mutate(
    logBrain     = log(Brain),
    logGestation = log(Gestation),
    logBody      = log(Body),
    BodyGroup    = cut(log(Body),
                       breaks         = quantile(log(Body), c(0, 1/3, 2/3, 1)),
                       labels         = c("Small", "Medium", "Large"),
                       include.lowest = TRUE)
  )

lm_marg <- lm(logBrain ~ logGestation, data = df)
lm_add  <- lm(logBrain ~ logGestation + BodyGroup, data = df)

b1_marg <- round(coef(lm_marg)["logGestation"], 2)
b1_part <- round(coef(lm_add)["logGestation"],  2)

x_seq <- seq(min(df$logGestation), max(df$logGestation), length.out = 200)

marg_line <- tibble(
  logGestation = x_seq,
  logBrain_hat = predict(lm_marg, data.frame(logGestation = x_seq))
)

group_lines <- expand_grid(logGestation = x_seq, BodyGroup = levels(df$BodyGroup)) |>
  mutate(logBrain_hat = predict(lm_add, data.frame(logGestation = logGestation,
                                                    BodyGroup = BodyGroup)))

cb_pal <- c("Small" = "#E69F00", "Medium" = "#56B4E9", "Large" = "#009E73")

p1 <- ggplot(df, aes(logGestation, logBrain)) +
  geom_point(color = "gray50", size = 1.5) +
  geom_line(data = marg_line, aes(y = logBrain_hat),
            color = "#0066cc", linewidth = 1.3) +
  labs(x = "log(Gestation, days)", y = "log(Brain, g)",
       subtitle = bquote(hat(beta)[1] == .(b1_marg) ~ "(marginal, all species)"))

p2 <- ggplot(df, aes(logGestation, logBrain, color = BodyGroup)) +
  geom_point(size = 1.5) +
  geom_line(data = marg_line, aes(logGestation, logBrain_hat),
            inherit.aes = FALSE, color = "black",
            linewidth = 1, linetype = "dashed") +
  scale_color_manual(values = cb_pal, name = "Body weight") +
  labs(x = "log(Gestation, days)", y = NULL,
       subtitle = "Body weight confounds the relationship")

p3 <- ggplot(df, aes(logGestation, logBrain, color = BodyGroup)) +
  geom_point(size = 1.5) +
  geom_line(data = group_lines, aes(logGestation, logBrain_hat, color = BodyGroup),
            linewidth = 1.3) +
  scale_color_manual(values = cb_pal, name = "Body weight") +
  labs(x = "log(Gestation, days)", y = NULL,
       subtitle = bquote(hat(beta)[1] == .(b1_part) ~
                           "(within groups — same slope, different heights)"))

(p1 | p2 | p3) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

---

# Fitting Multiple Linear Regression in R

Want to fit: $\mu(\text{brain} \mid \text{gestation},\,\text{body}) = \beta_0 + \beta_1\,\text{gestation} + \beta_2\,\text{body}$

- $\beta_0$, $\beta_1$, $\beta_2$ are parameters (we don't know them).
- We can **estimate** them by minimizing the sum of squared residuals.
- Residuals: $Y_i - (\beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i})$
- The resulting estimates are the OLS (ordinary least squares) estimates: $\hat\beta_0$, $\hat\beta_1$, $\hat\beta_2$.

Use `lm()` and **always save the output**:

```{r}
#| label: lm-two-vars
lmout <- lm(Brain ~ Gestation + Body, data = case0902)
lmout
```

| `(Intercept)` | `Gestation` | `Body` |
|:-:|:-:|:-:|
| $\hat\beta_0$ | $\hat\beta_1$ | $\hat\beta_2$ |

```{r}
#| label: lm-three-vars
lmout <- lm(Brain ~ Gestation + Body + Litter, data = case0902)
lmout
```

| `(Intercept)` | `Gestation` | `Body` | `Litter` |
|:-:|:-:|:-:|:-:|
| $\hat\beta_0$ | $\hat\beta_1$ | $\hat\beta_2$ | $\hat\beta_3$ |

We would interpret this output as:

- "We estimate that species with a 1-day longer gestation time have a mean brain weight 1.8 grams heavier, after adjusting for body weight and litter size."
- "We estimate that species with an average body weight 1 kg heavier have a mean brain weight 0.99 grams heavier, after adjusting for gestation time and litter size."
- "We estimate that species with a litter size one offspring larger have a mean brain weight 27.6 grams heavier, after adjusting for body weight and gestation time."

---

# Inference

- We are usually interested in testing if $\beta_i = 0$.
- We are usually interested in getting confidence intervals on the $\beta_i$'s.
- We can use the usual $t$-tools to get these.

```{r}
#| label: inference-tidy
lmout <- lm(Brain ~ Gestation + Body, data = case0902)
tidy(lmout)
```

| `term` | `estimate` | `std.error` | `statistic` | `p.value` |
|:--|:-:|:-:|:-:|:-:|
| `(Intercept)` | $\hat\beta_0$ | $SE(\hat\beta_0)$ | $\hat\beta_0/SE(\hat\beta_0)$ | p-value for $H_0:\beta_0=0$ |
| `Gestation`   | $\hat\beta_1$ | $SE(\hat\beta_1)$ | $\hat\beta_1/SE(\hat\beta_1)$ | p-value for $H_0:\beta_1=0$ |
| `Body`        | $\hat\beta_2$ | $SE(\hat\beta_2)$ | $\hat\beta_2/SE(\hat\beta_2)$ | p-value for $H_0:\beta_2=0$ |

```{r}
#| label: inference-glance
glance(lmout)
```

```{r}
#| label: confint-output
tidy(lmout, conf.int = TRUE)
```

## Interpreting $p$-values

The interpretation of significance **depends on what other variables are in the model**.

$\mu(\text{brain} \mid \text{gestation}) = \beta_0 + \beta_1\,\text{gestation}$

- $H_0: \beta_1 = 0$ vs $H_A: \beta_1 \neq 0$
- Model under $H_0$: $\mu(\text{brain}) = \beta_0$
- Model under $H_A$: $\mu(\text{brain} \mid \text{gestation}) = \beta_0 + \beta_1\,\text{gestation}$
- If we reject $H_0$: "We have strong evidence that gestation is related to brain weight."
- If we fail to reject $H_0$: "We do not have strong evidence that gestation is related to brain weight."

$\mu(\text{brain} \mid \text{gestation},\,\text{body}) = \beta_0 + \beta_1\,\text{gestation} + \beta_2\,\text{body}$

- $H_0: \beta_1 = 0$ vs $H_A: \beta_1 \neq 0$
- Model under $H_0$: $\mu(\text{brain} \mid \text{body}) = \beta_0 + \beta_2\,\text{body}$
- Model under $H_A$: $\mu(\text{brain} \mid \text{gestation},\,\text{body}) = \beta_0 + \beta_1\,\text{gestation} + \beta_2\,\text{body}$
- If we reject $H_0$: "We have strong evidence that gestation is related to brain weight after adjusting for body size."
- If we fail to reject $H_0$: "We do not have strong evidence that gestation is related to brain weight after adjusting for body size."
