---
title: "Multiple Regression EDA"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(Sleuth3)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(1)
```

# Learning Objectives

- Apply a systematic strategy for exploratory data analysis in multiple linear
  regression.
- Chapter 9.

---

# Case Study: Brain Size

- What variables are associated with brain weight?
- Collected information on 96 different species.
- We know that body weight is already associated with brain weight,
    - So what variables are associated with brain weight **after controlling for body weight**?
- Possible variables: Body weight (kg), gestation period (days), litter size.

```{r}
#| label: brain-data
data("case0902")
head(case0902)
```

---

# Plotting the Response Against Each Explanatory Variable

- The first step is almost always **making a ton of scatterplots**.
- Plots of the response against each explanatory variable show what variables seem to be **marginally** related to the response.
- "Marginally related" = related **unconditionally** on other variables.

```{r}
#| label: scatter-body
ggplot(case0902, aes(x = Body, y = Brain)) +
  geom_point()
```

```{r}
#| label: scatter-gestation
ggplot(case0902, aes(x = Gestation, y = Brain)) +
  geom_point()
```

```{r}
#| label: scatter-litter
ggplot(case0902, aes(x = Litter, y = Brain)) +
  geom_point()
```

Curvature and different spreads at each explanatory variable suggest a log transformation of `Brain`.

```{r}
#| label: log-brain
case0902$logBrain <- log(case0902$Brain)
```

```{r}
#| label: scatter-logbrain-body
ggplot(case0902, aes(x = Body, y = logBrain)) +
  geom_point()
```

```{r}
#| label: scatter-logbrain-gestation
ggplot(case0902, aes(x = Gestation, y = logBrain)) +
  geom_point()
```

```{r}
#| label: scatter-logbrain-litter
ggplot(case0902, aes(x = Litter, y = logBrain)) +
  geom_point()
```

Still a lot of curvature, so it looks like logging each variable might help.

```{r}
#| label: log-all-vars
case0902$logBody      <- log(case0902$Body)
case0902$logGestation <- log(case0902$Gestation)
case0902$logLitter    <- log(case0902$Litter)
```

```{r}
#| label: scatter-log-body
ggplot(case0902, aes(x = logBody, y = logBrain)) +
  geom_point()
```

```{r}
#| label: scatter-log-gestation
ggplot(case0902, aes(x = logGestation, y = logBrain)) +
  geom_point()
```

```{r}
#| label: scatter-log-litter
ggplot(case0902, aes(x = logLitter, y = logBrain)) +
  geom_point()
```

There looks like there is a lot of linearity now.

---

# Plotting Explanatory Variables Against Each Other

- It is often useful to look at scatterplots between each pair of explanatory variables.
- This tells us if some variables seem to be picking up a lot of the same information.
- E.g., gestation period might be larger just because body size is larger.
    - If brain is associated with gestation, it might only be through body weight.

```{r}
#| label: scatter-body-gestation
ggplot(case0902, aes(x = logBody, y = logGestation)) +
  geom_point()
```

```{r}
#| label: scatter-body-litter
ggplot(case0902, aes(x = logBody, y = logLitter)) +
  geom_point()
```

```{r}
#| label: scatter-gestation-litter
ggplot(case0902, aes(x = logGestation, y = logLitter)) +
  geom_point()
```

---

# Matrix Plot

- You can show scatterplots between all variables at the same time.
- This is called a **matrix plot** (or a **pairs plot**).

```{r}
#| label: pairs-plot
#| message: false
#| fig-width: 7
#| fig-height: 7
library(GGally)
ggpairs(case0902, columns = 6:9)
```

---

# Initial Model Fit

- We can first fit a very complicated model and check residuals.
- We would be looking for more curvature and outliers.

```{r}
#| label: complex-model
lm_comp <- lm(logBrain ~ logBody * logLitter * logGestation,
              data = case0902)
res_vec <- resid(lm_comp)
fit_vec <- fitted(lm_comp)
```

```{r}
#| label: resid-plot
ggplot(data.frame(fit_vec, res_vec), aes(x = fit_vec, y = res_vec)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

```{r}
#| label: qq-plot
ggplot(data.frame(res_vec), aes(sample = res_vec)) +
  geom_qq() +
  geom_qq_line()
```

---

# Coded Scatterplots

## Quantitative Third Variable

- If you want to explore the association between three quantitative variables, you can code one of them by **transparency** (more preferable) or **size** (less preferable).

```{r}
#| label: coded-alpha
ggplot(case0902, aes(x = logBody, y = logBrain, alpha = logGestation)) +
  geom_point()
```

```{r}
#| label: coded-size
ggplot(case0902, aes(x = logBody, y = logBrain, size = logGestation)) +
  geom_point()
```

## Categorical Third Variable

- If you have categorical explanatory variables, you can code their levels by **colors** (more preferable) and **shapes** (less preferable) and include this on a scatterplot of two quantitative variables.

```{r}
#| label: coded-color
data("case0901")
case0901$Time <- as.factor(case0901$Time)
ggplot(case0901, aes(x = Intensity, y = Flowers, color = Time)) +
  geom_point()
```

```{r}
#| label: coded-shape
ggplot(case0901, aes(x = Intensity, y = Flowers, shape = Time)) +
  geom_point()
```

It's polite to use colorblind-safe color palettes. The Okabe-Ito is one colorblind-safe palette.

```{r}
#| label: coded-colorblind
ggplot(case0901, aes(x = Intensity, y = Flowers, color = Time)) +
  geom_point() +
  scale_color_discrete(palette = "Okabe-Ito")
```
