---
title: "Simple Linear Regression Worksheet"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      eval       = FALSE,
                      fig.height = 3,
                      fig.width  = 4,
                      fig.align  = "center")
```

# LSD

In @wagner1968correlation, the authors intravenously injected d-lysergic acid
diethylamide (better known as LSD) into seven volunteer male subjects. They then
instructed the subjects to solve a set of simple arithmetic problems as quickly
as possible during a three-minute period. Load the data with:

```{r}
#| label: load-data
#| eval: true
#| message: false
lsd <- read_csv("https://dcgerard.github.io/stat_302/data/lsd.csv")
lsd
```

The variables are: `conc`, the tissue concentration of LSD (nanograms per mL);
and `score`, their performance score on the arithmetic questions as a percent
of a control score.

**1.** What is the explanatory variable? What is the response variable? What
are the observational units?

**2.** Make a scatterplot of `score` vs. `conc`. Does a linear model appear
appropriate?

```{r}

```

**3.** Write out the simple linear regression model for these data. Define each
term in the context of the study. What causal language (if any) is appropriate
for the slope interpretation, and why?

**4.** What assumptions does simple linear regression make about the
distribution of the **explanatory** variable?

**5.** Fit a linear model of `score` on `conc` using `lm()`.

```{r}

```

**6.** Make a residual plot (fitted values on the $x$-axis, residuals on the
$y$-axis). Does the constant-variance assumption look reasonable?

```{r}

```

**7.** State the null and alternative hypotheses to test whether concentration
is associated with score.

**8.** Use `tidy()` from the `broom` package to test the hypotheses from
Question 7. Report the test statistic, $p$-value, and a 95% confidence interval
for the slope. State a conclusion in plain language, including an interpretation
of the slope estimate.

```{r}

```

**9.** Suppose a new individual has a tissue concentration of 4 ng/mL. Compute
their predicted test score.

```{r}

```

**10.** Compute a 95% prediction interval for that individual's score. What
range of scores is plausible for a single person at that concentration?

```{r}

```

**11.** As the sample size approaches infinity, what does the standard error of
a prediction approach? Explain why.

**12.** A fellow researcher asks you to predict the score for an individual with
a concentration of 9 ng/mL. What concern would you raise?

# References
