---
title: "Chapter 1 Worksheet"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Sex Discrimination at a Bank (Case Study 1.1.2)

The data below come from a bank that was accused of paying female clerical
employees less than their male counterparts. Load the data and take a look.

```{r}
#| eval: true
#| message: false
library(tidyverse)
case0102 <- read_csv("https://dcgerard.github.io/stat_302/data/case0102.csv")
glimpse(case0102)
```

---

### Conceptual Questions

**1.** Is Case Study 1.1.2 a *randomized experiment* or an *observational study*?
Explain briefly.

**2.** Can we conclude from this data that sex *caused* the difference in
salaries? Why or why not? Identify at least one possible **confounder** that
could explain an observed salary gap.

**3.** The bank's records cover all clerical employees at this particular
institution. Does this limit how broadly we can generalize any findings?
What would need to be true for the results to apply to the broader
population of U.S. bank employees?

**4.** Let $\delta$ be the difference in population mean salaries (male minus
female) for clerical employees at this bank. State the null and alternative
hypotheses in words and in mathematical notation.

---

### Exploratory Data Analysis

**5.** Create a boxplot of `Salary` (y-axis) vs `Sex` (x-axis). Describe
what you see.

       - Functions to use: `ggplot()`, `geom_boxplot()`

```{r}

```

**6.** Calculate the mean salary for female employees. Save it as `femave`.

       - Functions to use: `mean()`, the dollar sign symbol `$`

```{r}

```

**7.** Calculate the mean salary for male employees. Save it as `malave`.
       
       - Functions to use: `mean()`, the dollar sign symbol `$`

```{r}

```

**8.** Calculate the observed difference in mean salaries (male minus female).

```{r}

```

---

### Permutation Test

The idea behind the randomization test: if sex had **no effect** on salary,
then the `Sex` labels could be shuffled randomly without changing anything
meaningful — any observed difference would just be due to chance.

**9.** The function `sample()` randomly shuffles the elements of a vector.
Try it on `case0102$Sex` and run it two or three times. What do you notice
about the output each time?

```{r}
sample(case0102$Sex)
```

**10.** Take one random permutation of `Sex` and recalculate the average
salary difference (male minus female) using the shuffled labels. How does
it compare to the observed difference from Question 8?

```{r}

```

**11.** The `replicate()` function runs a block of code `n` times and
collects the results. Run the demo below to see how it works.

```{r}
#| eval: true
demo_vec <- replicate(n = 10,
                      expr = {
                        a <- rnorm(1)
                        b <- rnorm(1)
                        b - a
                      })
demo_vec
```

Now modify the code above to build a **permutation distribution**: shuffle
`Sex` 1000 times, compute the mean salary difference (male minus female)
each time, and store all 1000 values in `sample_vec`.

```{r}

```

**12.** Plot `sample_vec` as a histogram. Add a vertical dashed line at the
observed difference from Question 8. What does the location of that line
tell you?

```{r}

```

**13.** Compute the **two-sided p-value**: the proportion of permutations
that produced an absolute difference at least as large as the one observed.

```{r}

```

---

### Interpretation

**14.** Interpret the p-value in context. Is there strong evidence that
salaries differ by sex at this bank? What would a very small p-value imply?

**15.** Suppose the p-value is very small and we conclude the salary gap
is real. Does this mean sex *caused* the difference? What is the role of
the possible confounder you identified in Question 2?

**16.** Refer back to the four-step framework from the notes:

  - **Step 1**: What is the model for this study?
  - **Step 2**: What are the hypotheses (from Question 4)?
  - **Step 3**: What test statistic did we use?
  - **Step 4**: How did we evaluate how extreme the test statistic is?
