---
title: "Spock Example"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 2.5, 
                      fig.width = 4, 
                      fig.align = "center")
```

## Objectives

- Demonstrate how to interact with ANOVA in R.

- Analyze the Spock trial data in R.

## Load in Data
```{r}
library(Sleuth3)
library(ggplot2)
data("case0502")
```


## Spock EDA

```{r}
qplot(Judge, Percent, data = case0502)
```

## Spock EDA

```{r}
qplot(Judge, Percent, data = case0502, geom = "boxplot")
```

## Before Fitting
- Always make sure the grouping variable (the explanatory variable) is either a "`factor`", a "`character`", or a "`logical`" with the `class()` function.
```{r}
class(case0502$Judge)
```
- Things will go wrong if this is any other type.
- You can force a variable to be a `factor` with the `as.factor()` function:
```{r}
case0502$Judge <- as.factor(case0502$Judge)
```


## Fit the full model
- Use `aov()` function (for **A**nalysis **O**f **V**ariance) to fit the model that assumes $\mu_1$, $\mu_2$, \ldots, $\mu_I$ are all *different*.

- Always save this output.

- The **response** variable goes on the left of the tilde (`~`) and the **explanatory** variable goes to the right of the tilde.

## Fit the full model
```{r}
aout_alldiff <- aov(Percent ~ Judge, data = case0502)
aout_alldiff
```


## Get $p$-values
- Apply `summary()` to this output to run the omnibus $F$-test
```{r}
summary(aout_alldiff)
```

## What is that Table?

- $H_0$: $\mu_1 = \mu_2 = \mu_3 = \mu_4 = \mu_5 = \mu_6 = \mu_7$
- $H_A$: At least two means are different.

```{r, echo = FALSE}
stemp <- summary(aout_alldiff)
stemp
```

\begin{center}
\small
\begin{tabular}{llllll}
 & Df & Sum Sq & Mean Sq & F value & Pr(>F) \\
Judge & $df_{extra}$ & $ESS$ & $ESS / df_{extra}$ & $F$-statistic & $p$-value \\
Residuals & $df_{full}$ & $RSS_{full}$ & $RSS_{full} / df_{full} = s_p^2$ & \\
\end{tabular}
\end{center}



## Pairwise Comparisons
- Use `pairwise.t.test()` to perform all pairwise comparisons, and then just extract the ones that you want.

## Pairwise Comparisons
\footnotesize
```{r}
ptout <- pairwise.t.test(x = case0502$Percent, 
                         g = case0502$Judge, 
                         p.adjust.method = "none")
ptout
```
\normalsize


## General Nested Comparisons
- To run tests that do not include the "all equal model", you must first *fit* both models using `aov()`, then run the `anova()` command.
- To compare the full model where all means are different to the reduced model where the non-Spock-judges have the same mean, fit a new model of Spock's judge vs the rest of the judges.

```{r}
case0502$isSpock <- case0502$Judge == "Spock's"
case0502$isSpock
```

## General Nested Comparisons
- Now fit the reduced model where $\mu_2 = \mu_3 = \cdots = \mu_7$
```{r}
aout_otherssame <- aov(Percent ~ isSpock, data = case0502)
```

## General Nested Comparisons
- Use `anova()` with both the full and reduced models to get the appropriate ANOVA table.
```{r}
anova(aout_otherssame, aout_alldiff)
```

## What is that Table?

- $H_0$: $\mu_1 \neq \mu_2 = \mu_3 = \mu_4 = \mu_5 = \mu_6 = \mu_7$
- $H_A$: At least two means are different among judges 2 through 7.

\small
```{r, echo = FALSE}
stemp <- anova(aout_otherssame, aout_alldiff)
stemp
```
\normalsize

\begin{center}
\small
\begin{tabular}{lllllll}
 & Res.Df & RSS & Df & Sum of Sq & F & Pr(>F) \\
1 & $df_{reduced}$ & $RSS_{reduced}$ & & & &\\
2 & $dr_{full}$ & $RSS_{full}$ & $df_{extra}$ & $ESS$ & $F$-statistic & $p$-value \\ 
\end{tabular}
\end{center}


## More nested comparisons
- use `anova()` with more than 2 models to get the nested sum of squares.
```{r}
aout_allequal <- aov(Percent ~ 1, data = case0502)
anova(aout_allequal, aout_otherssame, aout_alldiff)
```


