---
title: "One-Way ANOVA"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
set.seed(7)
library(broom)
library(patchwork)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(gt)
```

# Learning Objectives

- Explain the one-way ANOVA model for observational studies and randomized
  experiments.
- State and interpret the nested hypotheses.
- Compute and interpret the pooled variance estimate $s_p^2$.
- Conduct pairwise $t$-tests using the pooled variance.
- Set up and interpret the ANOVA $F$-test using the extra-sum-of-squares
  framework.
- Read a nested ANOVA table produced by R.
- Check ANOVA assumptions with residual plots.
- Describe the random-effects model and the intraclass correlation.

---

# Case Studies

```{r}
#| label: load-data
#| message: false
case0501 <- read_csv("https://dcgerard.github.io/stat_302/data/case0501.csv")
case0502 <- read_csv("https://dcgerard.github.io/stat_302/data/case0502.csv") |>
  mutate(Judge = factor(Judge))
```

## Case Study 5.1.1: Diet Restriction and Longevity

A series of studies found that restricting caloric intake can dramatically
increase life expectancy in animals. In one randomized experiment, female mice
were assigned at random to one of six diet groups:

| Label | Diet |
|:------|:-----|
| **NP** | Unlimited nonpurified standard diet |
| **N/N85** | Normal diet before and after weaning; controlled at 85 kcal/wk after (serves as the control) |
| **N/R50** | Normal before weaning; restricted to 50 kcal/wk after |
| **R/R50** | Restricted to 50 kcal/wk both before and after weaning |
| **N/R50 lopro** | Like N/R50, but dietary protein decreased with advancing age |
| **N/R40** | Normal before weaning; severely restricted to 40 kcal/wk after |

The response is **lifetime in months**. We are interested in:

1. Are *any* of the diets different from each other in terms of longevity?
2. Which specific diets differ, and by how much?


- Pairwise comparisons of interest:
  
  ![Relationship among the six diet groups. Arrows connect pairwise relationships of interest where there is only one condition that differs between diets. The slash represents pre- and post-weaning diets.](ch5_fig/diet_groups.svg){fig-align="center" width="85%"}

```{r}
#| label: diet-eda
#| fig-width: 7
case0501 |>
  mutate(Diet = factor(Diet, levels = c("NP", "N/N85", "N/R50", "R/R50", "lopro", "N/R40"))) |> ## Relevel to get boxplots in correct order
  ggplot(aes(x = Diet, y = Lifetime)) +
  geom_boxplot() +
  labs(y = "Lifetime (months)", x = "Diet group")
```

```{r}
#| label: diet-summary
case0501 |>
  mutate(Diet = factor(Diet, levels = c("NP", "N/N85", "N/R50", "R/R50", "lopro", "N/R40"))) |> ## Relevel to get boxplots in correct order
  group_by(Diet) |>
  summarise(n       = n(),
            Mean    = mean(Lifetime),
            SD      = sd(Lifetime),
            Median  = median(Lifetime)) |>
  gt() |>
  fmt_number(columns = c(3, 4, 5), decimals = 1)
```

## Case Study 5.1.2: The Spock Conspiracy Trial

In 1968, Dr. Benjamin Spock was tried in Boston on charges related to
encouraging draft resistance during the Vietnam War. The defense argued that
the Spock trial judge (referred to here as "Spock's" judge) systematically
underrepresented women on jury venires. They compared this judge's venires
with those of six other Boston-area district judges (labeled A–F).

This is an **observational study**: judge assignment to cases was not
randomized.

The response is `Percent` — the percentage of women on each venire.
We are interested in:

1. Is Spock's judge's mean percent-women lower than the other judges?
2. Are the other judges' means all the same?

```{r}
#| label: spock-eda
#| fig-width: 7
ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_boxplot() +
  labs(y = "Percent women on venire")
```

```{r}
#| label: spock-dotplot
#| fig-width: 7
ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_jitter(width = 0.15, height = 0) +
  stat_summary(fun = mean, geom = "crossbar",
               color = "tomato", linewidth = 0.5, width = 0.55) +
  labs(y = "Percent women on venire",
       caption = "Red bars = group sample means")
```

---

# The Statistical Model

In each case study we are comparing means across $I$ groups. We observe
$n_i$ individuals in group $i$, for a total of $n = n_1 + n_2 + \cdots + n_I$
observations.

**Notation:**

| Symbol | Meaning |
|:-------|:--------|
| $\mu_i$ | True population mean for group $i$, $i = 1, \ldots, I$ |
| $Y_{ij}$ | Observed value for the $j$th individual in group $i$ |
| $\varepsilon_{ij}$ | Individual-specific noise for the $j$th individual in group $i$: mean 0, variance $\sigma^2$ |

**Model:**
$$Y_{ij} = \mu_i + \varepsilon_{ij}$$

In words: each observed value equals the true group mean plus a random
individual-level noise term. The noise captures everything we can't explain —
genetic variation between mice, measurement error, day-to-day fluctuation in
a judge's venire composition, etc.

The key restriction is **equal variance**: $\sigma^2$ is assumed the same for
every group and every individual. A mouse in the N/R40 diet group has the same
background noise variance as a mouse in the NP group; they just have different
expected lifetimes. This is the $I$-group generalization of the pooled
two-sample $t$-test from Chapter 2.

**Spock trial interpretation:**

- $\mu_1$ = true mean percent-women for Spock's judge across all possible
  venires
- $\mu_i$ = true mean percent-women for judge $i$, $i = 2, \ldots, 7$
  (judges A–F)
- $Y_{ij}$ = observed percent women on venire $j$ of judge $i$
- $\varepsilon_{ij}$ = random variation in how that particular venire was drawn

**Observational vs. randomized:**
- The model equation $Y_{ij} = \mu_i + \varepsilon_{ij}$ looks the same in both the Spock trial and the longevity study.

- But only in the longevity study can we interpret any causal association (diet and lifetime).

- E.g., if $\mu_{NP}$ is the mean lifetime of mice on the NP diet, and $\mu_{N/N85}$ is the mean lifetime of the mice on the N/N85 diet, then $\delta = \mu_{N/N85} - \mu_{NP}$ is the treatment effect.

- That is, $\delta$ is how much longer, on average, a mouse would live if given the N/N85 diet over the NP diet.

---

# Hypotheses

With $I$ groups there are many possible hypotheses.

## Pairwise Hypotheses

Test whether any specific pair of means differs:
$$H_0: \mu_2 = \mu_5, \qquad H_A: \mu_2 \neq \mu_5$$

With $I = 7$ groups there are $\binom{7}{2} = 21$ possible pairwise tests.

## Omnibus Hypothesis ("Are *any* means different?")

$$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4 = \mu_5 = \mu_6 = \mu_7$$
$$H_A: \text{at least one } \mu_i \neq \mu_j$$

## One Group vs. the Rest

$$H_0: \mu_1 = \tfrac{1}{6}(\mu_2 + \mu_3 + \mu_4 + \mu_5 + \mu_6 + \mu_7)$$
$$H_A: \mu_1 \neq \tfrac{1}{6}(\mu_2 + \mu_3 + \mu_4 + \mu_5 + \mu_6 + \mu_7)$$

This asks: *Is Spock's judge's mean different from the average of the other
judges?*

---

# Estimation

## Group Sample Means

Estimate $\mu_i$ with the group sample mean:
$$\bar{Y}_{i\bullet} = \frac{1}{n_i}\sum_{j=1}^{n_i} Y_{ij}$$

## Pooled Variance

Each group has its own sample standard deviation $s_i$, measuring the spread
of observations *within* that group. Because we assume the same true variance
$\sigma^2$ for every group, we can combine these within-group estimates into one
more accurate estimate. We weight each group by the amount of information it
contributes — groups with more observations get more weight:

$$s_p^2 = \frac{(n_1-1)s_1^2 + (n_2-1)s_2^2 + \cdots + (n_I-1)s_I^2}
               {(n_1-1) + (n_2-1) + \cdots + (n_I-1)}$$

In words: each term $(n_i-1)s_i^2$ in the numerator is the sum of squared
deviations from group $i$'s mean. Adding these across all groups gives the
total within-group variation. Dividing by the total degrees of freedom gives
the average within-group variance.

The denominator is $\nu = n - I$ (the total degrees of freedom available for
estimating variance within groups — we lose one degree of freedom per group
because we estimated each group's mean).

Two observations:

- This is a weighted average of the within-group variance estimates, weighted
  by their degrees of freedom.
- When $I = 2$ this reduces exactly to the pooled variance from Chapter 2.

---

# Testing Pairwise Differences

To test $H_0: \mu_i = \mu_j$ vs $H_A: \mu_i \neq \mu_j$, we build a
$t$-statistic in three steps.

**Step 1 — Distribution of the difference under $H_0$.**
Each group mean is approximately normal: $\bar{Y}_{i\bullet} \sim N(\mu_i,\; \sigma^2/n_i)$.
Under $H_0$ ($\mu_i = \mu_j$), the two means are independent, so their
difference is:
$$\bar{Y}_{i\bullet} - \bar{Y}_{j\bullet}
  \;\sim\; N\!\left(0,\; \sigma^2\!\left(\tfrac{1}{n_i} + \tfrac{1}{n_j}\right)\right)$$

**Step 2 — Standardize.** Dividing by the true standard deviation gives a
standard normal:
$$\frac{\bar{Y}_{i\bullet} - \bar{Y}_{j\bullet}}
      {\sigma\sqrt{1/n_i + 1/n_j}} \;\sim\; N(0,1)$$

**Step 3 — Estimate $\sigma$.** We don't know $\sigma$, so we substitute the
pooled estimate $s_p$. Replacing a known standard deviation with an estimate
inflates uncertainty, and the result follows a $t$ distribution:

$$t^* = \frac{\bar{Y}_{i\bullet} - \bar{Y}_{j\bullet}}{s_p\sqrt{1/n_i + 1/n_j}}
       \;\sim\; t_\nu
       \quad \text{under } H_0$$

where $\nu = n - I$.

The advantage over running separate two-sample $t$-tests is that **we use all
the data to estimate $\sigma^2$**, giving more degrees of freedom and a more
accurate variance estimate.

A 95% confidence interval for $\mu_i - \mu_j$ is:

$$\bar{Y}_{i\bullet} - \bar{Y}_{j\bullet}
  \;\pm\; t_{\nu}(0.975) \cdot s_p\sqrt{1/n_i + 1/n_j}$$

::: {.callout-warning}
## Multiple Comparisons

Running 21 pairwise tests at $\alpha = 0.05$ inflates the overall Type I error
rate well above 5%. Chapter 6 covers methods that control for multiple
comparisons. For now, we treat each pairwise test independently and note the
concern.
:::

---

# The ANOVA $F$-Test

For the omnibus hypothesis ($H_0$: all means equal vs $H_A$: at least two differ),
a single summary statistic is needed. The ANOVA $F$-test compares two **nested
models** by asking: *how much does model fit improve when we allow separate group
means?*

## Full and Reduced Models

| | Reduced model ($H_0$) | Full model ($H_A$) |
|:--|:--|:--|
| **Means** | $\mu_1 = \mu_2 = \cdots = \mu_I = \mu$ (one shared mean) | $\mu_1, \mu_2, \ldots, \mu_I$ (separate means) |
| **Estimates** | $\hat{\mu} = \bar{Y}_{\bullet\bullet}$ | $\hat{\mu}_i = \bar{Y}_{i\bullet}$ |
| **# parameters** | 1 | $I$ |

Where
$$
Y_{i\bullet} = \frac{1}{n_i}\sum_{j=1}^{n_i}Y_{ij}
$$
$$
\bar{Y}_{\bullet\bullet} = \frac{1}{n}\sum_{i=1}^I\sum_{j=1}^{n_i}Y_{ij},
$$
$$
n = \sum_{i=1}^{I}n_i
$$

For the Spock data ($I = 7$ judges), here is what each model predicts for
each group:

| Group | Judge | Full model prediction | Reduced model prediction |
|:-----:|:------|:---------------------:|:------------------------:|
| 1 | Spock's | $\bar{Y}_{1\bullet}$ | $\bar{Y}_{\bullet\bullet}$ |
| 2 | Judge A | $\bar{Y}_{2\bullet}$ | $\bar{Y}_{\bullet\bullet}$ |
| 3 | Judge B | $\bar{Y}_{3\bullet}$ | $\bar{Y}_{\bullet\bullet}$ |
| 4 | Judge C | $\bar{Y}_{4\bullet}$ | $\bar{Y}_{\bullet\bullet}$ |
| 5 | Judge D | $\bar{Y}_{5\bullet}$ | $\bar{Y}_{\bullet\bullet}$ |
| 6 | Judge E | $\bar{Y}_{6\bullet}$ | $\bar{Y}_{\bullet\bullet}$ |
| 7 | Judge F | $\bar{Y}_{7\bullet}$ | $\bar{Y}_{\bullet\bullet}$ |

In the **full model**, each group gets its own sample mean as a prediction.
Spock's judge's venires are predicted by Spock's judge's average; Judge A's
venires are predicted by Judge A's average; and so on. There are 7 free
parameters — one per judge.

In the **reduced model**, every observation — regardless of judge — is
predicted by the same grand average $\bar{Y}_{\bullet\bullet}$ (the mean over
all 46 venires). This is the model that says $H_0$ is true: if all judges
have the same expected percent-women, then the best single prediction for any
venire is just the overall average.

**What is a residual?** A residual is simply the distance between an observed
value and the model's prediction for it:

- Full model residual: $Y_{ij} - \bar{Y}_{i\bullet}$ 

  - how far observation $(i,j)$ is from *its group's mean*
  
- Reduced model residual: $Y_{ij} - \bar{Y}_{\bullet\bullet}$ 

  - how far observation $(i,j)$ is from *the grand mean*

Intuitively, if the groups really have different means, the full model
residuals will be small (each observation is close to its own group mean),
while the reduced model residuals will be large (observations from Spock's
group, for example, are far from the overall grand mean).

## Sums of Squared Residuals

The **residual sum of squares** (RSS) for each model is:

$$RSS_{full} = \sum_i \sum_j (Y_{ij} - \bar{Y}_{i\bullet})^2,
\quad df_{full} = n - I$$

$$RSS_{reduced} = \sum_i \sum_j (Y_{ij} - \bar{Y}_{\bullet\bullet})^2,
\quad df_{reduced} = n - 1$$

- Each RSS adds up all the squared residuals across every observation.
  - a small RSS means predictions are close to the data
  - a large RSS means they are far away. 
  - Both RSS values involve squaring (so positive and negative residuals don't cancel) and summing (so we get a single overall measure).

- The **extra sum of squares** (ESS) measures how much the fit improves when we go from the reduced model to the full model:

  $$ESS = RSS_{reduced} - RSS_{full}, \quad df_{extra} = I - 1$$

- **NOTE: $RSS_{reduced} \geq RSS_{full}$** 
  - The full model has the reduced model as a special case: if all the $\mu_i$ happen to be equal, the full model's separate group means would all estimate the same value as the grand mean, and the two models would give identical fits. In general, the full model has more flexibility, so it can *never* fit worse than the reduced model. It can always do at least as well. Therefore $ESS \geq 0$ always.

- **What does ESS tell us?** 
  - If $H_0$ is true (all means really are equal), fitting separate group means should not improve the fit by very much — the data don't contain strong evidence for distinct groups. In that case, ESS will be positive but small. If the means truly differ (the data come from groups with different means), the full model will fit much better, and ESS will be large. The $F$-test formalizes what "small" and "large" mean.

The plots below illustrate this for the Spock data. Red bars/lines show the
model's fitted means; larger gaps between points and those lines = larger
residuals.

```{r}
#| label: null-vs-full
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
#| message: false
ybar <- mean(case0502$Percent)

grp_means <- case0502 |>
  group_by(Judge) |>
  summarise(mean_pct = mean(Percent))

rss_null <- sum((case0502$Percent - ybar)^2)
rss_full <- case0502 |>
  left_join(grp_means, by = "Judge") |>
  summarise(rss = sum((Percent - mean_pct)^2)) |>
  pull(rss)

p_null <- ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.7) +
  geom_hline(yintercept = ybar, color = "tomato",
             linetype = "dashed", linewidth = 1) +
  labs(title = "Null model: one overall mean",
       subtitle = paste0("RSS_reduced = ", round(rss_null, 0))) +
  theme(plot.subtitle = element_text(size = 9))

p_full <- ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.7) +
  stat_summary(fun = mean, geom = "crossbar",
               color = "tomato", linewidth = 0.5, width = 0.6) +
  labs(title = "Full model: separate group means",
       subtitle = paste0("RSS_full = ", round(rss_full, 0))) +
  theme(plot.subtitle = element_text(size = 9))

p_null + p_full
```

The squared residuals are much smaller under the full model, confirming that
group means differ substantially.

```{r}
#| label: sq-resid-boxplot
#| echo: false
#| fig-height: 3
sq_resid_df <- bind_rows(
  case0502 |>
    left_join(grp_means, by = "Judge") |>
    mutate(sq_resid = (Percent - mean_pct)^2, Model = "Full"),
  case0502 |>
    mutate(sq_resid = (Percent - ybar)^2, Model = "Reduced")
)

ggplot(sq_resid_df, aes(x = Model, y = sq_resid)) +
  geom_boxplot() +
  labs(y = "Squared residual",
       title = "Squared residuals are smaller in the full model")
```

## The $F$-Statistic

If $H_0$ is true, both $ESS/df_{extra}$ and $RSS_{full}/df_{full} = s_p^2$
estimate the same $\sigma^2$, so their ratio should be near 1. If $H_A$ is
true, the means differ and $ESS$ will be inflated, pushing the ratio above 1.

$$F = \frac{ESS / df_{extra}}{s_p^2}
    = \frac{ESS / (I-1)}{RSS_{full} / (n-I)}$$

Under $H_0$, $F \sim F_{I-1,\; n-I}$. The $p$-value uses only the upper
tail (large $F$ is evidence against $H_0$).

## The $F$-Distribution

The $F$ distribution is parameterized by two degrees of freedom: numerator
($df_1$) and denominator ($df_2$). Below are a few shapes:

```{r}
#| label: f-dist-shapes
#| echo: false
#| fig-width: 9
#| fig-height: 3
x_seq <- seq(0, 6, length = 300)
f_df <- bind_rows(
  data.frame(x = x_seq, y = df(x_seq, 2,   10),  label = "F(2, 10)"),
  data.frame(x = x_seq, y = df(x_seq, 6,   39),  label = "F(6, 39)"),
  data.frame(x = x_seq, y = df(x_seq, 10, 100),  label = "F(10, 100)")
)

ggplot(f_df, aes(x = x, y = y, color = label)) +
  geom_line(linewidth = 0.9) +
  labs(x = "F", y = "f(F)", color = NULL,
       title = "F-distribution shapes") +
  theme(legend.position = "bottom")
```

## $p$-Value for the Spock Omnibus Test

The Spock data has $I = 7$ judges and $n = 46$ venires, so $df_{extra} = 6$
and $df_{full} = 39$.

```{r}
#| label: spock-ftest-manual
n     <- nrow(case0502)
I     <- n_distinct(case0502$Judge)
df_extra <- I - 1          # 6
df_full  <- n - I          # 39
ess      <- rss_null - rss_full
sp2      <- rss_full / df_full
fstat    <- (ess / df_extra) / sp2
c(ESS = ess, RSS_full = rss_full, sp2 = sp2, F = fstat)
```

```{r}
#| label: spock-pvalue
pf(fstat, df1 = df_extra, df2 = df_full, lower.tail = FALSE)
```

```{r}
#| label: spock-f-plot
#| echo: false
#| fig-height: 2.8
x <- seq(0, 8, length = 500)
y <- df(x, df1 = 6, df2 = 39)
poly_x <- c(fstat, x[x >= fstat], max(x[x >= fstat]))
poly_y <- c(0, y[x >= fstat], 0)

ggplot(data.frame(x = x, y = y), aes(x, y)) +
  geom_line() +
  geom_polygon(data = data.frame(x = poly_x, y = poly_y),
               aes(x, y), fill = "steelblue", alpha = 0.6) +
  geom_vline(xintercept = fstat, linetype = "dashed") +
  annotate("text", x = fstat + 0.35, y = 0.05,
           label = paste0("F* = ", round(fstat, 2)), hjust = 0) +
  labs(x = "F", y = "f(F)",
       title = expression(F[list(6, 39)] ~ "null distribution"),
       subtitle = "Shaded area = p-value")
```

---

# A Hierarchy of Models

An F-test can be used to compare any two **nested** models. 

In the Spock example, we had three nested models:

- **Null**: all means equal $(\mu)$
- **Intermediate**: Spock's judge differs from a common mean for the rest
  $(\mu_\text{Spock}, \mu_\text{others})$
- **Full**: all seven means free $(\mu_1, \mu_2, \ldots, \mu_7)$

Here is what each model predicts for each judge in the Spock data.

**Null vs. Intermediate** — Is Spock's judge different from the rest?

- $H_0: \mu_1 = \cdots = \mu_7$ (all same)
- $H_A: \mu_1 \neq \mu_2 = \cdots = \mu_7$ (Spock is different)

| Group | Judge | Null (reduced) | Intermediate (full) |
|:-----:|:------|:--------------:|:-------------------:|
| 1 | Spock's | $\bar{Y}_{\bullet\bullet}$ | $\bar{Y}_{1\bullet}$ |
| 2 | Judge A | $\bar{Y}_{\bullet\bullet}$ | $\bar{Y}_{0}$ |
| 3 | Judge B | $\bar{Y}_{\bullet\bullet}$ | $\bar{Y}_{0}$ |
| 4 | Judge C | $\bar{Y}_{\bullet\bullet}$ | $\bar{Y}_{0}$ |
| 5 | Judge D | $\bar{Y}_{\bullet\bullet}$ | $\bar{Y}_{0}$ |
| 6 | Judge E | $\bar{Y}_{\bullet\bullet}$ | $\bar{Y}_{0}$ |
| 7 | Judge F | $\bar{Y}_{\bullet\bullet}$ | $\bar{Y}_{0}$ |

Here $\bar{Y}_{\bullet\bullet}$ is the grand mean of all 46 venires, and
$\bar{Y}_{0}$ is the average percent-women across all venires belonging to
judges A–F (groups 2 through 7). The intermediate model has 2 parameters:
one for Spock's judge ($\bar{Y}_{1\bullet}$) and one pooled mean for the
remaining six judges ($\bar{Y}_{0}$).

The plots below show how these two models fit the data. The drop in RSS
when moving from the null to the intermediate model is the ESS for this
test.

```{r}
#| label: null-vs-inter
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
spock_mean <- case0502 |>
  filter(Judge == "Spock's") |>
  pull(Percent) |>
  mean()
other_mean <- case0502 |>
  filter(Judge != "Spock's") |>
  pull(Percent) |>
  mean()

rss_inter <- case0502 |>
  mutate(fitted = if_else(Judge == "Spock's", spock_mean, other_mean)) |>
  summarise(rss = sum((Percent - fitted)^2)) |>
  pull(rss)

seg_inter <- data.frame(
  x    = c(0.65, 6.65),
  xend = c(6.35, 7.35),
  y    = c(other_mean, spock_mean),
  yend = c(other_mean, spock_mean)
)

p_null2 <- ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.7) +
  geom_hline(yintercept = ybar, color = "tomato",
             linetype = "dashed", linewidth = 1) +
  labs(title = "Null model: one overall mean",
       subtitle = paste0("RSS = ", round(rss_null, 0))) +
  theme(plot.subtitle = element_text(size = 9))

p_inter <- ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.7) +
  geom_segment(data = seg_inter,
               aes(x = x, xend = xend, y = y, yend = yend),
               color = "tomato", linetype = "dashed", linewidth = 1,
               inherit.aes = FALSE) +
  labs(title = "Intermediate model: Spock's judge differs",
       subtitle = paste0("RSS = ", round(rss_inter, 0))) +
  theme(plot.subtitle = element_text(size = 9))

p_null2 + p_inter
```

**Intermediate vs. Full** — Do judges A–F differ among themselves?

- $H_0: \mu_1 \neq \mu_2 = \cdots = \mu_7$ (Spock is different)
- $H_A:$ At least one $\mu_i \neq \mu_j$ for $i,j\in \{2,\ldots,7\}$

| Group | Judge | Intermediate (reduced) | Full |
|:-----:|:------|:----------------------:|:----:|
| 1 | Spock's | $\bar{Y}_{1\bullet}$ | $\bar{Y}_{1\bullet}$ |
| 2 | Judge A | $\bar{Y}_{0}$ | $\bar{Y}_{2\bullet}$ |
| 3 | Judge B | $\bar{Y}_{0}$ | $\bar{Y}_{3\bullet}$ |
| 4 | Judge C | $\bar{Y}_{0}$ | $\bar{Y}_{4\bullet}$ |
| 5 | Judge D | $\bar{Y}_{0}$ | $\bar{Y}_{5\bullet}$ |
| 6 | Judge E | $\bar{Y}_{0}$ | $\bar{Y}_{6\bullet}$ |
| 7 | Judge F | $\bar{Y}_{0}$ | $\bar{Y}_{7\bullet}$ |

This test asks: after accounting for Spock's judge, do the remaining six
judges' venires show any systematic differences from each other?

The plots below compare the fit of the intermediate model (reduced) against
the full model. The ESS here captures only the variation *among* judges A–F.

```{r}
#| label: inter-vs-full
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
p_inter2 <- ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.7) +
  geom_segment(data = seg_inter,
               aes(x = x, xend = xend, y = y, yend = yend),
               color = "tomato", linetype = "dashed", linewidth = 1,
               inherit.aes = FALSE) +
  labs(title = "Intermediate model: Spock's judge differs",
       subtitle = paste0("RSS = ", round(rss_inter, 0))) +
  theme(plot.subtitle = element_text(size = 9))

p_full2 <- ggplot(case0502, aes(x = Judge, y = Percent)) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.7) +
  stat_summary(fun = mean, geom = "crossbar",
               color = "tomato", linewidth = 0.5, width = 0.55) +
  labs(title = "Full model: all judges separate",
       subtitle = paste0("RSS = ", round(rss_full, 0))) +
  theme(plot.subtitle = element_text(size = 9))

p_inter2 + p_full2
```

Notice that we always fit *both* models using *all* the data — we never
discard Spock's judge's observations even when comparing only judges A–F. This
matters for the variance estimate: instead of testing on a subset of 39 venires
(judges A–F only, $df = 32$), we keep all 46 venires and use $s_p^2$ from the
full model ($df = 39$). More degrees of freedom means a better variance
estimate and a more powerful test.

![Hierarchy of models for the Spock trial. Arrows show $F$-tests between adjacent models. RSS = residual sum of squares; df = degrees of freedom.](ch5_fig/model_hierarchy.svg){fig-align="center" width="90%"}

The full sequential ANOVA table for the Spock data is:

| Model | $df_{RSS}$ | RSS | $df_{extra}$ | ESS | $F$ | $p$-value |
|:------|:----------:|----:|:------------:|----:|----:|----------:|
| All same | 45 | 3792 | | | | |
| Spock's differs | 44 | 2191 | 1 | 1601 | 33.48 | <0.001 |
| All different | 39 | 1864 | 5 | 326 | 1.37 | 0.26 |

**Conclusions:**

- Strong evidence that Spock's judge has a different (lower) mean than the
  other judges ($F_{1,39} = 33.48$, $p < 0.001$).
- No evidence that the six non-Spock judges differ among themselves
  ($F_{5,39} = 1.37$, $p = 0.26$).

---

# One-Way ANOVA in R

## Checking the Grouping Variable

Before fitting, confirm the grouping variable is a `factor` or `character`:

```{r}
class(case0502$Judge)
```

If it were numeric, `aov()` would treat it as a continuous predictor, not
a grouping variable. Use `as.factor()` or `mutate(Judge = factor(Judge))`
to fix this.

## Omnibus $F$-Test with `aov()` and `tidy()`

```{r}
#| label: aov-full
aout_full <- aov(Percent ~ Judge, data = case0502)
tidy(aout_full)
```

The table has two rows:

| Row | $df$ | Sum Sq | Mean Sq | $F$ | $p$-value |
|:----|:-----|-------:|--------:|----:|----------:|
| `Judge` | $df_{extra} = I - 1$ | $ESS$ | $ESS / df_{extra}$ | $F$-stat | $p$ |
| `Residuals` | $df_{full} = n - I$ | $RSS_{full}$ | $RSS_{full}/df_{full} = s_p^2$ | | |

## Pairwise Comparisons

```{r}
#| label: pairwise-t
ptout <- pairwise.t.test(x = case0502$Percent,
                         g = case0502$Judge,
                         p.adjust.method = "none")
ptout
```

Each cell is the raw (unadjusted) $p$-value for that pair. Note that
`pairwise.t.test()` uses $s_p^2$ from **all** groups to form the $t$-statistic,
even when comparing just two groups.

## Nested Model Comparisons with `anova()`

To test the intermediate hypothesis (Spock's differs; others pooled):

```{r}
#| label: nested-models
case0502 <- case0502 |>
  mutate(isSpock = Judge == "Spock's")

aout_intermediate <- aov(Percent ~ isSpock, data = case0502)
aout_null         <- aov(Percent ~ 1,       data = case0502)
```

```{r}
#| label: spock-vs-others
anova(aout_intermediate, aout_full)
```

The rows are:

| Row | $df_{RSS}$ | RSS | $df_{extra}$ | ESS | $F$ | $p$ |
|:----|:----------:|----:|:------------:|----:|----:|----:|
| Model 1 (reduced) | $df_{reduced}$ | $RSS_{reduced}$ | | | | |
| Model 2 (full) | $df_{full}$ | $RSS_{full}$ | $df_{extra}$ | ESS | $F$-stat | $p$ |

## Full Sequential ANOVA Table

Compare all three models at once:

```{r}
#| label: three-model-anova
anova(aout_null, aout_intermediate, aout_full)
```

This produces the sequential table shown in the hierarchy section above. Row 2
tests Null vs Intermediate; Row 3 tests Intermediate vs Full.

Note: the $F$-statistics shown here use $s_p^2$ from the **full** model
(`aout_full`) as the denominator.

---

# Assumptions and Diagnostics

The same three assumptions from the two-sample $t$-test apply, in the same
order of importance:

1. **Independence** of observations.
2. **Equal variance** across all groups.
3. **Normality** (approximately symmetric, no severe outliers).

**New tool for checking assumptions 2 and 3: residual plots.**

The **residual** for observation $(i,j)$ under the full model is:
$$e_{ij} = Y_{ij} - \bar{Y}_{i\bullet}$$

Plotting residuals removes the group-mean structure, letting us focus on
the *variation about the mean* without visual clutter from the differences
in means between groups.

## Residual Plots in R

The quickest way is `plot()` applied to an `aov()` object:

```{r}
#| label: base-resid-plots
#| fig-width: 7
#| fig-height: 6
par(mfrow = c(2, 2))
plot(aout_full)
par(mfrow = c(1, 1))
```

The two most important panels:

- **Residuals vs. Fitted**: residuals ($y$-axis) vs. group means ($x$-axis).
  Look for unequal spread across the fitted values.
- **Normal Q-Q**: QQ-plot of residuals. Look for departures from a straight
  line (skew, heavy tails, outliers).

## Manual `ggplot2` Residual Plots

```{r}
#| label: ggplot-resids
#| warning: false
aug_full <- augment(aout_full)

p_rv_fit <- ggplot(aug_full, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted values (group means)",
       y = "Residuals",
       title = "Residuals vs. fitted")

p_rv_grp <- ggplot(aug_full, aes(x = Judge, y = .resid)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "Residuals", title = "Residuals by group")

p_rv_fit + p_rv_grp
```

```{r}
#| label: qq-resid
ggplot(aug_full, aes(sample = .resid)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Normal Q-Q plot of residuals")
```

## What to Look For

**Unequal variance** appears as fan-shaped spread in the Residuals vs.
Fitted plot — larger spread for larger fitted values (or vice versa):

```{r}
#| label: bad-unequal-var
#| echo: false
set.seed(2)
resid_bad_var <- rnorm(n = nrow(case0502),
                       mean = 0,
                       sd   = aug_full$.fitted^3 / 3e6)
tibble(resid = resid_bad_var, fitted = aug_full$.fitted) |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted values", y = "Residuals",
       title = "Simulated unequal variance (fan shape)")
```

**An outlier** appears as an isolated point far from 0:

```{r}
#| label: bad-outlier
#| echo: false
resid_out <- aug_full$.resid
resid_out[1] <- 35
tibble(resid = resid_out, fitted = aug_full$.fitted) |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Fitted values", y = "Residuals",
       title = "Simulated outlier")
```

**Solutions** mirror Chapter 3:

| Violation | Solution |
|:----------|:---------|
| Independence | ANOVA with blocking (Chapters 12–13) |
| Unequal variance | Welch-style ANOVA; log transformation |
| Non-normality / outliers | Transformation; run with and without outlier |

---

# Random Effects

In the fixed-effects model above, the groups (judges, diets) are the
specific groups of interest — we want to compare these particular groups.

Sometimes the groups are **not** of direct interest. Instead, they are a
random sample from a larger population of groups, and we want to make
inferences about **variability among groups** in that population.

**Example**: Quality control — measuring output of 7 randomly selected
machines from a factory. We don't care about those specific 7 machines; we
want to know how much machine-to-machine variability exists across the entire
factory floor.

When groups are a random sample from a larger population of groups, use the
**random-effects model**.

## The Random-Effects Model

The model is written in two layers:

$$Y_{ij} = \mu_i + \varepsilon_{ij},
\quad \text{var}(\varepsilon_{ij}) = \sigma^2$$
$$\mu_i = \mu + \xi_i,
\quad \text{var}(\xi_i) = \sigma_p^2$$

The first line is identical to the fixed-effects model. The second line says
that each group mean $\mu_i$ is itself **random** — it is the overall average
$\mu$ plus a random group-specific deviation $\xi_i$. These deviations have
mean 0 and variance $\sigma_p^2$.

In words: every observation has two sources of randomness:

1. **Within-group noise** $\varepsilon_{ij}$: how much individual $j$ in
   group $i$ differs from its group's mean (variance $\sigma^2$).
2. **Between-group noise** $\xi_i$: how much group $i$'s mean differs from
   the overall mean $\mu$ (variance $\sigma_p^2$).

If $\sigma_p^2 = 0$, all groups have the same true mean ($\mu_i = \mu$ for
all $i$), and there is no between-group variability at all.

**Hypotheses:**
$$H_0: \sigma_p^2 = 0 \quad \text{(no between-group variability — all group means equal } \mu\text{)}$$
$$H_A: \sigma_p^2 > 0 \quad \text{(some groups have higher/lower means than others)}$$

We run the **same $F$-test as before**. Under $H_0$, the $F$-statistic has
the same $F_{I-1,\, n-I}$ null distribution.

## Why Use This Model?

In one-way ANOVA, the $p$-value from the $F$-test is identical whether you
treat groups as fixed or random. So why bother with random effects?

In more complex models — for example, when you have multiple random factors,
or when the same individuals appear in multiple groups — treating groups as
random leads to different (and generally better) estimates and inference.
The one-way case is the simplest introduction to this framework.

## Intraclass Correlation (ICC)

The ICC measures the **proportion of total variance explained by group
membership**:

$$\text{ICC} = \frac{\hat\sigma_p^2}{\hat\sigma_p^2 + \hat\sigma^2}$$

To see why: the total variance of any observation $Y_{ij}$ is
$\sigma_p^2 + \sigma^2$ (between-group plus within-group). The ICC is the
fraction of that total coming from between-group differences.

- **ICC close to 1**: most variation is *between* groups — knowing the group
  an observation belongs to tells you most of what you need to know about its
  value.
- **ICC close to 0**: most variation is *within* groups — group membership
  explains very little; individual-level noise dominates.

---

# Exercises

**1.** The diet study (Case Study 5.1.1) is a randomized experiment, while
the Spock trial (Case Study 5.1.2) is an observational study.

  a. For each study, identify the explanatory variable and the response
     variable.
  b. For each study, state whether a causal conclusion is justified if the
     $F$-test is significant. Explain.

**2.** Using the diet data (`case0501`):

  a. Fit the one-way ANOVA model with `aov()` and display the ANOVA table.
  b. State the null and alternative hypotheses for the omnibus $F$-test.
  c. Report the $F$-statistic, degrees of freedom, and $p$-value. What do
     you conclude?

```{r}

```

**3.** From the ANOVA table in Exercise 2, identify (by number, not name):

  a. $RSS_{full}$ and $df_{full}$.
  b. $s_p^2$ (the pooled variance estimate).
  c. $ESS$ and $df_{extra}$.
  d. The $F$-statistic — verify it equals $ESS/df_{extra}$ divided by $s_p^2$.

**4.** For the Spock data, use `pairwise.t.test()` to obtain all pairwise
$p$-values (unadjusted).

  a. How many pairwise comparisons are there?
  b. Which pairs, if any, show evidence of a difference at $\alpha = 0.05$?
  c. Why is it potentially problematic to use $\alpha = 0.05$ for each test
     individually when conducting this many comparisons?

```{r}

```

**5.** Fit the three nested models for the Spock data (null, intermediate,
full) and reproduce the sequential ANOVA table using `anova()`.

  a. What is Spock's judge's estimated percent-women on the venire?
  b. What is the pooled estimate for the other six judges?
  c. Interpret the two $F$-tests in the sequential table. What overall
     conclusion do you draw about the Spock trial jury selection?

```{r}

```

**6.** For the diet study, test whether the two most calorie-restricted groups
(N/R40 and R/R50) differ in mean lifetime from the control group (N/N85).

  a. Create a new binary variable that indicates whether the mouse is in one
     of those two groups vs. the control group.
  b. Fit the appropriate reduced model and use `anova()` to test the
     hypothesis.
  c. Interpret the result.

```{r}

```

**7.** For the Spock data, make the three residual plots:

  a. Residuals vs. fitted values.
  b. Residuals vs. judge (boxplots).
  c. QQ-plot of residuals.

Comment on whether the equal-variance and normality assumptions appear
reasonable.

```{r}

```

**8.** Explain in your own words why the pooled variance estimate $s_p^2$
uses **all** groups (not just the two being compared) when conducting a
pairwise $t$-test in the one-way ANOVA framework. What is the practical
benefit of this approach?

**9.** For the Spock data, the omnibus $F$-test gives a very small $p$-value.

  a. Does a significant omnibus $F$-test tell you *which* pairs of means differ?
     Explain.
  b. After seeing a significant omnibus result, a student decides to examine
     only the comparison that looks most different in the boxplot and ignores
     the others. What is the statistical problem with this strategy?

**10.** Consider a one-way ANOVA with $I = 4$ groups and $n = 40$ total
observations.

  a. What are $df_{extra}$ and $df_{full}$?
  b. If $RSS_{full} = 720$ and $ESS = 180$, compute $s_p^2$, the $F$-statistic,
     and the $p$-value.
  c. What is the $p$-value from `pf()` in R?

```{r}

```

**11.** Describe the difference between a **fixed-effects** and a
**random-effects** one-way ANOVA model. For each scenario below, state which
model is more appropriate and why.

  a. Comparing mean test scores across 5 specific schools that were chosen
     because they use different curricula.
  b. Comparing output quality across 6 machines randomly selected from a
     factory floor, to estimate how much machine-to-machine variability exists.
  c. Comparing soil nitrogen levels across 8 randomly chosen plots in a large
     field.

**12.** A researcher fits a one-way ANOVA with $I = 5$ groups and finds
$RSS_{full} = 400$ (df = 45), $RSS_{reduced} = 520$ (df = 49).

  a. Compute $ESS$ and $df_{extra}$.
  b. Compute $s_p^2$ and the $F$-statistic.
  c. Compute the $p$-value. Is there evidence of group differences?
  d. Compute the ICC estimate, given that the estimated between-group variance
     is $\hat\sigma_p^2 = 6.0$.
