---
title: "Chapter 5 Worksheet"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Star Trek: The Next Generation Dialogue

The data in `trek.csv` contain word counts and proportions of words spoken by
the main characters from [Star Trek: The Next Generation](https://en.wikipedia.org/wiki/Star_Trek:_The_Next_Generation).
The original dataset was obtained from [github.com/RMHogervorst/TNG](https://github.com/RMHogervorst/TNG).
Each row summarizes the dialogue for one character in one season. The variables
are:

- `season`: the season of the show (1–7).
- `character`: the character's name (lower case).
- `count`: total word count for that character in that season.
- `prop`: proportion of all words in that season spoken by that character.
- `logprop`: natural log of `prop`.

Captain Picard is generally considered the main character. We want to answer
three questions:

1. Do characters have about the same amount of dialogue?
2. Does Picard speak more than the other characters?
3. Do the other characters speak about the same amount as each other?

**Note:** ANOVA is not quite the ideal method here because the independence
assumption is somewhat violated — as one character speaks more, the others
must speak less within each season. We will proceed anyway and treat any
conclusions as preliminary.

We will use `logprop` as our response throughout.

```{r}
#| label: load-data
#| eval: true
#| message: false
library(tidyverse)
trek <- read_csv("https://dcgerard.github.io/stat_302/data/trek.csv")
glimpse(trek)
```

---

**1.** Make a plot to informally explore questions of interest 1 through 3.
Which characters appear to speak the most? Does it look like Picard stands
out from the rest?

```{r}

```

**2.** State the one-way ANOVA model and the omnibus null and alternative
hypotheses for exploring question 1. Define all symbols.

**3.** Test the omnibus hypothesis from Question 2 using `aov()` and
`tidy()`. Report the $F$-statistic, degrees of freedom, and $p$-value.

```{r}

```

**4.** Interpret the result from Question 3 in plain language. What does the
$p$-value tell you? Does a significant omnibus result tell you *which*
characters differ? Explain.

**5.** Create a new variable `isPicard` that is `TRUE` when `character ==
"picard"` and `FALSE` otherwise.

```{r}

```

**6.** Use `aov()` to fit the *intermediate* model in which only Picard has
his own mean and all other characters share a common mean.

```{r}

```

**7.** Use `aov()` to fit the *null* model in which all characters share the
same mean.

```{r}

```

**8.** Use `anova()` to produce the sequential ANOVA table comparing the null
model, the intermediate model, and the full model (from Question 3).

```{r}

```

  a. What are $H_0$ and $H_A$ for the first $F$-test in the table? Report the
     $F$-statistic, degrees of freedom, and $p$-value, and state your
     conclusion.

  b. What are $H_0$ and $H_A$ for the second $F$-test? Report the
     $F$-statistic, degrees of freedom, and $p$-value, and state your
     conclusion.

**9.** Use `plot()` on your full ANOVA model object from Question 3 to produce
residual diagnostic plots. Check the equal-variance and normality assumptions.
Do either appear violated?

```{r}

```

**10.** Use `pairwise.t.test()` (with no $p$-value adjustment) to obtain
unadjusted $p$-values for all pairwise comparisons of `logprop` across
characters.

```{r}

```

  a. How many pairwise comparisons are there for this dataset?
  b. Which pair has the largest $p$-value? What does that suggest about those
     two characters' dialogue share?
  c. Why is it problematic to use $\alpha = 0.05$ for each individual test
     when conducting this many comparisons?

**11.** Suppose we now want to test whether Riker and Data together have a
different mean from the remaining non-Picard characters (Beverly, Geordi,
Troi, Worf). Use `aov()` and `anova()` to carry out this test by comparing
the following two models:

  - **Reduced**: Picard has his own mean; all other characters share one mean.
  - **Full**: Picard has his own mean; Riker and Data share a mean; Beverly,
    Geordi, Troi, and Worf share a separate mean.

  Create the necessary grouping variable, fit both models, and produce the
  ANOVA table. Report the $F$-statistic and $p$-value, and state your
  conclusion.

```{r}

```

**12.** The introduction notes that the independence assumption is violated in
this dataset: within each season, if one character speaks more, the others
must speak less.

  a. Explain in your own words why this violates the ANOVA independence
     assumption.
  b. In which direction would you expect this violation to bias the $p$-values
     (too small or too large)? Explain your reasoning.
