---
title: "Confidence Interval Interpretation"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 2.7, 
                      fig.width = 4, 
                      fig.align = "center")
```

## Covering True Mean {.allowframebreaks}

```{r, echo = FALSE}
set.seed(2)
par(mar = c(2.2, 2.2, 1, 0),
    mgp = c(1.2, 0.5, 0))
mu        <- 10
sigma     <- 1
n         <- 100
simout    <- replicate(20, rnorm(n = n, mean = mu,
                                 sd = sigma))
xbar_vec  <- colMeans(simout)
s_vec     <- apply(simout, 2, sd)
lower_vec <- xbar_vec - 1.96 * s_vec / sqrt(n)
upper_vec <- xbar_vec + 1.96 * s_vec / sqrt(n)

for (plot_index in 1:length(xbar_vec)) {
  which_miss <- upper_vec < mu | lower_vec > mu
  plot(c(0), xlim = c(1, length(xbar_vec)),
       ylim = c(min(lower_vec), max(upper_vec)),
       type = "n", xlab = "sample", ylab = "value",
       main = "95% Confidence Intervals")
  arrows(x0 = 1:plot_index, y0 = lower_vec[1:plot_index],
         x1 = 1:plot_index, y1 = upper_vec[1:plot_index],
         length = 0, col = which_miss + 1, lwd = 4)
  abline(h = mu, lty = 2, col = 4)
}
```

## Correct/Incorrect Descriptions of CI

Let $l$ and $u$ be the lower and upper bounds, respectively, of a 95\% confidence interval.

What does ``With 95\% Confidence, $\mu$ is between $(l, u)$'' mean? Which interpretations are correct/incorrect?


1. The probability of $\mu$ being between $l$ and $u$ is 95\%.
2. Prior to sampling, the probability of $\mu$ being captured by our confidence interval is 95\%.
3. 95\% of the population's distribution is between $l$ and $u$.
4. If we were to draw another sample, the new $\bar{X}$ would be between $l$ and $u$ with 95\% probability.
5. 95\% of new $\bar{X}$'s would lie between $l$ and $u$.
6. We used a procedure that captures the true $\mu$ 95\% of the time in repeated samples.


