---
title: "04 Sampling Distribution"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 2.3, fig.width = 3, fig.align = "center", cols.print = 4)
```

## Learning Objectives

We will demonstrate the concept of a sampling distribution.

## Population

Suppose the population consists of 100 twins with the following histogram of differences

```{r, echo = FALSE}
set.seed(1)
library(ggplot2)
x <- rt(100, df = 10) + 1
qplot(x, bins = 20) + theme_bw()
```

## Sample
But we only observe one sample of 15 and calculated it's mean
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- mean(samp)
```

avearge = `r mean(samp)`.

## Another Sample

But we could have collected another sample
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

## Sampling Distribution

What if collect many samples and collect the corresponding sample averages in a histogram? This would be the sampling distribution of the sample average.

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

## Keep collecting samples
```{r, echo = FALSE}
samp <- sample(x, size = 15)
pander::pander(samp)
mean_vec <- c(mean_vec, mean(samp))
```

average = `r mean(samp)`.

```{r, echo=FALSE, warning=FALSE}
qplot(mean_vec, xlim = c(-1, 3), ylim = c(0, 15), bins = 30, xlab = "Sample Averages") + theme_bw()
```

