---
title: "Chapter 2 Worksheet"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Twin Study — Paired $t$-Test (Case Study 2.2.2)

Researchers measured left-hippocampus volumes (in $\text{cm}^3$) for 15 pairs
of identical twins, one of whom had schizophrenia (`Affected`) and one of
whom did not (`Unaffected`).

```{r}
#| eval: true
#| message: false
library(tidyverse)
case0202 <- read_csv("https://dcgerard.github.io/stat_302/data/case0202.csv")
glimpse(case0202)
```

**1.** Each row in `case0202` is a pair of twins rather than two independent
individuals. Why is a *paired* $t$-test more appropriate here than a
*two-sample* $t$-test?

**2.** Create a new vector `diff` containing the difference in left-hippocampus
volume for each pair (`Unaffected` minus `Affected`).

```{r}

```

**3.** Use `t.test()` to run a one-sample $t$-test on `diff` against
$H_0: \mu = 0$. Save the result as `tout` and print it.

```{r}

```

**4.** Now carry out the test manually. Compute the mean, standard deviation,
and sample size of `diff`.

```{r}

```

**5.** Compute the $t$-statistic:
$$t^* = \frac{\bar{d}}{s_d / \sqrt{n}}$$
Save it as `tstat`.

```{r}

```

**6.** What are the degrees of freedom for this test? Save it as `degfr`.

```{r}

```

**7.** Use `pt()` to compute the two-sided $p$-value from `tstat` and
`degfr`. Check that it matches the output of `t.test()` from Question 3.

```{r}

```

**8.** Manually construct the 95% confidence interval for $\mu$:
$$\bar{d} \;\pm\; t_{n-1}(0.975) \cdot \frac{s_d}{\sqrt{n}}$$
Use `qt(0.975, df = degfr)` for the critical value.

```{r}

```

**9.** Interpret the 95% confidence interval from Question 8 in the context
of this study. What does it say about the magnitude of the hippocampus
volume difference?

**10.** Based on Questions 3–9, write a complete statistical conclusion.
Address: (a) whether there is evidence of a difference, (b) the estimated
magnitude and direction of the difference, and (c) a 95% CI.

**11.** The Central Limit Theorem underpins the $t$-test. State the CLT in
your own words and explain why it allows us to use a $t$-distribution here
even if hippocampus volumes are not perfectly normally distributed in the
population.

---

# Beak Study — Two-Sample $t$-Test (Case Study 2.1.1)

Researchers measured beak depths (mm) of Galápagos finches on Daphne Major
in 1976 (before a drought) and in 1978 (after a drought that favored
larger, harder seeds).

```{r}
#| eval: true
case0201 <- read.csv("https://dcgerard.github.io/stat_302/data/case0201.csv") |>
  mutate(Year = factor(Year))
glimpse(case0201)
```

**1.** Let $\mu_1$ = population mean beak depth in 1976 and
$\mu_2$ = population mean beak depth in 1978. State the null and alternative
hypotheses in words and in mathematical notation.

**2.** Make a boxplot of beak `Depth` (y-axis) by `Year` (x-axis). Based on
the plot, do the two distributions appear to have similar spreads (variances)?
What does this suggest about which version of the $t$-test to use?

```{r}

```

**3.** Run a two-sample $t$-test assuming equal variances. Save the result
as `tout` and print it.

```{r}
t.test(formula = Depth ~ Year, data = case0201, var.equal = TRUE)
```

**4.** Write a complete statistical conclusion based on the output above.
Include the estimated difference in means, the 95% CI, and the $p$-value.

**5.** The confidence interval from Question 3 gives a range of plausible
values for $\mu_1 - \mu_2$ (1976 minus 1978). Is the interval entirely
above zero, entirely below zero, or does it straddle zero? What does that
tell you?

**6.** Suppose the researchers had a directional prediction: beaks will be
*deeper* in 1978 than in 1976 (i.e., the drought selected for larger beaks).

  a. State the one-sided alternative hypothesis in symbols.

  b. Add the appropriate argument to `t.test()` to run this one-sided test
     and report the resulting $p$-value. Is it larger or smaller than the
     two-sided $p$-value from Question 3? Explain why.

```{r}

```

**7.** By default, `t.test()` uses `var.equal = FALSE`, which applies the
**Satterthwaite approximation** to compute the degrees of freedom. In
Question 3 we overrode this with `var.equal = TRUE`. Under what conditions
is it safer to use `var.equal = FALSE`? Run the default version below and
compare the degrees of freedom and $p$-value to Question 3.

```{r}

```

**8.** In this study, finches were not randomly assigned to years — the year
was determined by when the study took place. Does this affect whether we can
claim that the drought *caused* a change in beak depth? What additional
reasoning or evidence would be needed to support a causal claim?
