---
title: "Probability Review"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
set.seed(3)
library(ggplot2)
theme_set(theme_bw(base_size = 12))
knitr::opts_chunk$set(
  echo       = FALSE,
  fig.height = 2.5,
  fig.width  = 3,
  fig.align  = "center"
)
```

# Randomness and Probability

## Example 1

:::{.panel-tabset}
## a
Toss a coin and record the result. What is the probability of heads? Tails?

## Solution
$P(\text{heads}) = 1/2 = 0.5$ and $P(\text{tails}) = 1/2 = 0.5$.
:::

:::{.panel-tabset}
## b
Roll a 6-sided die and record the result. What is the probability of getting a 1? A 2? Etc.

## Solution
$P(1) = P(2) = P(3) = P(4) = P(5) = P(6) = 1/6 \approx 0.167$.
:::

:::{.panel-tabset}
## c
Toss two coins (or one coin twice). What is the probability of getting two heads?

## Solution
The sample space is {HH, HT, TH, TT}. Since each outcome is equally likely and only
one gives HH: $P(\text{HH}) = 1/4 = 0.25$.
:::

:::{.panel-tabset}
## d
What does it mean that there is a 10% chance of rain today?

## Solution
If we could replay today's weather conditions a very large number of times, it would
rain in about 10% of those scenarios.
:::

A phenomenon is **random** if individual outcomes are uncertain, but there is nonetheless a
regular distribution of outcomes in a large number of repetitions. The **probability** of any
outcome of a random phenomenon can be defined as the proportion of times the outcome would
occur in a very long series of repetitions.

The following plot shows the running proportion of heads in two simulated sequences of 100
coin tosses. Both converge toward 0.5 as the number of tosses increases.

```{r}
#| label: convergence-plot
#| fig-width: 4
#| message: false
#| warning: false
library(tidyverse)
set.seed(3)
ntoss <- 100
series1 <- cumsum(sample(c(0, 1), size = ntoss, replace = TRUE)) / seq_len(ntoss)
series2 <- cumsum(sample(c(0, 1), size = ntoss, replace = TRUE)) / seq_len(ntoss)
dfdat <- data.frame(
  y      = c(series1, series2),
  x      = rep(seq_len(ntoss), 2),
  Series = rep(c("1", "2"), each = ntoss)
)
ggplot(dfdat, aes(x = x, y = y, color = Series)) +
  geom_line() +
  ggthemes::scale_color_colorblind() +
  geom_hline(yintercept = 0.5, lty = 2, color = "gray50") +
  xlab("Number of Tosses") +
  ylab("Proportion of Heads")
```

**Probability distributions** or **models** describe random processes and consist of two parts:

1. A list of all possible outcomes (called the **sample space**)
2. The probability of each outcome

## Example 2

:::{.panel-tabset}
## Question
Give the probability distribution for Example 1b (rolling a 6-sided die).

## Solution

| Outcome     |  1  |  2  |  3  |  4  |  5  |  6  |
|-------------|:---:|:---:|:---:|:---:|:---:|:---:|
| Probability | 1/6 | 1/6 | 1/6 | 1/6 | 1/6 | 1/6 |

:::

## Example 3

:::{.panel-tabset}
## Question
Suppose you count the number of heads after two flips of a coin. Find the probability
distribution of the number of heads.

## Solution
The sample space for two flips is {HH, HT, TH, TT}. Counting heads in each outcome:

| $X$ = # Heads |  0  |  1  |  2  |
|:--------------|:---:|:---:|:---:|
| Probability   | 1/4 | 2/4 | 1/4 |

:::

Distributions are frequently described by their histogram or density curve.

## Example 4

:::{.panel-tabset}
## Question
Suppose the probability distribution (model) of $X$ = the number of minutes that a
randomly selected student uses their cell phone (for any purpose) in a given hour has
the density shown below. Describe the distribution of $X$. (Note: $X$ is called a
**random variable**.)

```{r}
#| label: exp-histogram
set.seed(1)
x <- rexp(n = 1000, rate = 0.2)
ggplot(data = data.frame(x), mapping = aes(x = x, y = after_stat(density))) +
  geom_histogram(color = "black", fill = "white", bins = 30) +
  geom_density(color = "red", lty = 2) +
  xlab("Minutes on phone (x)") +
  ylab("f(x)")
```

## Solution
The distribution of $X$ is **right-skewed**: most students use their phone for a short
duration, but a few use it for much longer. There is a single mode near zero and a long
right tail.
:::

# Probability Distributions of a Population

- Distributions describe behavior: the values a random variable can take and how likely it
  is to take on those values.
- Probability histograms or density curves (called **pdf**, probability density function)
  provide graphical descriptions of probability distributions.
- The area under the density curve between two points gives the probability that the random
  variable falls between those values. The total area under any density curve equals 1.

**Probability Rules** — Let $A$ be an event.

- $0 \leq P(A) \leq 1$
- $P(A^c) = 1 - P(A)$, where $A^c$ is the **complement** of $A$.
- $P(A \text{ or } B) = P(A) + P(B)$, provided $A$ and $B$ are **disjoint**. Two events are
  disjoint if they have no outcomes in common and cannot occur simultaneously.

## Example 5

:::{.panel-tabset}
## Question
The event $A$ = "flip exactly two heads" and event $B$ = "flip exactly two tails."
Are $A$ and $B$ disjoint? If so, what is $P(A \text{ or } B)$?

## Solution
$A$ and $B$ are disjoint because they cannot both occur on the same pair of flips.
By the addition rule for disjoint events:
$$P(A \text{ or } B) = P(A) + P(B) = \frac{1}{4} + \frac{1}{4} = \frac{1}{2}$$
:::

- $P(A \text{ and } B) = P(A) \cdot P(B)$, provided $A$ and $B$ are **independent**. Two
  events are independent if knowing that one occurs does not change the probability that
  the other occurs.

## Example 6

:::{.panel-tabset}
## Question
Let $T$ = "get tails on the first toss" and $H$ = "get heads on the second toss."
Are $T$ and $H$ independent? If so, what is $P(T \text{ and } H)$?

## Solution
$T$ and $H$ are independent because the outcome of the first toss has no effect on the
second toss. Therefore:
$$P(T \text{ and } H) = P(T) \cdot P(H) = \frac{1}{2} \cdot \frac{1}{2} = \frac{1}{4}$$
:::

# Some Famous Distributions

## Normal

- R functions: `dnorm()`, `pnorm()`, `qnorm()`, `rnorm()`
- $\mu$ = mean (center of the distribution)
- $\sigma^2$ = variance; larger variance means more spread out.

```{r}
#| label: normal-plot
#| fig-width: 3
#| fig-height: 2
library(latex2exp)
set.seed(4)
x  <- seq(-3, 3, length = 100)
fx <- dnorm(x)
rs <- rnorm(5000)
ggplot(data.frame(rs = rs), aes(x = rs, y = after_stat(density))) +
  geom_histogram(bins = 30, fill = "red2", color = "black", alpha = 1/2) +
  geom_line(data = data.frame(x = x, fx = fx), mapping = aes(x = x, y = fx),
            linewidth = 1.2) +
  ggtitle(TeX("Normal, $X \\sim N(\\mu, \\sigma^2)$")) +
  xlab("x") +
  ylab("f(x)")
```

## t-Distribution

- R functions: `dt()`, `pt()`, `qt()`, `rt()`
- $\nu$ (or $df$) = degrees of freedom; smaller $df$ gives heavier tails.
- Mean is always 0; shape resembles the normal but with more probability in the tails.

```{r}
#| label: t-plot
#| fig-width: 3
#| fig-height: 2
x  <- seq(-3, 3, length = 100)
fx <- dt(x, df = 4)
rs <- rt(5000, df = 4)
ggplot(data.frame(rs = rs), aes(x = rs, y = after_stat(density))) +
  geom_histogram(bins = 30, fill = "blue1", color = "black", alpha = 1/2) +
  geom_line(data = data.frame(x = x, fx = fx), mapping = aes(x = x, y = fx),
            linewidth = 1.2) +
  ggtitle(TeX("t, $X \\sim t_{df}$")) +
  xlab("x") +
  ylab("f(x)")
```

## Chi-Squared

- R functions: `dchisq()`, `pchisq()`, `qchisq()`, `rchisq()`
- $\nu$ (or $df$) = degrees of freedom; smaller $df$ implies more spread out.
- Always non-negative and right-skewed, especially for small $df$.

```{r}
#| label: chisq-plot
#| fig-width: 3
#| fig-height: 2
x  <- seq(0, 15, length = 100)
fx <- dchisq(x, df = 3)
rs <- rchisq(5000, df = 3)
ggplot(data.frame(rs = rs), aes(x = rs, y = after_stat(density))) +
  geom_histogram(bins = 30, fill = "green3", color = "black", alpha = 1/2) +
  geom_line(data = data.frame(x = x, fx = fx), mapping = aes(x = x, y = fx),
            linewidth = 1.2) +
  ggtitle(TeX("Chi-squared, $X \\sim \\chi^2(\\nu)$")) +
  xlab("x") +
  ylab("f(x)")
```

## Exponential

- R functions: `dexp()`, `pexp()`, `qexp()`, `rexp()`
- $\lambda$ = rate parameter; larger $\lambda$ implies shorter typical durations.
- Often used for waiting-time questions: "How long until a light bulb fails?"

```{r}
#| label: exp-plot
#| fig-width: 3
#| fig-height: 2
x  <- seq(0.001, 3, length = 100)
fx <- dexp(x, rate = 2)
rs <- rexp(5000, rate = 2)
ggplot(data.frame(rs = rs), aes(x = rs, y = after_stat(density))) +
  geom_histogram(bins = 30, fill = "purple2", color = "black", alpha = 1/2) +
  geom_line(data = data.frame(x = x, fx = fx), mapping = aes(x = x, y = fx),
            linewidth = 1.2) +
  ggtitle(TeX("Exponential, $X \\sim Exp(\\lambda)$")) +
  xlab("x") +
  ylab("f(x)")
```

## Binomial

- R functions: `dbinom()`, `pbinom()`, `qbinom()`, `rbinom()`
- $n$ = number of trials, $p$ = probability of success on each trial.
- Use when: (i) there are $n$ independent trials, (ii) each results in success or failure,
  and (iii) the probability of success is $p$ for every trial.

```{r}
#| label: binom-plot
#| fig-width: 3
#| fig-height: 2
x  <- 0:10
fx <- dbinom(x, size = 10, prob = 0.7)
ggplot(data.frame(x = x, fx = fx), aes(x = x, xend = x, y = 0, yend = fx)) +
  geom_segment(linewidth = 1.2, color = "black") +
  ggtitle(TeX("Binomial, $X \\sim Bin(n, p)$")) +
  xlab("x") +
  ylab("Pr(x)")
```

## Uniform

- R functions: `dunif()`, `punif()`, `qunif()`, `runif()`
- $a$ = lower bound, $b$ = upper bound.
- Every value in $[a, b]$ has equal probability density.

```{r}
#| label: unif-plot
#| fig-width: 3
#| fig-height: 2
rs <- runif(5000)
ggplot(data.frame(rs = rs), aes(x = rs, y = after_stat(density))) +
  geom_histogram(bins = 30, fill = "yellow1", color = "black", alpha = 1/2) +
  geom_hline(yintercept = 1) +
  ggtitle(TeX("Uniform, $X \\sim Unif(a, b)$")) +
  xlab("x") +
  ylab("f(x)")
```

## Poisson

- R functions: `dpois()`, `ppois()`, `qpois()`, `rpois()`
- $\lambda$ = rate; both the mean and variance equal $\lambda$.
- Used for count data: "How many phone calls will we receive in the next hour?"

```{r}
#| label: pois-plot
#| fig-width: 3
#| fig-height: 2
x  <- 0:15
fx <- dpois(x, lambda = 3)
ggplot(data.frame(x = x, fx = fx), aes(x = x, xend = x, y = 0, yend = fx)) +
  geom_segment(linewidth = 1.2, color = "orange") +
  ggtitle(TeX("Poisson, $X \\sim Poi(\\lambda)$")) +
  xlab("x") +
  ylab("Pr(x)")
```

# The Normal Distribution in Depth

## Density Function

The normal distribution $N(\mu, \sigma^2)$ has density:
$$f(x \mid \mu, \sigma^2) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\!\left(-\frac{(x - \mu)^2}{2\sigma^2}\right)$$

- **Mean** $\mu$: shifts the curve left or right along the $x$-axis without changing shape.
- **Standard deviation** $\sigma$: stretches or squeezes the curve.

Normal density curves with different means (same $\sigma = 2$):

```{r}
#| label: normal-means
#| message: false
#| warning: false
#| fig-width: 4
x <- seq(-10, 10, length = 100)
dfdat <- data.frame(
  x = rep(x, 3),
  y = c(dnorm(x, mean =  0, sd = 2),
        dnorm(x, mean = -4, sd = 2),
        dnorm(x, mean =  4, sd = 2)),
  z = factor(rep(c(1, 2, 3), each = length(x)))
)
ggplot(dfdat, aes(x = x, y = y, color = z, lty = z)) +
  geom_line() +
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "none") +
  ylab("f(x)") +
  xlab("x") +
  ggtitle("Different means, same standard deviation")
```

Normal density curves with different standard deviations (same $\mu = 0$):

```{r}
#| label: normal-sds
#| message: false
#| warning: false
#| fig-width: 4
x <- seq(-10, 10, length = 100)
dfdat <- data.frame(
  x = rep(x, 3),
  y = c(dnorm(x, mean = 0, sd = 1),
        dnorm(x, mean = 0, sd = 2),
        dnorm(x, mean = 0, sd = 4)),
  z = factor(rep(c(1, 2, 3), each = length(x)))
)
ggplot(dfdat, aes(x = x, y = y, color = z, lty = z)) +
  geom_line() +
  ggthemes::scale_color_colorblind() +
  theme(legend.position = "none") +
  ylab("f(x)") +
  xlab("x") +
  ggtitle("Same mean, different standard deviations")
```

## The 68–95–99.7 Rule

For any $N(\mu, \sigma^2)$ distribution:

- Approximately **68%** of observations fall within $1\sigma$ of $\mu$
- Approximately **95%** of observations fall within $2\sigma$ of $\mu$ (exactly 1.96$\sigma$)
- Approximately **99.7%** of observations fall within $3\sigma$ of $\mu$

```{r}
#| label: rule-plot
x <- seq(-3.5, 3.5, length = 500)
y <- dnorm(x)
dat <- data.frame(x = x, y = y)
poly1 <- data.frame(x = c(-1, x[x <= 1 & x >= -1], 1, -1),
                    y = c(0, y[x <= 1 & x >= -1], 0, 0))
poly2 <- data.frame(x = c(-2, x[x <= 2 & x >= -2], 2, -2),
                    y = c(0, y[x <= 2 & x >= -2], 0, 0))
poly3 <- data.frame(x = c(-3, x[x <= 3 & x >= -3], 3, -3),
                    y = c(0, y[x <= 3 & x >= -3], 0, 0))
ggplot(dat, aes(x = x, y = y)) +
  geom_line() +
  geom_polygon(data = poly1, aes(x, y), alpha = 0.4, fill = "red") +
  geom_polygon(data = poly2, aes(x, y), alpha = 0.3, fill = "blue") +
  geom_polygon(data = poly3, aes(x, y), alpha = 0.2, fill = "green") +
  ylab("Density") +
  annotate("text", x = 0,   y = 0.20, label = "68%",   size = 3.5) +
  annotate("text", x = 1.6, y = 0.08, label = "95%",   size = 3.5) +
  annotate("text", x = 2.8, y = 0.02, label = "99.7%", size = 3.0) +
  scale_x_continuous(breaks = -3:3,
                     labels = c("-3σ", "-2σ", "-σ",
                                "μ", "+σ", "+2σ", "+3σ")) +
  xlab("")
```

## Standardization

A variable $X \sim N(\mu, \sigma^2)$ can be **standardized** to $Z \sim N(0, 1)$ by:
$$z = \frac{x - \mu}{\sigma}$$

Standardizing lets us compare observations from different normal distributions and use
a single table (the standard normal table) or R to find any probability.

## Example: Calf Circumference

Suppose the calf circumference of adults is approximately:
$$X \sim N(\mu = 37,\ \sigma^2 = 25) \quad \text{(cm, so } \sigma = 5\text{)}$$

:::{.panel-tabset}
## a
What is the probability that a randomly selected person has a calf circumference less
than 37 cm?

## Solution
Since 37 cm is the mean, and the normal distribution is symmetric about its mean:
$$P(X < 37) = 0.5$$
:::

:::{.panel-tabset}
## b
What proportion has calf circumference greater than 37 cm?

## Solution
By the complement rule:
$$P(X > 37) = 1 - P(X < 37) = 0.5$$
:::

:::{.panel-tabset}
## c
What is the probability of a calf circumference between 27 cm and 47 cm?

## Solution
Note $27 = 37 - 2(5) = \mu - 2\sigma$ and $47 = 37 + 2(5) = \mu + 2\sigma$. By the
68–95–99.7 rule:
$$P(27 < X < 47) \approx 0.95$$
:::

:::{.panel-tabset}
## d
What is the probability that two independently selected adults each have a calf
circumference greater than 47 cm?

## Solution
First, $P(X > 47) = P(X > \mu + 2\sigma) = (1 - 0.95)/2 = 0.025$.

Since the individuals are independent:
$$P(\text{both} > 47) = 0.025 \times 0.025 = 0.000625$$
:::


# Computing Probabilities in R

For $X \sim N(\mu, \sigma^2)$, `pnorm(q, mean, sd)` returns $P(X < q)$.

**Note:** R takes the standard deviation $\sigma$, not the variance $\sigma^2$.

:::{.panel-tabset}
## a
Let $X \sim N(37, 25)$. Find $P(X < 30)$.

## Solution
```{r}
#| echo: true
pnorm(30, mean = 37, sd = 5)
```
:::

:::{.panel-tabset}
## b
Let $X \sim N(37, 25)$. Find $P(30 < X < 35)$.

## Solution
$$P(30 < X < 35) = P(X < 35) - P(X < 30)$$

```{r}
#| echo: true
pnorm(35, mean = 37, sd = 5) - pnorm(30, mean = 37, sd = 5)
```
:::

Use `qnorm(p, mean, sd)` to find the $p$th quantile.

:::{.panel-tabset}
## c
At what value do 20% of individuals have a smaller calf circumference?

## Solution
We need the 20th percentile of $N(37, 25)$, which we find with `qnorm()`:

```{r}
#| echo: true
qnorm(0.20, mean = 37, sd = 5)
```

About 32.8 cm.
:::

