---
title: "Multi-Factor ANOVA"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Write out the general multi-factor ANOVA model with main effects and interactions.
- Distinguish two-way interactions from three-way interactions using interaction plots.
- Fit a multi-factor ANOVA model in R and test for interactions top-down.
- Chapter 13 of the textbook.

---

# The General Model

With **three** categorical variables ($A$, $B$, $C$):

$$Y_{ijk\ell} = \mu + \alpha_i + \beta_j + \gamma_k + (\alpha\beta)_{ij} + (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\alpha\beta\gamma)_{ijk} + \epsilon_{ijk\ell}$$

| Term | Meaning |
|:--|:--|
| $\mu$ | baseline mean |
| $\alpha_i, \beta_j, \gamma_k$ | main effects |
| $(\alpha\beta)_{ij}, (\alpha\gamma)_{ik}, (\beta\gamma)_{jk}$ | two-way interactions |
| $(\alpha\beta\gamma)_{ijk}$ | three-way interaction |

- A **two-way interaction** means the effect of one variable depends on the level of another.
- A **three-way interaction** means the two-way interaction between $A$ and $B$ itself
  changes depending on the level of $C$.

---

# Interaction Plots for Three Factors

With three factors, interaction plots use:

- **x-axis**: one variable.
- **Color/line**: a second variable.
- **Facets**: the third variable.

The key question: do the lines look parallel within each facet, and do the patterns look the same across facets?

## No Interactions

Parallel lines, same pattern in every facet — no interactions at all:

```{r}
#| label: no-interaction
#| echo: false
x <- c(1, 1, 1, 1, 0, 0, 0, 0)
y <- c(1, 1, 0, 0, 1, 1, 0, 0)
z <- c(1, 0, 1, 0, 1, 0, 1, 0)
w <- x + y + z
tibble(x = factor(x), y = factor(y), z = factor(z, labels = c("z=0","z=1")), w) |>
  ggplot(aes(x = x, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~z) +
  labs(title = "No interactions: parallel lines in all facets",
       x = "x", y = "Mean response", color = "y")
```

## Two-Way Interaction Between $x$ and $y$ Only

When only $x$ and $y$ interact (but $z$ has no interaction with either), we can check
different arrangements of the three variables across the axes and facets.

**Facets = $z$, color = $y$:** Non-parallel lines within each facet, but the **same** non-parallel
pattern in every facet. The $x$:$y$ interaction is the same regardless of $z$.

```{r}
#| label: two-way-xy-1
#| echo: false
w2 <- x + y + z + x * y
tibble(x = factor(x), y = factor(y), z = factor(z, labels = c("z=0","z=1")), w = w2) |>
  ggplot(aes(x = x, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~z) +
  labs(title = "x:y interaction only — same non-parallel pattern across z facets",
       x = "x", y = "Mean response", color = "y")
```

**Facets = $y$, color = $z$:** Parallel lines within each facet ($x$ and $z$ do not interact),
but the gap between lines shifts across facets of $y$ (because $x$:$y$ interact, so the effect
of $x$ depends on $y$).

```{r}
#| label: two-way-xy-2
#| echo: false
tibble(x = factor(x), y = factor(y, labels = c("y=0","y=1")), z = factor(z), w = w2) |>
  ggplot(aes(x = x, y = w, color = z, group = z)) +
  geom_point() + geom_line() +
  facet_wrap(~y) +
  labs(title = "x:y interaction only — parallel lines (no x:z), but gap varies by y facet",
       x = "x", y = "Mean response", color = "z")
```

**Facets = $y$, x-axis = $z$, color = $x$:** $x$ and $z$ do not interact, so lines are parallel
within each facet and the pattern is the same across facets of $y$.

```{r}
#| label: two-way-xy-3
#| echo: false
tibble(x = factor(x), y = factor(y, labels = c("y=0","y=1")), z = factor(z), w = w2) |>
  ggplot(aes(x = z, y = w, color = x, group = x)) +
  geom_point() + geom_line() +
  facet_wrap(~y) +
  labs(title = "x:y interaction only — no x:z interaction visible here",
       x = "z", y = "Mean response", color = "x")
```

**Facets = $x$, x-axis = $z$, color = $y$:** $y$ and $z$ do not interact, so lines are parallel.
But the gap between $y$ lines differs across facets of $x$ — again reflecting the $x$:$y$ interaction.

```{r}
#| label: two-way-xy-4
#| echo: false
tibble(x = factor(x, labels = c("x=0","x=1")), y = factor(y), z = factor(z), w = w2) |>
  ggplot(aes(x = z, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~x) +
  labs(title = "x:y interaction only — parallel lines (no y:z), gap changes across x facets",
       x = "z", y = "Mean response", color = "y")
```

## Two-Way Interactions Between All Pairs (No Three-Way)

When all three pairs interact but there is no three-way interaction, non-parallel lines appear
within each facet, **and** the non-parallel pattern changes somewhat across facets — but
the change itself is consistent (not dependent on the third variable in a deeper sense).

```{r}
#| label: two-way-all
#| echo: false
w_all2 <- x + y + z + x * y + x * z + y * z
tibble(x = factor(x), y = factor(y), z = factor(z, labels = c("z=0","z=1")), w = w_all2) |>
  ggplot(aes(x = x, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~z) +
  scale_y_continuous(breaks = 1:6) +
  labs(title = "All two-way interactions, no three-way",
       x = "x", y = "Mean response", color = "y")
```

## Three-Way Interaction

Non-parallel lines **and** the interaction pattern itself changes across facets:

```{r}
#| label: three-way
#| echo: false
w3 <- x + y + z + x * y + x * z + y * z + x * y * z
tibble(x = factor(x), y = factor(y), z = factor(z, labels = c("z=0","z=1")), w = w3) |>
  ggplot(aes(x = x, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~z) +
  scale_y_continuous(breaks = 1:7) +
  labs(title = "Three-way interaction — interaction pattern differs across facets",
       x = "x", y = "Mean response", color = "y")
```

- The relationship between $x$ and $y$ is **different** at different levels of $z$.

---

# Case Study: Poker Winnings

**Research question:** Does skill level or hand quality affect poker winnings? Does the type of game (fixed vs. no limit) interact with skill? These data come from @meyer2013poker.

**Variables:**

- `Skill`: `"Expert"` or `"Average"` — skill level of the player.
- `Hand`: `"Bad"`, `"Neutral"`, or `"Good"` — quality of the cards dealt.
- `Limit`: `"Fixed"` or `"None"` — whether the game has a fixed betting limit.
- `Cash`: final cash balance (in dollars) after a session of play.

```{r}
#| label: poker-load
#| message: false
poker <- read_csv("https://dcgerard.github.io/stat_302/data/poker.csv") |>
  mutate(Hand = factor(Hand, levels = c("Bad", "Neutral", "Good")))
glimpse(poker)
```

```{r}
#| label: poker-table
table(poker$Skill, poker$Hand, poker$Limit)
```

The design is balanced: equal sample sizes across all 12 combinations of the three factors.

## Interaction Plots

With three factors we have $3 \times 2 = 6$ ways to assign variables to axes and facets.
Looking at all of them gives a fuller picture of which interactions may be present.

```{r}
#| label: poker-plot-1
#| fig-width: 7
ggplot(poker, aes(x = Skill, y = Cash, color = Hand, group = Hand)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_wrap(~Limit, labeller = label_both) +
  labs(x = "Skill", y = "Mean cash ($)", color = "Hand",
       title = "x = Skill, color = Hand, facets = Limit")
```

```{r}
#| label: poker-plot-2
#| fig-width: 7
ggplot(poker, aes(x = Hand, y = Cash, color = Skill, group = Skill)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_wrap(~Limit, labeller = label_both) +
  labs(x = "Hand", y = "Mean cash ($)", color = "Skill",
       title = "x = Hand, color = Skill, facets = Limit")
```

```{r}
#| label: poker-plot-3
#| fig-width: 7
ggplot(poker, aes(x = Limit, y = Cash, color = Skill, group = Skill)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_wrap(~Hand, labeller = label_both) +
  labs(x = "Limit", y = "Mean cash ($)", color = "Skill",
       title = "x = Limit, color = Skill, facets = Hand")
```

```{r}
#| label: poker-plot-4
#| fig-width: 7
ggplot(poker, aes(x = Skill, y = Cash, color = Limit, group = Limit)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_wrap(~Hand, labeller = label_both) +
  labs(x = "Skill", y = "Mean cash ($)", color = "Limit",
       title = "x = Skill, color = Limit, facets = Hand")
```

```{r}
#| label: poker-plot-5
#| fig-width: 7
ggplot(poker, aes(x = Hand, y = Cash, color = Limit, group = Limit)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_wrap(~Skill, labeller = label_both) +
  labs(x = "Hand", y = "Mean cash ($)", color = "Limit",
       title = "x = Hand, color = Limit, facets = Skill")
```

```{r}
#| label: poker-plot-6
#| fig-width: 7
ggplot(poker, aes(x = Limit, y = Cash, color = Hand, group = Hand)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_wrap(~Skill, labeller = label_both) +
  labs(x = "Limit", y = "Mean cash ($)", color = "Hand",
       title = "x = Limit, color = Hand, facets = Skill")
```

**Observations from the plots:**

- The effect of `Hand` on `Cash` appears to differ between experts and average players
  (non-parallel lines when Skill is on the facet or color), suggesting a Skill × Hand interaction.
- The effect of `Limit` also appears to differ between skill levels, suggesting a
  Skill × Limit interaction.
- The Hand × Limit interaction looks less clear.

## Top-Down Testing Strategy

Start with the **full interaction model** (all main effects + all interactions). Test
the **three-way interaction** first; if not significant, drop it and test the two-way
interactions.

```{r}
#| label: poker-full
aout_full <- aov(Cash ~ Skill * Hand * Limit, data = poker)
tidy(aout_full)
```

- The three-way interaction `Skill:Hand:Limit` is not significant — drop it.

```{r}
#| label: poker-twoway
aout_2way <- aov(Cash ~ Skill + Hand + Limit +
                   Skill:Hand + Skill:Limit + Hand:Limit,
                 data = poker)
tidy(aout_2way)
```

- `Skill:Hand` and `Skill:Limit` are significant; `Hand:Limit` is not.
- Drop `Hand:Limit` and adopt the model with only those two interactions.

```{r}
#| label: poker-sub
aout_sub <- aov(Cash ~ Skill + Hand + Limit + Skill:Hand + Skill:Limit,
                data = poker)
tidy(aout_sub)
```

## Residual Plot

```{r}
#| label: poker-resid
#| warning: false
augment(aout_sub) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, lty = 2, color = "red") +
  labs(x = "Fitted values", y = "Residuals")
```

No obvious patterns — the equal-variance assumption looks reasonable.

## Coefficient Estimates

```{r}
#| label: poker-coef
lm_sub <- lm(Cash ~ Skill + Hand + Limit + Skill:Hand + Skill:Limit,
             data = poker)
tidy(lm_sub)
```

Using the notation $\alpha$ for Skill, $\beta$ for Hand, $\gamma$ for Limit (reference levels:
Average, Bad, Fixed):

| Parameter | Level | Estimate |
|:--|:--|--:|
| $\mu$ | — | 5.400 |
| $\alpha_1$ | Average | 0 (reference) |
| $\alpha_2$ | Expert | 1.404 |
| $\beta_1$ | Bad | 0 (reference) |
| $\beta_2$ | Neutral | 5.180 |
| $\beta_3$ | Good | 9.270 |
| $\gamma_1$ | Fixed | 0 (reference) |
| $\gamma_2$ | None | −1.910 |
| $(\alpha\beta)_{11}$ | Average × Bad | 0 (reference) |
| $(\alpha\beta)_{12}$ | Average × Neutral | 0 (reference) |
| $(\alpha\beta)_{13}$ | Average × Good | 0 (reference) |
| $(\alpha\beta)_{21}$ | Expert × Bad | 0 (reference) |
| $(\alpha\beta)_{22}$ | Expert × Neutral | −1.440 |
| $(\alpha\beta)_{23}$ | Expert × Good | −4.124 |
| $(\alpha\gamma)_{11}$ | Average × Fixed | 0 (reference) |
| $(\alpha\gamma)_{12}$ | Average × None | 0 (reference) |
| $(\alpha\gamma)_{21}$ | Expert × Fixed | 0 (reference) |
| $(\alpha\gamma)_{22}$ | Expert × None | 2.520 |

## Interpretations

**Skill × Hand interaction:** How much does hand quality matter?

The mean difference between players with **good hands** and players with **bad hands** depends
on skill level. For **expert** players:

$$\mu_{\text{Expert, Good}} - \mu_{\text{Expert, Bad}} = \beta_3 + (\alpha\beta)_{23} = 9.270 - 4.124 = 5.146$$

For **average** players (where $(\alpha\beta)_{13} = 0$ by definition):

$$\mu_{\text{Average, Good}} - \mu_{\text{Average, Bad}} = \beta_3 + (\alpha\beta)_{13} = 9.270 + 0 = 9.270$$

Average players gain **more** from receiving a good hand than experts do. Experts can partially
compensate for a bad hand through skill, so hand quality matters less to them.

**Skill × Limit interaction:** Who benefits from a no-limit game?

The mean difference between no-limit and fixed-limit games by skill level:

$$\mu_{\text{Average, None}} - \mu_{\text{Average, Fixed}} = \gamma_2 = -1.910$$

$$\mu_{\text{Expert, None}} - \mu_{\text{Expert, Fixed}} = \gamma_2 + (\alpha\gamma)_{22} = -1.910 + 2.520 = 0.610$$

On average, removing the betting limit **hurts** average players (−\$1.91) but **helps** experts
(+\$0.61). The no-limit format rewards skill: experts can leverage their advantage with larger bets.

---

# R Formula Quick Reference

| Formula | Interpretation |
|:--|:--|
| `A + B + C` | main effects only (additive) |
| `A * B + C` | A, B, C, and A:B |
| `A * B * C` | all main effects and all interactions |
| `(A + B + C)^2` | all main effects and all two-way interactions |
| `A:B` | A×B interaction term only (no main effects) |

---

# Exercises

**1.** Using the `npk` data (built into R — run `data(npk)` to load):

a. Create an interaction plot of `yield` vs. `N`, colored by `P`, faceted by `K`.
   Does the plot suggest any N × P interaction?

b. Create an interaction plot of `yield` vs. `N`, colored by `K`, faceted by `P`.
   Does this plot suggest any N × K interaction?

c. Formally test whether the three-way interaction `N:P:K` is significant using
   `tidy(aov(yield ~ N * P * K, data = npk))`. Report the $p$-value and conclusion.

d. Drop non-significant two-way interactions and interpret the significant main effects
   in plain language.

---

**2.** Explain in plain language the difference between a two-way and a three-way
interaction. Give an example from everyday life (not from this lecture) of:

a. A plausible two-way interaction.

b. A plausible three-way interaction.

---

**3.** An experiment studies the effects of exercise type (Cardio vs. Strength),
diet (Low-carb vs. High-carb), and sleep quality (Poor vs. Good) on weight loss.

a. Write out the full interaction ANOVA model using the notation from this lecture.
   Identify which terms are main effects, which are two-way interactions, and which
   is the three-way interaction.

b. If you were conducting this study, what testing strategy would you use?
   Start with what model and what test?

c. How many parameters does the full interaction model have (including the baseline)?
   How many parameters does the additive model have?

---

**4.** In the poker case study, the fitted model included `Skill:Hand` and `Skill:Limit`
interactions.

a. Using the coefficient table above, compute the estimated mean cash for an expert
   player with a good hand in a no-limit game.

b. Compute the estimated mean cash for an average player with a neutral hand in a
   fixed-limit game.

c. Based on the coefficients, does having no limit help or hurt average players?
   By how much?
   
# References
