---
title: "Multi-Factor ANOVA"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Write out the general multi-factor ANOVA model with main effects and interactions.
- Distinguish two-way interactions from three-way interactions using interaction plots.
- Fit a multi-factor ANOVA model in R and test for interactions top-down.
- Chapter 13 of the textbook.

---

# The General Model

With **three** categorical variables ($A$, $B$, $C$):

$$Y_{ijk\ell} = \mu + \alpha_i + \beta_j + \gamma_k + (\alpha\beta)_{ij} + (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\alpha\beta\gamma)_{ijk} + \epsilon_{ijk\ell}$$

| Term | Meaning |
|:--|:--|
| $\mu$ | baseline mean |
| $\alpha_i, \beta_j, \gamma_k$ | main effects |
| $(\alpha\beta)_{ij}, (\alpha\gamma)_{ik}, (\beta\gamma)_{jk}$ | two-way interactions |
| $(\alpha\beta\gamma)_{ijk}$ | three-way interaction |

- A **two-way interaction** means the effect of one variable depends on the level of another.
- A **three-way interaction** means the two-way interaction between $A$ and $B$ itself
  changes depending on the level of $C$.

---

# Interaction Plots for Three Factors

With three factors, interaction plots use:

- **x-axis**: one variable.
- **Color/line**: a second variable.
- **Facets**: the third variable.

The key question for each arrangement: do the lines look parallel within each facet,
and do the patterns look the same across facets?

## No Interactions

Parallel lines, same pattern in every facet:

```{r}
#| label: no-interaction
#| echo: false
x <- c(1, 1, 1, 1, 0, 0, 0, 0)
y <- c(1, 1, 0, 0, 1, 1, 0, 0)
z <- c(1, 0, 1, 0, 1, 0, 1, 0)
w <- x + y + z
tibble(x = factor(x), y = factor(y), z = factor(z, labels = c("z=0","z=1")), w) |>
  ggplot(aes(x = x, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~z) +
  labs(title = "No interactions: parallel lines in all facets",
       x = "x", y = "Mean response", color = "y")
```

## Two-Way Interaction Between $x$ and $y$ Only

Non-parallel lines within each facet, but the **same** non-parallel pattern in every facet:

```{r}
#| label: two-way-xy
#| echo: false
w2 <- x + y + z + x * y
tibble(x = factor(x), y = factor(y), z = factor(z, labels = c("z=0","z=1")), w = w2) |>
  ggplot(aes(x = x, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~z) +
  labs(title = "x:y interaction only — same pattern across facets",
       x = "x", y = "Mean response", color = "y")
```

- The interaction between $x$ and $y$ is the **same** regardless of the level of $z$.
- No three-way interaction.

## Three-Way Interaction

Non-parallel lines, and the **interaction pattern changes** across facets:

```{r}
#| label: three-way
#| echo: false
w3 <- x + y + z + x * y + x * z + y * z + x * y * z
tibble(x = factor(x), y = factor(y), z = factor(z, labels = c("z=0","z=1")), w = w3) |>
  ggplot(aes(x = x, y = w, color = y, group = y)) +
  geom_point() + geom_line() +
  facet_wrap(~z) +
  labs(title = "Three-way interaction — pattern differs across facets",
       x = "x", y = "Mean response", color = "y")
```

- The relationship between $x$ and $y$ is **different** at different levels of $z$.
- This is a three-way interaction.

---

# Case Study: Pea Yield

- Data: `npk` (built into R), from a split-plot agricultural experiment.
- Three binary fertilizers applied to peas: **N** (nitrogen), **P** (phosphate), **K** (potassium).
- Response: `yield` in pounds per plot.
- Question: Do N, P, K, and their interactions affect pea yield?

```{r}
#| label: npk-load
data(npk)
glimpse(npk)
```

```{r}
#| label: npk-table
table(npk$N, npk$P, npk$K)
```

- Balanced design: 3 observations per combination.

## Interaction Plots

```{r}
#| label: npk-plot-np
#| fig-width: 7
ggplot(npk, aes(x = N, y = yield, color = P, group = P)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  facet_wrap(~K, labeller = label_both) +
  labs(x = "Nitrogen (0 = absent, 1 = present)",
       y = "Mean yield (lbs)",
       color = "Phosphate",
       title = "Interaction plot: N × P, faceted by K")
```

- Lines are roughly parallel within facets — no obvious N×P interaction.
- Yields are slightly higher when N is present (lines shifted up at N=1).
- K does not appear to change the N×P pattern much — no strong three-way interaction.

## Top-Down Testing Strategy

- Start with the **full interaction model** (all main effects + all interactions).
- Test the **three-way interaction** first. If not significant, drop it.
- Then test **two-way interactions**. If none significant, use the additive model.
- Then test **main effects**.

```{r}
#| label: npk-full
aout_full <- aov(yield ~ N * P * K, data = npk)
tidy(aout_full)
```

- The three-way interaction `N:P:K` has $p = 0.54$ — no evidence.
- Drop it and refit with only two-way interactions.

```{r}
#| label: npk-twoway
aout_2way <- aov(yield ~ N + P + K + N:P + N:K + P:K, data = npk)
tidy(aout_2way)
```

- None of the two-way interactions are significant ($p > 0.05$).
- Adopt the additive model.

```{r}
#| label: npk-additive
aout_add <- aov(yield ~ N + P + K, data = npk)
tidy(aout_add)
```

- **Nitrogen** ($p = 0.004$): strong evidence that nitrogen increases yield.
- **Phosphate** ($p = 0.057$): marginal evidence.
- **Potassium** ($p = 0.028$): moderate evidence that potassium **decreases** yield.

## Coefficient Estimates

```{r}
#| label: npk-coef
lm_add <- lm(yield ~ N + P + K, data = npk)
tidy(lm_add, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high, p.value)
```

- Adding nitrogen is estimated to increase yield by about 5.6 lbs per plot
  (95% CI: 2.0 to 9.3 lbs).
- Adding potassium is estimated to decrease yield by about 3.9 lbs per plot
  (95% CI: −7.6 to −0.3 lbs).

---

# R Formula Quick Reference

| Formula | Interpretation |
|:--|:--|
| `A + B + C` | main effects only (additive) |
| `A * B + C` | A, B, C, and A:B |
| `A * B * C` | all main effects and all interactions |
| `(A + B + C)^2` | all main effects and all two-way interactions |
| `A:B` | A×B interaction term only (no main effects) |

---

# Exercises

**1.** Using the `npk` data:

a. Create an interaction plot of `yield` vs. `N`, colored by `K`, faceted by `P`.
   Does the plot suggest any interaction between N and K?

b. Formally test whether the three-way interaction `N:P:K` is significant using
   `tidy(aov(yield ~ N * P * K, data = npk))`. Report $p$-value and conclusion.

c. The additive model fitted in this lecture found significant effects of N and K.
   Interpret both coefficients in plain language. Do the signs of the effects make
   agricultural sense?

---

**2.** Explain in plain language the difference between a two-way and a three-way
interaction. Give an example from everyday life (not from this lecture) of:

a. A plausible two-way interaction.

b. A plausible three-way interaction.

---

**3.** An experiment studies the effects of exercise type (Cardio vs. Strength),
diet (Low-carb vs. High-carb), and sleep quality (Poor vs. Good) on weight loss.

a. Write out the full interaction ANOVA model using the notation from this lecture.
   Identify which terms are main effects, which are two-way interactions, and which
   is the three-way interaction.

b. If you were conducting this study, what testing strategy would you use?
   Start with what model and what test?

c. How many parameters does the full interaction model have (including the baseline)?
   How many parameters does the additive model have?
