---
title: "Two-Way ANOVA"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Know when two-way ANOVA is appropriate.
- Write out the additive model and the interaction model.
- Visually understand how the F-test for main effects works (reduced vs. full model residuals).
- Test for main effects and interactions using F-tests.
- Obtain coefficient estimates and confidence intervals using `tidy()`.
- Chapter 13 of the textbook.

---

# Case Study: Pygmalion Effect

- **Pygmalion Effect**: high expectations from a supervisor translate into improved
  performance of a subordinate.
- A study used U.S. Army companies, each containing several platoons.
- Within each company, one platoon was randomly designated the "Pygmalion platoon."
- The commanding officer of the Pygmalion platoon was told (falsely) that his platoon
  had been predicted to be superior soldiers.
- At the end of basic training, all soldiers took a skill test.
- Data: **average score** for each platoon. Company is a nuisance variable (grouping variable).

```{r}
#| label: load-data
#| message: false
case1302 <- read_csv("https://dcgerard.github.io/stat_302/data/case1302.csv") |>
  mutate(Company = factor(Company),
         Treat   = factor(Treat, levels = c("Control", "Pygmalion")),
         com_num = as.numeric(Company),
         lower   = com_num - 0.2,
         upper   = com_num + 0.2)
glimpse(case1302)
```

---

# When to Use Two-Way ANOVA

1. **Quantitative** response variable.
2. **Two categorical** explanatory variables.
3. It is called *two-way* because each observational unit can be placed into a
   two-way table according to its group in both categorical variables.

|           | C1 | C2 | $\cdots$ | C10 |
|-----------|----|----|----------|-----|
| Pygmalion |    |    |          |     |
| Control   |    |    |          |     |

---

# The Models

## The Additive Model (Main Effects Only)

$$Y_{ijk} = \mu + \alpha_i + \beta_j + \epsilon_{ijk}$$

- $\mu$: baseline mean (Company C1, Control group).
- $\alpha_i$: **main effect** of Company $i$ — accounts for between-company variability.
- $\beta_j$: **main effect** of Treatment $j$ — what we actually care about.
- $\epsilon_{ijk} \sim N(0, \sigma^2)$: noise.
- **Additivity assumption**: the treatment effect $\beta_j$ is the **same** for every company.

## The Interaction Model (Cell Means)

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}$$

- $(\alpha\beta)_{ij}$: **interaction** — allows the treatment effect to differ by company.
- Equivalent to letting every company × treatment combination have its own mean.
- In R: `Score ~ Company * Treat`.

---

# Testing for Main Effects

## The Research Question

- **Research question**: Does the Pygmalion treatment improve scores?
- We need to test $H_0: \beta_j = 0$ for all $j$ (treatment has no effect).
- Equivalently: does the **full additive model** fit better than the **reduced model**
  (Company only)?

$$\text{Reduced: } Y_{ijk} = \mu + \alpha_i + \epsilon_{ijk}$$
$$\text{Full: } Y_{ijk} = \mu + \alpha_i + \beta_j + \epsilon_{ijk}$$

## Reduced Model: Company Only

- Under the reduced model, each company has one estimated mean for all platoons.

```{r}
#| label: reduced-fit
#| echo: false
aout_com <- aov(Score ~ Company, data = case1302)
case1302$est_com <- fitted(aout_com)

ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = est_com, yend = est_com),
               color = "black") +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Reduced model: one mean per company")
```

The residuals under the reduced model — gaps between points and their company mean:

```{r}
#| label: reduced-resid
#| echo: false
ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = est_com, yend = est_com),
               color = "black") +
  geom_segment(aes(x = com_num, xend = com_num, y = Score, yend = est_com),
               color = "steelblue", alpha = 0.6, linewidth = 1) +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Reduced model: residuals (blue)")
```

Notice the residuals tend to be **signed by treatment**: Pygmalion points above their
company mean, Control points below. This suggests treatment matters.

## Full (Additive) Model: Company + Treatment

- Under the additive model, each company gets **two** estimated means — one per treatment.
- The treatment gap is the same for every company (additivity).

```{r}
#| label: full-fit
#| echo: false
aout_add <- aov(Score ~ Company + Treat, data = case1302)
case1302$est_add <- fitted(aout_add)

ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = est_add, yend = est_add)) +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Full model: two means per company (same gap everywhere)")
```

The residuals under the additive model are much smaller — the systematic
treatment pattern has been absorbed by $\beta_j$:

```{r}
#| label: full-resid
#| echo: false
ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = est_add, yend = est_add)) +
  geom_segment(aes(x = com_num, xend = com_num, y = Score, yend = est_add),
               color = "steelblue", alpha = 0.6, linewidth = 1) +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Full model: residuals (blue) — much smaller")
```

## Estimated Treatment Effects

The arrows show the Control → Pygmalion gap at each company's estimated means.
Under the additive model, all arrows are forced to be **the same length** — this is
exactly what $\hat{\beta}_j$ being a single constant means:

```{r}
#| label: arrows-additive
#| echo: false
arrow_add <- case1302 |>
  group_by(Company, com_num) |>
  summarize(Control   = mean(est_add[Treat == "Control"]),
            Pygmalion = mean(est_add[Treat == "Pygmalion"]),
            lower = first(lower), upper = first(upper),
            .groups = "drop")

ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point(alpha = 0) +
  geom_segment(aes(x = lower, xend = upper, y = est_add, yend = est_add)) +
  geom_segment(data = arrow_add,
               aes(x = com_num, xend = com_num, y = Control, yend = Pygmalion),
               color = "black", linewidth = 0.7,
               arrow = arrow(length = unit(0.12, "inches")),
               inherit.aes = FALSE) +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Additive model: equal treatment effect at every company")
```

## The F-Statistic

```{r}
#| label: rss-calc
#| echo: false
n        <- nrow(case1302)
I_val    <- nlevels(case1302$Company)
J_val    <- nlevels(case1302$Treat)
rss_full <- sum(aout_add$residuals^2)
df_full  <- n - I_val - J_val + 1
rss_red  <- sum(aout_com$residuals^2)
df_red   <- n - I_val
ess      <- rss_red - rss_full
df_extra <- df_red - df_full
f_stat   <- (ess / df_extra) / (rss_full / df_full)
```

| Quantity | Formula | Value |
|:--|:--|--:|
| $RSS_{\text{full}}$ | | `r round(rss_full, 2)` |
| $df_{\text{full}}$ | $n - I - J + 1 = `r n` - `r I_val` - `r J_val` + 1$ | `r df_full` |
| $RSS_{\text{reduced}}$ | | `r round(rss_red, 2)` |
| $df_{\text{reduced}}$ | $n - I = `r n` - `r I_val`$ | `r df_red` |

$$ESS = RSS_{\text{reduced}} - RSS_{\text{full}} = `r round(rss_red, 2)` - `r round(rss_full, 2)` = `r round(ess, 2)`$$

$$df_{\text{extra}} = df_{\text{reduced}} - df_{\text{full}} = `r df_red` - `r df_full` = `r df_extra`$$

$$F = \frac{ESS / \, df_{\text{extra}}}{RSS_{\text{full}} / \, df_{\text{full}}} = \frac{`r round(ess, 2)` / `r df_extra`}{`r round(rss_full, 2)` / `r df_full`} = `r round(f_stat, 2)`$$

Compare the observed $F$ to an $F_{`r df_extra`, `r df_full`}$ distribution — the shaded area
is the $p$-value:

```{r}
#| label: f-density-plot
#| echo: false
f_seq  <- seq(0, 12, length.out = 500)
f_dens <- df(f_seq, df1 = df_extra, df2 = df_full)
p_val  <- pf(f_stat, df1 = df_extra, df2 = df_full, lower.tail = FALSE)

tail_x <- f_seq[f_seq >= f_stat]
tail_y <- df(tail_x, df1 = df_extra, df2 = df_full)

ggplot(tibble(x = f_seq, y = f_dens), aes(x, y)) +
  geom_line() +
  geom_ribbon(data = tibble(x = tail_x, y = tail_y),
              aes(x = x, ymin = 0, ymax = y),
              fill = "steelblue", alpha = 0.7, inherit.aes = FALSE) +
  geom_vline(xintercept = f_stat, linetype = "dashed") +
  annotate("text", x = f_stat + 0.35, y = 0.25,
           label = paste0("F = ", round(f_stat, 2), "\np = ", round(p_val, 3)),
           hjust = 0, size = 3.5) +
  labs(x = paste0("F(", df_extra, ", ", df_full, ")"),
       y = "Density",
       title = paste0("p-value = shaded area to the right of F = ",
                      round(f_stat, 2)))
```

## F-Test in R

```{r}
#| label: main-effects-test
tidy(aout_add)
```

- The `Treat` row is the F-test for the main effect of treatment controlling for Company.
- $F = 7.84$ on 1 and 18 df, $p = 0.012$.
- **Conclusion**: moderate evidence that the Pygmalion treatment increased average scores.

Or equivalently, compare the two models directly:

```{r}
#| label: main-effects-anova
anova(aout_com, aout_add)
```

---

# Testing for Interactions

## But Wait — Is the Additivity Assumption Valid?

- The test above assumes the treatment effect is the **same for every company**.
- What if the effect of the Pygmalion treatment depends on which company?
- We should check this assumption — i.e., test for **interactions**.

## Interaction Plots

- An **interaction plot** shows the group means connected by lines.
- X-axis: levels of one variable. Color: levels of the other variable.
- **Parallel lines** $\Rightarrow$ additive model is plausible.
- **Non-parallel lines** $\Rightarrow$ interaction may be present.

**Additive (parallel lines):**

```{r}
#| label: additive-example
#| echo: false
meanvalsc <- c(70, 77, 71, 64, 76, 72, 65, 68, 67, 74)
tibble(
  score   = c(meanvalsc, meanvalsc + 8),
  group   = rep(c("Control", "Pygmalion"), each = length(meanvalsc)),
  company = rep(seq_along(meanvalsc), 2)
) |>
  ggplot(aes(x = company, y = score, color = group, group = group)) +
  geom_point() + geom_line() +
  scale_x_continuous(breaks = seq_along(meanvalsc)) +
  labs(title = "Additive: parallel lines", y = "Mean score", x = "Company",
       color = "Treatment")
```

**Non-additive (non-parallel lines):**

```{r}
#| label: interaction-example
#| echo: false
effect_vec <- c(-2, 1, 3, 5, 8, 9, 1, 5, -4, 1)
tibble(
  score   = c(meanvalsc, meanvalsc + effect_vec),
  group   = rep(c("Control", "Pygmalion"), each = length(meanvalsc)),
  company = rep(seq_along(meanvalsc), 2)
) |>
  ggplot(aes(x = company, y = score, color = group, group = group)) +
  geom_point() + geom_line() +
  scale_x_continuous(breaks = seq_along(meanvalsc)) +
  labs(title = "Non-additive: non-parallel lines", y = "Mean score", x = "Company",
       color = "Treatment")
```

## Interaction Plot for the Pygmalion Data

```{r}
#| label: interaction-plot-pygmalion
ggplot(case1302, aes(x = Company, y = Score, color = Treat, group = Treat)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(x = "Company", y = "Average Score", color = "Treatment",
       title = "Pygmalion: interaction plot")
```

- Lines are roughly parallel — additivity is plausible.

## Interpreting the Interaction Model

Under the **cell means model**, each company × treatment combination gets its own
estimated mean:

```{r}
#| label: cellmeans-fit
#| echo: false
aout_int <- aov(Score ~ Company * Treat, data = case1302)
case1302$est_int <- fitted(aout_int)

ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = est_int, yend = est_int)) +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Cell means model: each cell has its own mean")
```

The arrows make the company-specific treatment effects explicit — each company gets
its own arrow length:

```{r}
#| label: arrows-cellmeans
#| echo: false
arrow_int <- case1302 |>
  group_by(Company, com_num) |>
  summarize(Control   = mean(est_int[Treat == "Control"]),
            Pygmalion = mean(est_int[Treat == "Pygmalion"]),
            lower = first(lower), upper = first(upper),
            .groups = "drop")

ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point(alpha = 0) +
  geom_segment(aes(x = lower, xend = upper, y = est_int, yend = est_int)) +
  geom_segment(data = arrow_int,
               aes(x = com_num, xend = com_num, y = Control, yend = Pygmalion),
               color = "black", linewidth = 0.7,
               arrow = arrow(length = unit(0.12, "inches")),
               inherit.aes = FALSE) +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Cell means model: treatment effect varies by company")
```

The residuals under the cell means model are smaller than under the additive model —
each company's idiosyncratic effect is absorbed:

```{r}
#| label: cellmeans-resid
#| echo: false
ggplot(case1302, aes(x = com_num, y = Score, color = Treat)) +
  geom_point() +
  geom_segment(aes(x = lower, xend = upper, y = est_int, yend = est_int)) +
  geom_segment(aes(x = com_num, xend = com_num, y = Score, yend = est_int),
               color = "steelblue", alpha = 0.6, linewidth = 1) +
  scale_x_continuous(breaks = 1:10, labels = levels(case1302$Company)) +
  labs(x = "Company", y = "Score", color = "Treatment",
       title = "Cell means model: residuals (blue)")
```

## Hypotheses

$$H_0: (\alpha\beta)_{ij} = 0 \text{ for all } i, j$$
$$H_A: \text{at least one } (\alpha\beta)_{ij} \neq 0$$

## Degrees of Freedom

| Model | Parameters | Residual df |
|:--|:--|:--|
| Full (cell means, with interaction) | $I \times J$ | $n - IJ$ |
| Reduced (additive, no interaction) | $I + J - 1$ | $n - I - J + 1$ |
| **Extra** | $IJ - I - J + 1$ | — |

## The F-Statistic

```{r}
#| label: rss-calc-int
#| echo: false
rss_int      <- sum(aout_int$residuals^2)
df_int       <- nrow(case1302) - nlevels(case1302$Company) * nlevels(case1302$Treat)
ess_int      <- rss_full - rss_int   # rss_full is from additive model (computed earlier)
df_extra_int <- df_full - df_int
f_stat_int   <- (ess_int / df_extra_int) / (rss_int / df_int)
```

| Quantity | Formula | Value |
|:--|:--|--:|
| $RSS_{\text{full}}$ (cell means) | | `r round(rss_int, 2)` |
| $df_{\text{full}}$ | $n - IJ = `r n` - `r I_val` \times `r J_val`$ | `r df_int` |
| $RSS_{\text{reduced}}$ (additive) | | `r round(rss_full, 2)` |
| $df_{\text{reduced}}$ | $n - I - J + 1 = `r n` - `r I_val` - `r J_val` + 1$ | `r df_full` |

$$ESS = RSS_{\text{reduced}} - RSS_{\text{full}} = `r round(rss_full, 2)` - `r round(rss_int, 2)` = `r round(ess_int, 2)`$$

$$df_{\text{extra}} = df_{\text{reduced}} - df_{\text{full}} = `r df_full` - `r df_int` = `r df_extra_int`$$

$$F = \frac{ESS / \, df_{\text{extra}}}{RSS_{\text{full}} / \, df_{\text{full}}} = \frac{`r round(ess_int, 2)` / `r df_extra_int`}{`r round(rss_int, 2)` / `r df_int`} = `r round(f_stat_int, 2)`$$

Compare the observed $F$ to an $F_{`r df_extra_int`, `r df_int`}$ distribution:

```{r}
#| label: f-density-int
#| echo: false
f_seq2  <- seq(0, 5, length.out = 500)
f_dens2 <- df(f_seq2, df1 = df_extra_int, df2 = df_int)
p_val2  <- pf(f_stat_int, df1 = df_extra_int, df2 = df_int, lower.tail = FALSE)
tail_x2 <- f_seq2[f_seq2 >= f_stat_int]
tail_y2 <- df(tail_x2, df1 = df_extra_int, df2 = df_int)

ggplot(tibble(x = f_seq2, y = f_dens2), aes(x, y)) +
  geom_line() +
  geom_ribbon(data = tibble(x = tail_x2, y = tail_y2),
              aes(x = x, ymin = 0, ymax = y),
              fill = "steelblue", alpha = 0.7, inherit.aes = FALSE) +
  geom_vline(xintercept = f_stat_int, linetype = "dashed") +
  annotate("text", x = f_stat_int + 0.1, y = max(f_dens2) * 0.6,
           label = paste0("F = ", round(f_stat_int, 2), "\np = ", round(p_val2, 3)),
           hjust = 0, size = 3.5) +
  labs(x = paste0("F(", df_extra_int, ", ", df_int, ")"),
       y = "Density",
       title = paste0("p-value = shaded area to the right of F = ",
                      round(f_stat_int, 2)))
```

## F-Test in R

```{r}
#| label: interaction-test
aout_int <- aov(Score ~ Company * Treat, data = case1302)
tidy(aout_int)
```

The `Company:Treat` row is the interaction test:

- $p = 0.67$ — no evidence of an interaction.
- The additive model is appropriate; the main effects test above is valid.

---

# Coefficient Estimates

Use `lm()` with `tidy(conf.int = TRUE)` to get estimates and confidence intervals:

```{r}
#| label: coef-estimates
lm_add <- lm(Score ~ Company + Treat, data = case1302)
tidy(lm_add, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high, p.value)
```

- `(Intercept)`: estimated mean score for Company C1, Control platoon.
- `CompanyC2`, `CompanyC3`, etc.: difference in baseline score vs. Company C1.
- `TreatPygmalion`: estimated treatment effect — mean score increase in Pygmalion vs.
  Control platoons, **after controlling for Company**.
- 95% CI for the treatment effect: approximately 1.5 to 14.1 points.

---

# R Formula Syntax

| Formula | Model fitted |
|:--|:--|
| `A + B` | $\mu + \alpha_i + \beta_j$ (additive) |
| `A * B` | $\mu + \alpha_i + \beta_j + (\alpha\beta)_{ij}$ (with interaction) |
| `A:B` | $(\alpha\beta)_{ij}$ only (rarely used alone) |
| `A * B - A:B` | Same as `A + B` |

---

# Analysis Workflow

1. Make an **interaction plot** to assess whether additivity is plausible.
2. **Test for interaction**: if no evidence ($p > 0.05$), adopt the additive model.
3. **Test for main effects** using the additive model.
4. **Estimate and interpret** coefficients with confidence intervals.

---

# Exercises

**1.** Use the Pygmalion data (`case1302`) for this exercise.

a. Make an interaction plot with Company on the x-axis and Treatment as the grouping
   variable. Do the lines look roughly parallel?

b. Formally test for an interaction. State the $F$-statistic, degrees of freedom,
   $p$-value, and conclusion.

c. Based on part (b), which model is appropriate? State the estimated treatment effect
   and a 95% confidence interval.

---

**2.** Consider the following two interaction plots (not from real data). For each:

a. Would you say the data are consistent with an additive model?

b. If an interaction is present, describe what it means in words.

```{r}
#| label: exercise-plots
#| echo: false
#| fig-width: 7
library(patchwork)
p1_df <- tibble(
  x      = rep(c("A", "B", "C"), 2),
  mean_y = c(5, 8, 11, 9, 12, 15),
  group  = rep(c("Group 1", "Group 2"), each = 3)
)
p2_df <- tibble(
  x      = rep(c("A", "B", "C"), 2),
  mean_y = c(5, 8, 11, 9, 7, 5),
  group  = rep(c("Group 1", "Group 2"), each = 3)
)
p1 <- ggplot(p1_df, aes(x, mean_y, color = group, group = group)) +
  geom_point() + geom_line() +
  labs(title = "Plot 1", y = "Mean response", x = NULL, color = NULL)
p2 <- ggplot(p2_df, aes(x, mean_y, color = group, group = group)) +
  geom_point() + geom_line() +
  labs(title = "Plot 2", y = "Mean response", x = NULL, color = NULL)
p1 + p2
```

---

**3.** The data file `case1301.csv` contains results from a randomized experiment on
plant cover. `Treat` is the treatment and `Block` is the experimental block (a grouping
variable included to reduce residual variance).

```{r}
#| label: ex3-load
#| message: false
#| eval: false
case1301 <- read_csv("https://dcgerard.github.io/stat_302/data/case1301.csv")
```

a. Make an interaction plot of `Cover` vs. `Treat`, grouped by `Block`.
b. Formally test for a `Treat:Block` interaction.
c. Test for a main effect of `Treat`. State the conclusion in context.
