---
title: "Why Use Two Categorical Explanatory Variables?"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
set.seed(20)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(3)
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
```

# Learning Objectives

- Know the four main reasons for including a second categorical variable in an ANOVA model.
- Understand how a grouping variable reduces residual variance and increases power.
- Recognize Simpson's paradox and explain when it arises.
- Chapter 13 of the textbook.

---

# Four Reasons

Why would we use two categorical explanatory variables instead of one?

## Reason 1: Interest in Both Variables

- You are genuinely interested in the effects of **both** variables.
- Example: Does both pressure **and** temperature affect espresso foam quality?
- With no interactions, every observation contributes to estimating **both** effects.
- More efficient than running two separate one-factor experiments.

## Reason 2: Detecting and Describing Interactions

- The effect of the primary variable may **depend on** the level of a second variable.
- Example: a drug's effectiveness might differ between males and females.
- Two-way ANOVA lets you test for and characterize interaction effects.
- If no interaction is found, the result generalizes across all levels of the second variable.

## Reason 3: Variance Reduction via Grouping

- Including a second variable can **reduce residual variance**, making it easier to detect
  an effect of the primary variable.
- Analogous to using a paired design instead of a two-sample test.

### Simulation

```{r}
#| label: var-reduction-setup
#| echo: false
n     <- 30
cat1  <- c(rep("A", n / 2), rep("B", n / 2))
cat2  <- factor(rep(c("Low", "High"), n / 2), levels = c("Low", "High"))
y     <- 10 + 2.5 * (cat1 == "B") + 2.6 * (cat2 == "High") + rnorm(n)
sim_df <- tibble(cat1, cat2, y)
```

When both variables are plotted, the groups separate clearly:

```{r}
#| label: var-reduction-grouped
#| echo: false
ggplot(sim_df, aes(x = cat1, y = y, color = cat2)) +
  geom_jitter(width = 0.1) +
  labs(x = "Primary variable", color = "Grouping var")
```

If `cat2` is ignored, the within-group spread looks much larger — harder to see the `cat1` effect:

```{r}
#| label: var-reduction-ungrouped
#| echo: false
ggplot(sim_df, aes(x = cat1, y = y)) +
  geom_jitter(width = 0.1) +
  labs(x = "Primary variable")
```

Residuals are smaller when `cat2` is included:

```{r}
#| label: var-reduction-residuals
#| echo: false
bind_rows(
  augment(aov(y ~ cat1 + cat2, data = sim_df)) |> mutate(model = "With cat2"),
  augment(aov(y ~ cat1,        data = sim_df)) |> mutate(model = "Without cat2")
) |>
  ggplot(aes(x = model, y = .resid)) +
  geom_boxplot() +
  labs(y = "Residuals", x = NULL)
```

- **Grouping variable**: included in the model solely to reduce variance, not because you
  are directly interested in it.
- Include a grouping variable **regardless** of whether it is statistically significant.

## Reason 4: Controlling for Confounders — Simpson's Paradox

- In observational studies, the direction or size of an effect can **reverse** after
  adjusting for another variable.
- This is **Simpson's Paradox**.
- Arises when one group is over-represented at a "better" (or "worse") level of the
  confounder.

### Simulation

```{r}
#| label: simpsons-setup
#| echo: false
set.seed(4)
n    <- 40
cat2 <- c(1, 1, 1, 1, rep(0, n / 2 - 4), 0, 0, 0, 0, rep(1, n / 2 - 4))
cat1 <- rep(c("Group B", "Group A"), each = n / 2)
y2   <- 10 + 5 * (cat1 == "Group B") + 2 * cat2 + rnorm(n)
simp_df <- tibble(cat1, cat2 = factor(cat2, labels = c("Low", "High")), y = y2)
```

Within each level of the confounder, Group High is clearly above Group Low:

```{r}
#| label: simpsons-grouped
#| echo: false
ggplot(simp_df, aes(x = cat2, y = y, color = cat1)) +
  geom_boxplot(color = "black", coef = Inf) +
  geom_jitter(width = 0.1) +
  facet_wrap(~cat1) +
  labs(x = "Primary variable", y = "Response", color = "Confounder") +
  theme(legend.position = "none")
```

But aggregated (ignoring the confounder), Group Low looks higher — the effect **reverses**:

```{r}
#| label: simpsons-aggregate
#| echo: false
ggplot(simp_df, aes(x = cat2, y = y)) +
  geom_boxplot(coef = Inf) +
  geom_jitter(width = 0.1) +
  labs(x = "Primary variable", y = "Response",
       title = "Confounder ignored — Simpson's Paradox")
```

- Group "Low" happens to be over-represented in the "B" level of the confounder — a favorable condition — which inflates its overall mean.
- **Real-world example**: United Airlines appeared to have worse on-time performance than competitors, but United has a hub at O'Hare, which has frequent weather delays. After controlling for airport, United was competitive.
- Simpson's paradox is less likely in **randomized experiments**: randomization ensures no group is systematically over- or under-represented at any level of a confounder.

---

# Exercises

**1.** A study compares crop yield under two fertilizer treatments (A and B) across 16
fields. The researchers also recorded soil quality for each field (Poor vs. Good).

a. Even if the researchers are not interested in the effect of soil quality on yield,
   why might they include soil quality in the two-way ANOVA model?

b. What would be the consequence of *not* including soil quality if it truly affects yield?

---

**2.** In the 1970s, it was observed that UC Berkeley's graduate school admitted female
applicants at a lower rate than male applicants overall. But when individual departments
were examined, female applicants had equal or higher admission rates in every
department.

a. What statistical phenomenon does this illustrate?

b. Give one explanation for how this pattern could arise.

c. What conclusion is more valid: the overall comparison or the department-by-department
   comparison? Explain.

---

**3.** A researcher compares two diets (A vs. B) on weight loss and finds that Diet A
produces greater average weight loss. However, when participants are split by age group
(Young vs. Old), Diet B produces greater weight loss within both age groups.

a. Name the statistical phenomenon this illustrates.

b. Sketch a scenario in which this outcome is possible. What would the distribution of
   age groups across diets have to look like?

c. In a randomized controlled trial where participants are randomly assigned to Diet A
   or Diet B, would this phenomenon be likely? Why or why not?
