---
title: "Chapter 13 Worksheet"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      eval       = FALSE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Quality Espresso

@masella2015new developed a new method of brewing espresso. Traditional methods use steam from hot water to create pressure that drives water through coffee grounds. In this new method, pressure differentials between the interior and exterior of the brewing chamber control the flow. This allows finer control of both pressure and temperature. The authors investigated the effects of pressure and temperature on foam index.

| Variable | Description |
|:--|:--|
| `foamIndx` | Ratio of foam to liquid volume (%) measured 30 s after extraction |
| `trt_id` | Label for each treatment group (temperature × pressure combination) |
| `tempC` | Water temperature (°C): 75, 85, or 95 |
| `prssBar` | Brewing pressure (bar): 15 or 20 |

Load the data (you may need to adjust the path):

```{r}
#| label: load-data
#| eval: true
#| message: false
espresso <- read_csv("../data/espresso.csv") |>
  mutate(tempC   = factor(tempC),
         prssBar = factor(prssBar))
glimpse(espresso)
```

---

**1.** Informally explore the effects of pressure and temperature on foam index by
creating an interaction plot with `tempC` on the x-axis and `prssBar` as the grouping
variable (color and line). Does pressure appear to increase or decrease the foam index?
Does temperature appear to increase or decrease it? Does the plot suggest a strong
interaction?

```{r}
#| label: q1-plot

```

::: {.callout-note collapse="true"}
## Solution

```{r}
#| label: q1-sol
#| eval: true
#| echo: false
ggplot(espresso, aes(x = tempC, y = foamIndx, color = prssBar, group = prssBar)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(x = "Temperature (°C)", y = "Foam index (%)", color = "Pressure (bar)")
```

With pointwise 95% confidence intervals:

```{r}
#| label: q1-sol-ci
#| eval: true
#| echo: false
espresso |>
  group_by(tempC, prssBar) |>
  summarize(mean  = mean(foamIndx),
            se    = sd(foamIndx) / sqrt(n()),
            .groups = "drop") |>
  mutate(lower = mean - 2 * se,
         upper = mean + 2 * se) |>
  ggplot(aes(x = tempC, y = mean, color = prssBar, group = prssBar)) +
  geom_line() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(x = "Temperature (°C)", y = "Mean foam index (%)", color = "Pressure (bar)")
```

Temperature appears to decrease the foam index and higher pressure appears to increase
it. There might be a weak interaction: the foam index seems to level off at higher
temperatures for higher pressures but not for lower pressures, though the confidence
intervals overlap substantially.
:::

---

**2.** Fit the cell means model (full interaction model) and evaluate whether the
assumptions of the ANOVA model seem appropriate. Make a residual plot (residuals vs.
fitted values).

```{r}
#| label: q2-fit

```

::: {.callout-note collapse="true"}
## Solution

```{r}
#| label: q2-sol
#| eval: true
#| echo: false
aout_cell <- aov(foamIndx ~ prssBar * tempC, data = espresso)
augment(aout_cell) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, lty = 2) +
  labs(x = "Fitted values", y = "Residuals")
```

The residual plot looks good: no obvious fan shape and residuals are scattered randomly
around zero. The constant variance assumption appears reasonable.
:::

---

**3.** Write out the two-way ANOVA model with interactions for this study. Use the
notation $Y_{ijk} = \ldots$ and define all terms.

::: {.callout-note collapse="true"}
## Solution

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}$$

- $Y_{ijk}$: foam index of the $k$th observation at pressure $i$ and temperature $j$
- $\mu$: baseline mean
- $\alpha_i$: main effect of pressure level $i$
- $\beta_j$: main effect of temperature level $j$
- $(\alpha\beta)_{ij}$: interaction effect of pressure $i$ and temperature $j$
- $\epsilon_{ijk}$: random error (assumed $\sim N(0, \sigma^2)$, independent)
:::

---

**4.** Write out the null and alternative hypotheses in terms of the
$(\alpha\beta)_{ij}$ parameters to test whether there is an interaction between
pressure and temperature.

::: {.callout-note collapse="true"}
## Solution

- $H_0: (\alpha\beta)_{ij} = 0$ for all $i$ and all $j$
- $H_A:$ At least one $(\alpha\beta)_{ij} \neq 0$
:::

---

**5.** Formally test whether there is an interaction between pressure and temperature.
Report the $F$-statistic, degrees of freedom, and $p$-value, and state a conclusion.

```{r}
#| label: q5-test

```

::: {.callout-note collapse="true"}
## Solution

```{r}
#| label: q5-sol
#| eval: true
#| echo: false
aout_cell <- aov(foamIndx ~ prssBar * tempC, data = espresso)
tidy(aout_cell)
```

The $p$-value for `prssBar:tempC` is large (approximately 0.48), providing no evidence
of an interaction between pressure and temperature. We proceed with the additive model.
:::

---

**6.** Fit the additive model (main effects only). Is there a significant effect of
pressure or temperature on the foam index? Report the $F$-statistics and $p$-values for
each main effect.

```{r}
#| label: q6-add

```

::: {.callout-note collapse="true"}
## Solution

```{r}
#| label: q6-sol
#| eval: true
#| echo: false
aout_add <- aov(foamIndx ~ prssBar + tempC, data = espresso)
tidy(aout_add)
```

Yes — both pressure ($p \approx 0.0035$) and temperature ($p \approx 0.0066$) have
significant effects on foam index.
:::

---

**7.** Extract the coefficient estimates and confidence intervals from the additive model
using `tidy(lm(...), conf.int = TRUE)`. Interpret each coefficient in context.

*Hint:* The default reference level for `tempC` is 75°C. To also obtain the contrast
between 85°C and 95°C, refit with 85°C as the reference level using
`mutate(tempC = relevel(tempC, ref = "85"))` and extract that model's coefficients too.

```{r}
#| label: q7-coef

```

::: {.callout-note collapse="true"}
## Solution

```{r}
#| label: q7-sol
#| eval: true
#| echo: false
lm_add <- lm(foamIndx ~ prssBar + tempC, data = espresso)
tidy(lm_add, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high)
```

```{r}
#| label: q7-sol2
#| eval: true
#| echo: false
espresso2 <- espresso |> mutate(tempC = relevel(tempC, ref = "85"))
lm_add2 <- lm(foamIndx ~ prssBar + tempC, data = espresso2)
tidy(lm_add2, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high)
```

- **Pressure:** Increasing pressure from 15 bar to 20 bar is estimated to increase the
  foam index by about 19.8% (95% CI: 9.5% to 30.2%).
- **Temperature 75→85°C:** Increasing temperature from 75°C to 85°C is estimated to
  decrease the foam index by about 16.1% (95% CI: 3.4% to 28.8% decrease).
- **Temperature 85→95°C:** A further increase from 85°C to 95°C is estimated to
  decrease the foam index by about 3.6% additional (95% CI: 9.0% increase to
  16.5% decrease) — not significant on its own.
:::

---

**8.** Now run three separate two-sample $t$-tests comparing foam index at the two
pressure levels (15 bar vs. 20 bar), one test for each temperature level (75°C, 85°C,
and 95°C). Do any of these individual tests produce a significant result? How does this
compare to the conclusion from the additive ANOVA in Question 6?

```{r}
#| label: q8-ttests

```

::: {.callout-note collapse="true"}
## Solution

```{r}
#| label: q8-sol
#| eval: true
#| echo: false
dat75 <- espresso |> filter(tempC == "75")
dat85 <- espresso |> filter(tempC == "85")
dat95 <- espresso |> filter(tempC == "95")

t.test(foamIndx ~ prssBar, data = dat75)
t.test(foamIndx ~ prssBar, data = dat85)
t.test(foamIndx ~ prssBar, data = dat95)
```

None of the individual $t$-tests is significant at the 0.05 level when run separately.
This contrasts with the additive ANOVA, which found a significant pressure effect
($p \approx 0.0035$). The ANOVA is more powerful because it pools the residual variance
estimate across all temperature groups, giving a much more precise estimate of $\sigma^2$
and therefore more power to detect the pressure effect.
:::

# References
