---
title: "F-test in Multiple Regression"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    slide_level: 2
    theme: "metropolis"
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 2.3, 
                      fig.width = 3.7, 
                      fig.align = "center")
```


```{r}
set.seed(1)
library(ggplot2)
```

## Learning Objective

- Test for including multiple variables at the same time.

- Section 10.3 in the book

## Case Study

- Kentucky Derby

- `Speed` vs `Year` and `Yeat`$^2$.

```{r, echo = TRUE}
library(Sleuth3)
data(ex0920)
head(ex0920)
```

## Year vs Speed

```{r, echo = TRUE, message=FALSE}
qplot(Year, Speed, data = ex0920) +
  geom_smooth(se = FALSE)
```

## Goal

- Get a $p$-value for the association between year and Speed.

- It is clear that a quadratic model would be better than a linear model.

- $\mu(Speed|Year) = \beta_0 + \beta_1 Year + \beta_2 Year^2$

- So to see if year is important, we need to test:

    - $H_0: \beta_1 = \beta_2 = 0$
    
    - $H_A: \text{ either } \beta_1 \neq 0 \text{ or } \beta_2 \neq 0$
    
## Full and Reduced Models:

- Full Model: $\mu(Speed|Year) = \beta_0 + \beta_1 Year + \beta_2 Year^2$

- Reduced Model: $\mu(Speed|Year) = \beta_0$

- Use $F$-test strategy to run this hypothesis test.

    1. Fit both full and reduced models.
    2. Calculate sum of squared residuals under both models and the corresponding degrees of freedom.
    3. Calculate the $F$-statistic.
    4. Compare to theoretical $F$-distribution under $H_0$
    

## Fit Under Full

```{r}
ex0920$Year2 <- ex0920$Year ^ 2
lmfull <- lm(Speed ~ Year + Year2, data = ex0920)
lmreduced <- lm(Speed ~ 1, data = ex0920)

ex0920$fit_full <- fitted(lmfull)
ex0920$fit_red  <- fitted(lmreduced)

qplot(Year, Speed, data = ex0920) +
  geom_line(mapping = aes(y = fit_full), lwd = 1, color = "blue")
```

## Residuals under Full

```{r}
qplot(Year, Speed, data = ex0920) +
  geom_line(mapping = aes(y = fit_full), lwd = 1, color = "blue") +
  geom_segment(aes(x = Year, xend = Year, y = fit_full, yend = Speed), alpha = 1/2)
```

## Fit under Reduced

```{r}
qplot(Year, Speed, data = ex0920) +
  geom_line(mapping = aes(y = fit_red), lwd = 1, color = "blue")
```

## Residuals under Reduced

```{r}
qplot(Year, Speed, data = ex0920) +
  geom_line(mapping = aes(y = fit_red), lwd = 1, color = "blue") +
  geom_segment(aes(x = Year, xend = Year, y = fit_red, yend = Speed), alpha = 1/2)
```

    
## In R

- First, fit both models

```{r, echo = TRUE}
ex0920$Year2 <- ex0920$Year ^ 2
lmfull <- lm(Speed ~ Year + Year2, data = ex0920)
lmreduced <- lm(Speed ~ 1, data = ex0920)
```

- Then use `anova()` with the reduced model as the first argument.

\footnotesize
```{r, echo = TRUE}
anova(lmreduced, lmfull)
```
\normalsize


## What is that Table?

```{r}
anova(lmreduced, lmfull)
```

\begin{center}
\begin{tabular}{llllll}
  Res.Df  & RSS & Df &  Sum of Sq   & F &  Pr(>F)\\
  $df_{reduced}$ & $RSS_{reduced}$ & & & & \\
  $df_{full}$ & $RSS_{full}$ & $df_{extra}$ & $ESS$ & $F$-stat & $p$-value\\
\end{tabular}
\end{center}


## $F$-test

- We can use the $F$-test for any two **nested** models.

- **Nested**: The reduced model is a special case of the full model by setting constraints on some of the parameters of the full.

## Another Example

- $\mu(Speed|Year,Starters) = \beta_0 + \beta_1Year + \beta_2Year^2 + \beta_3Starters + \beta_4Starters^2$

- $H_0: \beta_3 = \beta_4 = 0$
    
- $H_A: \text{ either } \beta_3 \neq 0 \text{ or } \beta_4 \neq 0$

- Full Model: $\mu(Speed|Year,Starters) = \beta_0 + \beta_1Year + \beta_2Year^2 + \beta_3Starters + \beta_4Starters^2$

- Reduced Model: $\mu(Speed|Year,Starters) = \beta_0 + \beta_1Year + \beta_2Year^2$

## Another Example

```{r, echo = TRUE}
ex0920$Starters2 <- ex0920$Starters ^ 2
lmfull <- lm(Speed ~ Year + Year2 + Starters + 
               Starters2, data = ex0920)
lmreduced <- lm(Speed ~ Year + Year2, data = ex0920)
anova(lmreduced, lmfull)
```

## Example of a **non**-nested model

- Model 1: $\mu(Speed|Year,Starters) = \beta_0 + \beta_1Year + \beta_2Year^2$

- Model 2: $\mu(Speed|Year,Starters) = \beta_0 + \beta_1Starters + \beta_2Starters^2$

- **Cannot** use an $F$-test to compare these two models.

- Why? Mathematical theory only gaurantees the $F$-distribution when the models are nested.

- When models are not nested, use adjusted $R^2$, $C_p$, AIC, or BIC methods from section 12.4 (more on this later).
