---
title: "F-tests for Nested Models"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
set.seed(15)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(1)
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
```

# Learning Objectives

- Test whether a **group** of explanatory variables jointly improves a model.
- Distinguish *nested* from *non-nested* model pairs.
- Interpret the $F$-statistic and its degrees of freedom.
- Use `anova()` in R to run the $F$-test.
- Section 10.3 in the textbook.

---

# Motivation

## The Problem with Individual $t$-tests

- In multiple regression we can test each coefficient with a $t$-test.
- But some variables enter the model through **multiple terms** (e.g., $X$ and $X^2$).
- Testing $\beta_1 = 0$ and $\beta_2 = 0$ separately does not answer "does $X$ matter at all?"
- We need to test $H_0: \beta_1 = 0$ **and** $\beta_2 = 0$ **simultaneously** — a joint test.

## Case Study: Kentucky Derby

- The Kentucky Derby is a famous horse race held annually since 1875.
- We have the winning speed (mph) and the year of each race.
- Does year explain speed? (Speed has increased over time as breeding and track conditions improved.)

```{r}
#| label: derby-load
#| message: false
ex0920 <- read_csv("https://dcgerard.github.io/stat_302/data/ex0920.csv")
glimpse(ex0920)
```

```{r}
#| label: derby-scatter
#| message: false
ggplot(ex0920, aes(x = Year, y = Speed)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(y = "Winning Speed (mph)")
```

- The trend is clearly **nonlinear**: speed rose sharply in the early 20th century then leveled off.
- A quadratic model is appropriate:
$$\mu(\text{Speed} \mid \text{Year}) = \beta_0 + \beta_1\,\text{Year} + \beta_2\,\text{Year}^2$$
- **Question**: Is year *at all* associated with winning speed?
- $H_0: \beta_1 = 0$ and $\beta_2 = 0$
- $H_A: \beta_1 \neq 0$ or $\beta_2 \neq 0$

---

# Nested Models

## Definition

- A model is **nested** inside another if it can be obtained by setting some coefficients
  in the larger model to zero.
- The larger model is the **full model**.
- The smaller model is the **reduced model**.

## Kentucky Derby Example

| Model | Equation |
|:--|:--|
| **Full** | $\mu(\text{Speed} \mid \text{Year}) = \beta_0 + \beta_1\,\text{Year} + \beta_2\,\text{Year}^2$ |
| **Reduced** (under $H_0$) | $\mu(\text{Speed}) = \beta_0$ |

- The reduced model is the full model with $\beta_1 = \beta_2 = 0$.

## Non-Nested Models (Counter-example)

- Model A: $\mu(\text{Speed} \mid \text{Year}) = \beta_0 + \beta_1\,\text{Year} + \beta_2\,\text{Year}^2$
- Model B: $\mu(\text{Speed} \mid \text{Starters}) = \beta_0 + \beta_1\,\text{Starters} + \beta_2\,\text{Starters}^2$
- Neither is a special case of the other — **cannot** use an $F$-test.
- For non-nested models, use AIC or BIC (see the model selection lecture).

## Intuition: Residuals Tell the Story

- If the full model fits much better (smaller residuals), we have evidence against $H_0$.
- We compare the **residual sum of squares** (RSS) under each model.

```{r}
#| label: derby-models
ex0920 <- ex0920 |> mutate(Year2 = Year^2)
lm_full    <- lm(Speed ~ Year + Year2, data = ex0920)
lm_reduced <- lm(Speed ~ 1,            data = ex0920)

ex0920 <- ex0920 |>
  mutate(fit_full = fitted(lm_full),
         fit_red  = fitted(lm_reduced))
```

```{r}
#| label: resid-full
#| echo: false
ggplot(ex0920, aes(x = Year, y = Speed)) +
  geom_segment(aes(xend = Year, yend = fit_full), alpha = 0.35, color = "steelblue") +
  geom_point(size = 1.5) +
  geom_line(aes(y = fit_full), color = "steelblue", linewidth = 1) +
  labs(title = "Full model residuals (small)", y = "Speed (mph)")
```

```{r}
#| label: resid-reduced
#| echo: false
ggplot(ex0920, aes(x = Year, y = Speed)) +
  geom_segment(aes(xend = Year, yend = fit_red), alpha = 0.35, color = "tomato") +
  geom_point(size = 1.5) +
  geom_line(aes(y = fit_red), color = "tomato", linewidth = 1) +
  labs(title = "Reduced model residuals (large)", y = "Speed (mph)")
```

- The full model fits much better — the residuals are far smaller.
- This large difference in fit is strong evidence against $H_0$.

---

# The $F$-Statistic

## Key Quantities

Let $p_F$ = number of parameters in the full model, $p_R$ = number in the reduced, $n$ = sample size.

| Symbol | Meaning |
|:--|:--|
| $RSS_F$ | Sum of squared residuals under the full model |
| $RSS_R$ | Sum of squared residuals under the reduced model |
| $ESS$ | **Extra sum of squares** $= RSS_R - RSS_F$ (how much better the full fits) |
| $df_{\text{extra}}$ | $p_F - p_R$ (extra parameters in the full model) |
| $df_F$ | $n - p_F$ (residual degrees of freedom for the full model) |

## Formula and Distribution

$$F = \frac{ESS / \, df_{\text{extra}}}{RSS_F / \, df_F}$$

- **Numerator**: average extra variability explained per additional parameter.
- **Denominator**: average residual variability per degree of freedom (an estimate of $\sigma^2$).
- Under $H_0$: $F$ follows an $F$-distribution with $(df_{\text{extra}},\, df_F)$ degrees of freedom.
- A **large** $F$ means the full model explains substantially more — evidence against $H_0$.

---

# Running the $F$-Test in R

Use `anova()` with the **reduced** model first, the **full** model second:

```{r}
#| label: anova-derby
anova(lm_reduced, lm_full) |>
  tidy()
```

## Reading the Output

| Column | Meaning |
|:--|:--|
| `Res.Df` | Residual degrees of freedom for each model |
| `RSS` | Residual sum of squares for each model |
| `Df` | $df_{\text{extra}} = p_F - p_R$ |
| `Sum of Sq` | $ESS = RSS_R - RSS_F$ |
| `F` | $F$-statistic |
| `Pr(>F)` | $p$-value |

- **Conclusion**: We have overwhelming evidence that year is associated with winning speed
  ($F = 104.6$ on 2 and 119 df, $p < 0.001$).

---

# Another Example: Do Starters Matter?

## Setup

- `Starters`: number of horses that started the race.
- After accounting for a quadratic trend in year, does the number of starters add anything?

$$H_0: \beta_3 = \beta_4 = 0 \text{ in:}$$
$$\mu(\text{Speed} \mid \text{Year}, \text{Starters}) = \beta_0 + \beta_1\,\text{Year} + \beta_2\,\text{Year}^2
  + \beta_3\,\text{Starters} + \beta_4\,\text{Starters}^2$$

- **Full**: Year + Year² + Starters + Starters²
- **Reduced**: Year + Year² (same as the full model above)

```{r}
#| label: anova-starters
ex0920 <- ex0920 |> mutate(Starters2 = Starters^2)
lm_full2    <- lm(Speed ~ Year + Year2 + Starters + Starters2, data = ex0920)
lm_reduced2 <- lm(Speed ~ Year + Year2,                        data = ex0920)
anova(lm_reduced2, lm_full2) |>
  tidy()
```

- **Conclusion**: We do have some evidence that the number of starters is associated with speed after adjusting for year ($F = 3.92$ on 2 and 111 df, $p = 0.023$).

---

# Exercises

**1.** The brain size dataset (`case0902`) has three potential explanatory variables for
`Brain` weight (grams): `Body` (kg), `Gestation` (days), and `Litter` size.

Load the data:
```r
case0902 <- read_csv("https://dcgerard.github.io/stat_302/data/case0902.csv")
```

Then consider:
- **Reduced model**: `Brain ~ Body`
- **Full model**: `Brain ~ Body + Gestation + Litter`

a. Write out $H_0$ and $H_A$ in terms of the $\beta$ coefficients. In plain English,
   what does the null hypothesis claim?

b. Run `anova(lm_reduced, lm_full)` in R. Report the $F$-statistic, the two degrees
   of freedom, and the $p$-value.

c. State a conclusion in context.

---

**2.** Which of the following pairs of models are nested? For nested pairs, identify
which is the reduced and which is the full model.

a. Model A: $\mu(Y \mid X_1, X_2) = \beta_0 + \beta_1 X_1 + \beta_2 X_2$
   Model B: $\mu(Y \mid X_1) = \beta_0 + \beta_1 X_1$

b. Model A: $\mu(Y \mid X_1, X_2) = \beta_0 + \beta_1 X_1 + \beta_2 X_2$
   Model B: $\mu(Y \mid X_1, X_3) = \beta_0 + \beta_1 X_1 + \beta_3 X_3$

c. Model A: $\mu(Y \mid X_1) = \beta_0 + \beta_1 X_1 + \beta_2 X_1^2$
   Model B: $\mu(Y \mid X_1) = \beta_0 + \beta_1 X_1$

d. Model A: $\mu(Y \mid X_1, X_2, X_3) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3$
   Model B: $\mu(Y \mid X_1, X_2) = \beta_0 + \beta_1 X_1 + \beta_2 X_2$

---

**3.** Using the Kentucky Derby data, test whether the **quadratic** term `Year²` adds
significantly on top of the **linear** term `Year` alone.

- Reduced model: `Speed ~ Year`
- Full model: `Speed ~ Year + Year2`

a. Write out $H_0$ and $H_A$.

b. Run the test and report the results.

c. Compare the $p$-value to the one from the bigger test (intercept-only reduced model)
   earlier in this lecture. Why are they different?
