---
title: "Non-nested Comparisons"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    slide_level: 2
    theme: "metropolis"
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 2.3, 
                      fig.width = 3.7, 
                      fig.align = "center")
```


```{r}
set.seed(1)
library(ggplot2)
```

## Learning Objective

- Sections 10.4.1 and 12.4

- Choosing Between Non-nested Models

# Case Study and EDA


## Case Study: Sex Descrimination

- Same study as in Case Study 0102

- Looked at beginning salary at a bank with respect to sex.

- Want to control for many different variables.

## Case Study: Sex Descrimination

```{r, echo = TRUE}
library(Sleuth3)
data(case1202)
head(case1202)
```


## EDA

```{r, echo = TRUE, message=FALSE}
library(GGally)
ggpairs(case1202, columns = c(1, 2, 4, 5, 6, 7), 
        aes(size = I(0.1)))
```

## EDA

```{r, echo = TRUE, message=FALSE}
case1202$logBsal <- log(case1202$Bsal)
ggpairs(case1202, columns = c(2, 4, 5, 6, 7, 8), 
        aes(size = I(0.1)))
```

## EDA Summary

- Loging `Bsal` seems to help a lot.

- Age and Experience might need a quadratic transformation.

# Step-wise Procedures (Section 12.3)

## Step-wise Regression

- Start with a complicated model.

- Look at $p$-values (when testing that a coefficient is 0)

- Drop the one with the largest $p$-value.

- Continue until all $p$-values are less than some threshold (usually 0.05).

- Note, you cannot interpret $p$-values the way we define them anymore if you do this.

## Step-wise Regression, the manual way

```{r, echo = TRUE}
case1202$Age2   <- case1202$Age ^ 2
case1202$Exper2 <- case1202$Exper ^ 2
lm1 <- lm(logBsal ~ Senior + Age + Age2 + 
            Senior + Educ + Exper + Exper2,
          data = case1202)
```

## Step-wise Regression, the manual way

```{r, echo = TRUE}
coef(summary(lm1))
```

- Drop `Age2` ($p$-value of 0.97)

## Step-wise Regression, the manual way

```{r}
lm2 <- lm(logBsal ~ Senior + Age + 
            Senior + Educ + Exper + Exper2,
          data = case1202)
coef(summary(lm2))
```

## Step-wise Regression

- Can also start at the simplest model,

    - add the variable that has the smallest $p$-value
    
    - continue until no new variables would have a $p$-value less than 0.05
    
- Can also both add and drop variables based on $p$-values.

## Step-wise Regression in R

- Use the `step()` function to do this automatically
    - It actually uses AIC (not $p$-values) to choose between models, but the idea is similar. See later for AIC.

\footnotesize
```{r, echo = TRUE}
lm1 <- lm(logBsal ~ Senior + Age + Age2 + 
            Senior + Educ + Exper + Exper2,
          data = case1202)
stepout <- step(object = lm1, trace = FALSE)
stepout
```
\normalsize

## Step-wise Regression in R

- The output of `step()` is also an `lm` object, so you can get coefficients, $p$-values, confidence intervals, fits, predictions, residuals, etc directly from it.

```{r, echo = TRUE}
confint(stepout)
```

# Comparing Non-nested Models (Section 12.4)

## Motivation

- What if we want to decide between the following two models

- $\mu(logBsal|...) = Senior + Educ + Exper + Exper^2$

- $\mu(logBsal|...) = Senior + Educ + Age + Age^2$

- These models are non-nested, so we cannot apply $F$-test techniques to them.

## BIC and AIC

- BIC (Bayesian Information Criterion) and AIC (Akaike Information Criterion) return the log of the sum of square residuals **plus** a penalty due to the number of parameters in the model.

- Best model has the smallest BIC or AIC.

- BIC: $n\log(SSR/n) + \log(n)(p + 1)$

- AIC: $n\log(SSR/n) + 2(p+1)$

- BIC penalizes more when the sample size is larger.

- BIC is better for model selection (get interpretable model), AIC is better for prediction (goal is prediction).

## BIC and AIC in R
- Fit both models, then use the `AIC()` and `BIC()` functions.
\tiny
```{r, echo = TRUE}
lm_mod1 <- lm(logBsal ~ Senior + Educ + Exper + Exper2, 
              data = case1202)
lm_mod2 <- lm(logBsal ~ Senior + Educ + Age + Age2, 
              data = case1202)
BIC(lm_mod1)
BIC(lm_mod2)
AIC(lm_mod1)
AIC(lm_mod2)
```
\normalsize

## Mallow's $C_p$ statistic

- $Bias(\hat{Y}_i) = \mu(\hat{Y}_i) - \mu(Y_{i})$

- $MSE(\hat{Y}_i) = Bias(\hat{Y}_i)^2 + Var(\hat{Y}_i)$

- $TMSE = \sum_{i = 1}^n MSE(\hat{Y}_i)$

- We don't know the $TMSE$, but Mallow's $C_p$ **estimates** it.

## $C_p$ plot

- You obtain Mallow's $C_p$ for **every** possible model.

- Only feasible if you have less than $p = 10$ or so explanatory variables ($2^p$ models are possible).

- Plot $C_p$ on the $y$-axis and the number of parameters on the $x$-axis.

- Models below the $y = x$ line are candidate models 

    - Models without bias should have a $C_p$ of about $p$
    
    - So if $C_p$ is below $p$, the model probably does not have any bias issues.
    
## $C_p$ in R

- We will use the `leaps()` function in the `leaps` library.

```{r, echo = TRUE}
library(leaps)
lm1 <- lm(logBsal ~ Senior + Age + Age2 + 
            Senior + Educ + Exper + Exper2,
          data = case1202)
X <- model.matrix(lm1)
leapsout <- leaps(x =  X, 
                  y = case1202$logBsal,
                  int = FALSE)
```

## $C_p$ in R

```{r, echo = TRUE}
qplot(leapsout$size, leapsout$Cp, 
      xlab = "Number of Parameters",
      ylab = "Cp") +
  geom_abline(slope = 1, intercept = 0)
```

## $C_p$ in R

```{r, echo = TRUE}
goodmodel <- leapsout$Cp < 1000
qplot(leapsout$size[goodmodel], leapsout$Cp[goodmodel], 
      xlab = "Number of Parameters",
      ylab = "Cp") +
  geom_abline(slope = 1, intercept = 0) +
  ylim(0, 60)
```




# Back to Case Study

## Back to Case Study

- We chose a model with 
$$\mu(logBsal|...) = Senior + Age + Senior + Educ + Exper + Exper^2$$

- Now let's answer the question if `Sex` is still associated with base salary after adjusting for these variables.

## Results

```{r, echo=TRUE}
lmfinal <- lm(logBsal ~ Sex + Senior + Age + 
                Senior + Educ + Exper + Exper2,
              data = case1202)
coef(summary(lmfinal))
```

## Results

```{r, echo=TRUE}
cbind(coef(lmfinal), confint(lmfinal))
```

## Results

```{r, echo=TRUE}
exp(coef(lmfinal)[2])
exp(confint(lmfinal)[2, ])
```




