---
title: "Model Selection Worksheet"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      eval       = FALSE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Pollution and Mortality

Does air pollution kill people? Data from an early study were collected for 60 Standard
Metropolitan Statistical Areas (cities) in the United States, covering the years 1959–1961.
The goal is to determine whether the pollution variables are associated with mortality **after**
accounting for climate and socioeconomic differences between cities.

Variables:

| Variable | Description |
|:--|:--|
| `CITY` | City name |
| `Mortality` | Total age-adjusted mortality (deaths per 100,000) |
| `Precip` | Mean annual precipitation (inches) |
| `Humidity` | Percent relative humidity (annual average at 1 pm) |
| `JanTemp` | Mean January temperature (°F) |
| `JulyTemp` | Mean July temperature (°F) |
| `Over65` | Percent of population aged 65 or older |
| `House` | Population per household |
| `Educ` | Median school years completed (age 25+) |
| `Sound` | Percent of housing that is sound with all facilities |
| `Density` | Population density (persons per square mile) |
| `NonWhite` | Percent of population that is non-white |
| `WhiteCol` | Percent employed in white-collar occupations |
| `Poor` | Percent of households with income under \$3,000 (1960) |
| `HC` | Relative pollution potential of hydrocarbons |
| `NOX` | Relative pollution potential of nitrogen oxides |
| `SO2` | Relative pollution potential of sulfur dioxide |

Load the data with:

```{r}
#| label: load-data
#| eval: true
#| message: false
library(tidyverse)
library(broom)
ex1217 <- read_csv("https://dcgerard.github.io/stat_302/data/ex1217.csv")
glimpse(ex1217)
```

---

**1.** What are the observational units? What is the response variable? Which variables are
the primary explanatory variables of interest (the pollution variables)? Which are
considered control variables?

---

**2.** Make scatterplots of each pollution variable (`HC`, `NOX`, `SO2`) against
`Mortality`. Based on these plots, do any transformations seem appropriate?

```{r}

```

---

**3.** Create log-transformed versions of the three pollution variables. Then make new
scatterplots using the log-transformed versions. Does the log transformation help?

```{r}

```

---

**4.** Fit a rich model that includes all non-pollution control variables (no interactions,
no transformations of the controls needed):

```r
lm_rich <- lm(Mortality ~ Precip + Humidity + JanTemp + JulyTemp +
                Over65 + House + Educ + Sound + Density +
                NonWhite + WhiteCol + Poor,
              data = ex1217)
```

Create a residual plot and a QQ-plot of the residuals. Do the model assumptions look
reasonable?

```{r}

```

---

**5.** Compute leverage values, studentized residuals, and Cook's distances for the rich
model from Question 4. Produce plots of each against observation number. Are there any
cities you are particularly concerned about?

```{r}

```

---

**6.** Use `step()` on the rich model from Question 4 to select a subset of control
variables. Do **not** include the pollution variables in this step — we will add them
separately. What variables were retained?

```{r}

```

---

**7.** Write out the full regression model you will use to test for a pollution effect.
The model should include the variables selected in Question 6 plus the three
log-transformed pollution variables.

---

**8.** Fit the model from Question 7. Use `anova()` with the stepwise-selected model
(no pollution variables) as the reduced model and the model with pollution as the full
model.

a. State $H_0$ and $H_A$ for this $F$-test.

b. Report the $F$-statistic, degrees of freedom, and $p$-value.

c. State a conclusion in the context of the study.

```{r}

```

---

**9.** Use `tidy(lm_final, conf.int = TRUE)` to examine the individual pollution
coefficients. Are any of the pollution variables individually significant? Do the signs
of the coefficients make intuitive sense? If not, what might explain a counterintuitive
sign?

```{r}

```

---

**10.** Rerun the stepwise selection and $F$-test from Questions 6 and 8, but this time
**exclude the high-leverage cities** you identified in Question 5. Does the conclusion
about the pollution-mortality association change? What does this tell you about the
robustness of the result?

```{r}

```
