---
title: "R Graphics with `{ggplot2}`"
author: "David Gerard"
date: today
format: html
bibliography: "../bib.bib"
---

```{r setup}
#| include: false
set.seed(1)
knitr::opts_chunk$set(echo = TRUE,
            fig.width = 4,
            fig.height = 3,
            fig.align = "center")
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r}
#| echo: false
set.seed(1)
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
```

# Learning objectives

- Basic plotting in R using the `{ggplot2}` package.

# Introduction

- `{ggplot2}` is very powerful, so I am just going to show you the most important and basic plots that are necessary for data analysis.

-   Before using the plotting functions from `{ggplot2}` in a new R session, always first load the `{tidyverse}` library that has `{ggplot2}` as part of it.

    ```{r}
    #| message: false
    library(tidyverse)
    ```

-   For this lecture, let's consider the data [ex1027.csv](https://dcgerard.github.io/stat_302/data.html#chapter-10), from @pimm1988risk. They were interested in if nesting pair size was related to rate of extinction. They measured the average nesting pair size and time to extinction across 16 islands for 62 bird species. They also measued the size of the species (large/small) and the migratory status of the species (migratory/resident).
  
    ```{r}
    #| message: false
    birds <- read_csv("https://dcgerard.github.io/stat_302/data/ex1027.csv")
    glimpse(birds)
    ```

# `ggplot()`

- The first function you use in making a plot is always `ggplot()`.

- It takes two main arguments:
  - `data`: The data frame that holds the variables you want to plot.
  - `mapping`: The "aesthetic map"

- An "aesthetic map" says what variables go on the $x$-axis, what variables go on the $y$-axis, what variables are represented by color, or size, or point shape, etc...

- You place all aesthetic maps inside an `aes()` function.

-   E.g. here, we are mapping `Time` to be on the $x$-axis, and different values of `Size` should be different colors.

    ```{r}
    #| eval: false
    #| fig-alt: "Basic ggplot graphic."
    ggplot(data = birds, mapping = aes(x = Time, color = Size))
    ```

- This function just sets the data and the aesthetic mapping, but it won't produce any useful plot by itself.

- You add additional functions to the plot to state the type of plot you want.

# One Variable

## Continuous

-   Histogram:
    - Variable should be on the $x$-axis.
    - Use the `geom_histogram()` function.

    ```{r}
    #| message: false
    #| warning: false
    #| fig-alt: "Histogram of Time."
    ggplot(data = birds, mapping = aes(x = Time)) +
      geom_histogram()
    ```

-   Make the bin lines black and the fill white, and change the number of bins.

    ```{r}
    #| warning: false
    #| fig-alt: "Histogram of IQF with white fill and block color."
    ggplot(data = birds, mapping = aes(x = Time)) +
      geom_histogram(bins = 20, color = "black", fill = "white")
    ```

::: {.panel-tabset}
## Exercise
Load in the "ex0223.csv" data from the [course website](https://dcgerard.github.io/stat_302/data.html#chapter-2). Call the resulting data frame `highway`. 

Make a histogram of the 1996 fatalities.

## Solution

```{r}
#| eval: true
#| message: false
highway <- read_csv("https://dcgerard.github.io/stat_302/data/ex0223.csv")
ggplot(highway, aes(x = Fatalities1996)) +
  geom_histogram(bins = 20)
```
:::

## Discrete

-   Barplot:
    - Put the variable on the $x$-axis.
    - Use `geom_bar()`.
  
    ```{r}
    #| fig-alt: "Barplot of hyperactivity rating."
    ggplot(data = birds, mapping = aes(x = Size)) +
      geom_bar()
    ```

::: {.panel-tabset}
## Exercise 
What variable(s) from the `highway` data is(are) appropriately plotted using a bar plot? Plot them.

## Solution

Just the `SpeedLimit` variable
```{r}
ggplot(highway, aes(x = SpeedLimit)) +
  geom_bar()
```
:::

# Two Variables

## Continuous X, Continuous Y

-   Scatterplot:
    - Say what variables should be on the $x$- and $y$-axes.
    - Use `geom_point()`.
  
    ```{r}
    ggplot(data = birds, mapping = aes(x = Pairs, y = Time)) +
      geom_point()
    ```

-   Add a Loess Smoother by adding `geom_smooth()`. (Don't take the loess smoother too seriously)

    ```{r}
    ggplot(data = birds, mapping = aes(x = Pairs, y = Time)) +
      geom_point() +
      geom_smooth(formula = y ~ x, method = "loess", se = FALSE)
    ```

::: {.panel-tabset}
## Exercise
Using the `highway` data, make a scatterplot exploring the association between the fatility numbers of the two years.

## Solution
```{r}
ggplot(data = highway, mapping = aes(x = Fatalities1995, y = Fatalities1996)) +
  geom_point()
```
:::

## Discrete X, Continuous Y

-   Boxplot
    - Place one variable on $x$-axis and other on $y$-axis.
    - Typically, but not always, continuous goes on $y$-axis.
    - Use `geom_boxplot()`.
  
    ```{r}
    ggplot(data = birds, mapping = aes(x = Size, y = Time)) +
      geom_boxplot()
    ```

::: {.panel-tabset}
## Exercise
Using the `highway` data, explore the association between increasing the speedlimit and the percent change.

## Solution
```{r}
ggplot(data = highway, mapping = aes(x = SpeedLimit, y = PctChange)) +
  geom_boxplot()
```
:::

# Color Coding and Legends

-   Color code a scatterplot by a categorical variable and add a legend.
    - Just add a color mapping.
  
    ```{r}
    ggplot(data = birds, mapping = aes(x = Pairs, y = Time, color = Size)) +
      geom_point()
    ```

::: {.panel-tabset}
## Exercise 

Using the `highway` data, create a scatterplot between the fatality rates of the two years and color code by speed limit policy.

## Solution
```{r}
ggplot(data = highway, mapping = aes(x = Fatalities1995, y = Fatalities1996, color = SpeedLimit)) +
  geom_point()
```
:::

# Changing a legend title

-   Add a `scale_*()` call to change the name:

    ```{r}
    ggplot(data = birds, mapping = aes(x = Pairs, y = Time, color = Size)) +
      geom_point() +
      scale_color_discrete(name = "Bird Size")
    ```

# Change Theme

-   Add a `theme_*()` function to change the theme:
    - I always use `theme_bw()`.

    ```{r}
    ggplot(data = birds, mapping = aes(x = Pairs, y = Time, color = Size)) +
      geom_point() +
      scale_color_discrete(name = "Bird Size") +
      theme_classic()
    ```
