---
title: "Linear Combinations of Group Means"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(car)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Define a linear combination and a contrast of group means.
- Estimate a linear combination and compute its standard error.
- Conduct a hypothesis test and form a confidence interval for a linear
  combination.
- Use `linearHypothesis()` from the `car` package to test linear combinations
  in R.

---

# The Spock Conspiracy Trial (Recap)

```{r}
#| label: load-data
#| message: false
case0502 <- read_csv("https://dcgerard.github.io/stat_302/data/case0502.csv") |>
  mutate(Judge = factor(Judge))
```

The model is $Y_{ij} = \mu_i + \varepsilon_{ij}$, where $Y_{ij}$ is the
percent women on venire $j$ of judge $i$, $\mu_i$ is judge $i$'s true mean,
and $\varepsilon_{ij}$ is individual noise with variance $\sigma^2$. There
are $I = 7$ judges and $n = 46$ venires.

The omnibus $F$-test from Chapter 5 gave very strong evidence that the seven
judges do not all have the same mean ($p < 0.001$). Now we want to ask more
specific questions.

---

# Linear Combinations

## Motivation

We want to compare Spock's judge ($\mu_1$) to the average of the other six
judges ($\mu_2, \ldots, \mu_7$), **without** having to assume those six means
are equal. That is:

$$H_0: \mu_1 = \tfrac{1}{6}(\mu_2 + \mu_3 + \cdots + \mu_7),
\qquad
H_A: \mu_1 \neq \tfrac{1}{6}(\mu_2 + \mu_3 + \cdots + \mu_7)$$

Rewriting the null:

$$\mu_1 - \tfrac{1}{6}\mu_2 - \tfrac{1}{6}\mu_3 - \cdots - \tfrac{1}{6}\mu_7 = 0$$

The left-hand side is a **linear combination** of the group means.

## Definition

A **linear combination** of $\mu_1, \mu_2, \ldots, \mu_I$ with coefficients
$c_1, c_2, \ldots, c_I$ is:

$$\gamma = c_1\mu_1 + c_2\mu_2 + \cdots + c_I\mu_I$$

If the coefficients sum to zero ($c_1 + c_2 + \cdots + c_I = 0$), the linear
combination is called a **contrast**.

For the Spock comparison: $c_1 = 1$ and $c_2 = c_3 = \cdots = c_7 = -\tfrac{1}{6}$.
Check: $1 + 6 \times (-\tfrac{1}{6}) = 0$. So this is a contrast.

## Hypotheses

For any linear combination $\gamma$, the hypotheses are:

$$H_0: \gamma = 0, \qquad H_A: \gamma \neq 0$$

## Estimation

Estimate each $\mu_i$ with the group sample mean $\bar{Y}_{i\bullet}$:

$$\hat{\gamma} = c_1 \bar{Y}_{1\bullet} + c_2 \bar{Y}_{2\bullet} + \cdots + c_I \bar{Y}_{I\bullet}$$

## Standard Error

Because the group means are independent and
$\mathrm{Var}(\bar{Y}_{i\bullet}) = \sigma^2 / n_i$:

$$\mathrm{Var}(\hat{\gamma}) = \sigma^2 \left(\frac{c_1^2}{n_1} + \frac{c_2^2}{n_2} + \cdots + \frac{c_I^2}{n_I}\right)$$

We estimate $\sigma$ with the pooled standard deviation $s_p$ from the full
model, giving:

$$SE(\hat{\gamma}) = s_p \sqrt{\frac{c_1^2}{n_1} + \frac{c_2^2}{n_2} + \cdots + \frac{c_I^2}{n_I}}$$

## Hypothesis Test

Under $H_0$, the standardized estimate follows a $t$ distribution:

$$t^* = \frac{\hat{\gamma}}{SE(\hat{\gamma})} \;\sim\; t_\nu, \quad \nu = n - I$$

The two-sided $p$-value is $2 \cdot P(t_\nu \leq -|t^*|)$.

## Confidence Interval

A 95% confidence interval for $\gamma$:

$$\hat{\gamma} \;\pm\; t_\nu(0.975) \cdot SE(\hat{\gamma})$$

---

# Implementation in R

## The Default Parameterization

Fitting a one-way ANOVA with `aov()` uses R's default treatment-contrast
parameterization: the model is $Y_{ij} = \mu + \delta_i + \varepsilon_{ij}$
with $\delta_1 = 0$ (the first factor level is the reference group). The
coefficients returned are $\hat{\mu}$ and $\hat{\delta}_2, \ldots, \hat{\delta}_I$
— **not** the group means $\mu_1, \ldots, \mu_I$.

```{r}
#| label: default-param
aout_full <- aov(Percent ~ Judge, data = case0502)
coef(aout_full)
```

The interpretation: `(Intercept)` = $\hat{\mu}_A$ (judge A's mean);
`JudgeB` = $\hat{\mu}_B - \hat{\mu}_A$; and so on. The group means are
$\hat{\mu}_i = \hat{\mu} + \hat{\delta}_i$, but this parameterization
makes setting up coefficient vectors awkward.

## Forcing the Group-Means Parameterization

Subtracting 1 from the formula removes the intercept and forces R to fit
$Y_{ij} = \mu_i + \varepsilon_{ij}$ directly, returning each group mean:

```{r}
#| label: no-intercept
aout_lc <- aov(Percent ~ Judge - 1, data = case0502)
coef(aout_lc)
```

::: {.callout-warning}
## Use `- 1` only for `linearHypothesis()`

The no-intercept fit gives correct group means, but it **corrupts the ANOVA
table**: `summary()` and `anova()` will give wrong $F$-statistics and
$p$-values. Only switch to `~ Judge - 1` immediately before calling
`linearHypothesis()`. Use the ordinary `~ Judge` fit for everything else.
:::

## Checking the Factor Level Order

The coefficient vector must be in the **same order as the factor levels**:

```{r}
#| label: factor-levels
levels(case0502$Judge)
```

The order is `A, B, C, D, E, F, Spock's` — so position 7 is Spock's judge.

## Setting Up Coefficient Vectors

**Test 1:** Is Spock's mean different from the average of the other six?

$$H_0: \mu_{\text{Spock's}} - \tfrac{1}{6}\mu_A - \tfrac{1}{6}\mu_B - \cdots - \tfrac{1}{6}\mu_F = 0$$

```{r}
#| label: coef-vec1
combo_vec1 <- c(-1/6, -1/6, -1/6, -1/6, -1/6, -1/6, 1)
```

**Test 2:** Is the average of judges A–C equal to the average of judges D–F?

$$H_0: \tfrac{1}{3}\mu_A + \tfrac{1}{3}\mu_B + \tfrac{1}{3}\mu_C - \tfrac{1}{3}\mu_D - \tfrac{1}{3}\mu_E - \tfrac{1}{3}\mu_F = 0$$

```{r}
#| label: coef-vec2
combo_vec2 <- c(1/3, 1/3, 1/3, -1/3, -1/3, -1/3, 0)
```

## Running the Tests

```{r}
#| label: lh-test1
lhout1 <- linearHypothesis(model = aout_lc, hypothesis.matrix = combo_vec1)
lhout1
```

```{r}
#| label: lh-test2
lhout2 <- linearHypothesis(model = aout_lc, hypothesis.matrix = combo_vec2)
lhout2
```

Strong evidence that Spock's judge's mean differs from the pooled mean of the
other six ($p < 0.001$). No evidence that the first three non-Spock judges
(A–C) differ from the last three (D–F) ($p = 0.26$).

## Correspondence to the $t$-Test

`linearHypothesis()` returns an $F$-statistic. For a **single** linear
combination, $F = t^2$, so $|t| = \sqrt{F}$:

```{r}
#| label: t-from-f
sqrt(lhout1$F[2])
```

We can verify this manually. First, compute $\hat\gamma$ and $SE(\hat\gamma)$:

```{r}
#| label: manual-t
#| message: false
sumdat <- case0502 |>
  group_by(Judge) |>
  summarise(ybar = mean(Percent), n = n())

gamma_hat <- sum(combo_vec1 * sumdat$ybar)

sp  <- sqrt(sum(aout_lc$residuals^2) / (nrow(case0502) - 7))
se_gamma <- sp * sqrt(sum(combo_vec1^2 / sumdat$n))

tstat <- gamma_hat / se_gamma

cat("gamma_hat =", round(gamma_hat, 3), "\n",
    "SE       =", round(se_gamma, 3), "\n",
    "t*       =", round(tstat, 3), "\n")
```

```{r}
#| label: pvalue-match
2 * pt(-abs(tstat), df = nrow(case0502) - 7)
lhout1$`Pr(>F)`[2]
```

The $p$-values are identical, confirming the equivalence.

