---
title: "Linear Combinations of Group Means"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
set.seed(8)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(multcomp)
library(broom)
```

# Learning Objectives

- Define a linear combination and a contrast of group means.
- Estimate a linear combination and compute its standard error.
- Conduct a hypothesis test and form a confidence interval for a linear
  combination.
- Use `glht()` from the `multcomp` package to test linear combinations
  in R.

---

# The Spock Conspiracy Trial (Recap)

```{r}
#| label: load-data
#| message: false
case0502 <- read_csv("https://dcgerard.github.io/stat_302/data/case0502.csv") |>
  mutate(Judge = factor(Judge))
```

The model is $Y_{ij} = \mu_i + \varepsilon_{ij}$, where $Y_{ij}$ is the
percent women on venire $j$ of judge $i$, $\mu_i$ is judge $i$'s true mean,
and $\varepsilon_{ij}$ is individual noise with variance $\sigma^2$. There
are $I = 7$ judges and $n = 46$ venires.

The omnibus $F$-test from Chapter 5 gave very strong evidence that the seven
judges do not all have the same mean ($p < 0.001$). Now we want to ask more
specific questions.

---

# Linear Combinations

## Motivation

We want to compare Spock's judge ($\mu_1$) to the average of the other six
judges ($\mu_2, \ldots, \mu_7$), **without** having to assume those six means
are equal. That is:

$$H_0: \mu_1 = \tfrac{1}{6}(\mu_2 + \mu_3 + \cdots + \mu_7),
\qquad
H_A: \mu_1 \neq \tfrac{1}{6}(\mu_2 + \mu_3 + \cdots + \mu_7)$$

Rewriting the null:

$$\mu_1 - \tfrac{1}{6}\mu_2 - \tfrac{1}{6}\mu_3 - \cdots - \tfrac{1}{6}\mu_7 = 0$$

The left-hand side is a **linear combination** of the group means.

## Definition

A **linear combination** of $\mu_1, \mu_2, \ldots, \mu_I$ with coefficients
$c_1, c_2, \ldots, c_I$ is:

$$\gamma = c_1\mu_1 + c_2\mu_2 + \cdots + c_I\mu_I$$

If the coefficients sum to zero ($c_1 + c_2 + \cdots + c_I = 0$), the linear
combination is called a **contrast**.

For the Spock comparison: $c_1 = 1$ and $c_2 = c_3 = \cdots = c_7 = -\tfrac{1}{6}$.
Check: $1 + 6 \times (-\tfrac{1}{6}) = 0$. So this is a contrast.

## Hypotheses

For any linear combination $\gamma$, the hypotheses are:

$$H_0: \gamma = 0, \qquad H_A: \gamma \neq 0$$

## Estimation

Estimate each $\mu_i$ with the group sample mean $\bar{Y}_{i\bullet}$:

$$\hat{\gamma} = c_1 \bar{Y}_{1\bullet} + c_2 \bar{Y}_{2\bullet} + \cdots + c_I \bar{Y}_{I\bullet}$$

## Standard Error

Because the group means are independent and
$\mathrm{Var}(\bar{Y}_{i\bullet}) = \sigma^2 / n_i$:

$$\mathrm{Var}(\hat{\gamma}) = \sigma^2 \left(\frac{c_1^2}{n_1} + \frac{c_2^2}{n_2} + \cdots + \frac{c_I^2}{n_I}\right)$$

We estimate $\sigma$ with the pooled standard deviation $s_p$ from the full
model, giving:

$$SE(\hat{\gamma}) = s_p \sqrt{\frac{c_1^2}{n_1} + \frac{c_2^2}{n_2} + \cdots + \frac{c_I^2}{n_I}}$$

## Hypothesis Test

Under $H_0$, the standardized estimate follows a $t$ distribution:

$$t^* = \frac{\hat{\gamma}}{SE(\hat{\gamma})} \;\sim\; t_\nu, \quad \nu = n - I$$

The two-sided $p$-value is $2 \cdot P(t_\nu \leq -|t^*|)$.

## Confidence Interval

A 95% confidence interval for $\gamma$:

$$\hat{\gamma} \;\pm\; t_\nu(0.975) \cdot SE(\hat{\gamma})$$

---

# Implementation in R

## The Default Parameterization

Fitting a one-way ANOVA with `aov()` uses R's default treatment-contrast
parameterization: the model is $Y_{ij} = \mu + \delta_i + \varepsilon_{ij}$
with $\delta_1 = 0$ (the first factor level is the reference group). The
coefficients returned are $\hat{\mu}$ and $\hat{\delta}_2, \ldots, \hat{\delta}_I$
— **not** the group means $\mu_1, \ldots, \mu_I$.

```{r}
#| label: default-param
aout_full <- aov(Percent ~ Judge, data = case0502)
coef(aout_full)
```

The interpretation: `(Intercept)` = $\hat{\mu}_A$ (judge A's mean);
`JudgeB` = $\hat{\mu}_B - \hat{\mu}_A$; and so on. The group means are
$\hat{\mu}_i = \hat{\mu} + \hat{\delta}_i$, but this parameterization
makes setting up coefficient vectors awkward.

## Forcing the Group-Means Parameterization

Subtracting 1 from the formula removes the intercept and forces R to fit
$Y_{ij} = \mu_i + \varepsilon_{ij}$ directly, returning each group mean:

```{r}
#| label: no-intercept
aout_lc <- aov(Percent ~ Judge - 1, data = case0502)
coef(aout_lc)
```

::: {.callout-warning}
## Use `- 1` only for `linearHypothesis()`

The no-intercept fit gives correct group means, but it **corrupts the ANOVA
table**: `summary()` and `anova()` will give wrong $F$-statistics and
$p$-values. Only switch to `~ Judge - 1` immediately before calling
`linearHypothesis()`. Use the ordinary `~ Judge` fit for everything else.
:::

## Checking the Factor Level Order

The coefficient vector must be in the **same order as the coefficient estimates**:

```{r}
#| label: factor-levels
coef(aout_lc)
```

The order is `A, B, C, D, E, F, Spock's` — so position 7 is Spock's judge.

## Setting Up Coefficient Vectors

- Is Spock's mean different from the average of the other six?

  $$H_0: \mu_{\text{Spock's}} - \tfrac{1}{6}\mu_A - \tfrac{1}{6}\mu_B - \cdots - \tfrac{1}{6}\mu_F = 0$$

-   We set up the coefficient vector. The syntax for the `{multcomp}` package requires it to be a matrix. Just modify this syntax for future use:
    ```{r}
    #| label: coef-vec1
    combo_vec1 <- matrix(c(-1/6, -1/6, -1/6, -1/6, -1/6, -1/6, 1), nrow = 1)
    ```


## Running the Tests

-   Put the model object as `model` and the coefficients as `linfct`:
    ```{r}
    #| label: lh-test1
    glht(model = aout_lc, linfct = combo_vec1) |>
      tidy(conf.int = TRUE)
    ```



- Strong evidence that Spock's judge's mean differs from the pooled mean of the other six ($p < 0.001$). With 95% confidence, we estimate that Spock's Judge's venires have an average of 9.6 to 20.3 percentage-points fewer women than the average of the other judges.

# Other Contrasts of Interest

## Comparing Averages of Group Means

- The Spock hypothesis above is an example of comparing averages of group means.
  - We had two groups in that case, Spock's judge (group 1), and all other judges (group 2).
  
- More generally, we can compare groups 1.1,1.2,...,1.$J$ to groups 2.1,2.2,...,2.$K$ 
  - $J$ groups in group 1
  - $K$ groups in group 2.
  
- Suppose the means are $\mu_{1.1}, \mu_{1.2},\ldots,\mu_{1.J}$ and $\mu_{2.1}, \mu_{2.2},\ldots,\mu_{2.K}$. Then the contrast is
  $$
  \gamma = \frac{\mu_{1.1} + \cdots + \mu_{1.J}}{J} - \frac{\mu_{2.1}+\cdots+\mu_{2.K}}{K}.
  $$

- The $J$ groups have coefficients $1/J$ and the $K$ groups have coefficients $-1/K$.

- E.g., suppose we want to test if the average of means of Judges A, B, and C was the same as the average of means of Judges C, D, and E
  - We are not really interested in this hypothesis. This is for demonstration purposes.
  
  $$H_0: \tfrac{1}{3}\mu_A + \tfrac{1}{3}\mu_B + \tfrac{1}{3}\mu_C - \tfrac{1}{3}\mu_D - \tfrac{1}{3}\mu_E - \tfrac{1}{3}\mu_F = 0$$

-   Set up aov object and constrast vector:
    ```{r}
    #| label: coef-vec2
    aout_lc <- aov(Percent ~ Judge - 1, data = case0502)
    coef(aout_lc)
    combo_vec2 <- matrix(c(1/3, 1/3, 1/3, -1/3, -1/3, -1/3, 0), nrow = 1)
    ```

-   Run linear hypothesis test:
    ```{r}
    #| label: lh-test2
    glht(model = aout_lc, linfct = combo_vec2) |>
      tidy(conf.int = TRUE)
    ```
  
- Some evidence that the first three non-Spock judges
(A–C) differ from the last three (D–F) ($p = 0.045$).

## Pairwise Comparisons

- A pairwise comparison is an example of a contrast.

- E.g., suppose we want to just compare Spock's Judge to Judge F, then the contrast is

  $$
  \gamma = \mu_{\text{Spock's}} - \mu_{F}
  $$

- And we are testing the null

  $$
  H_0: \gamma = 0
  $$
  
- We learned how to do this using `pairwise.t.test()`, but we can also use the linear hypothesis framework

-   Set up aov object and constrast vector:
    ```{r}
    #| label: coef-vec3
    aout_lc <- aov(Percent ~ Judge - 1, data = case0502)
    coef(aout_lc)
    combo_vec3 <- matrix(c(0, 0, 0, 0, 0, -1, 1), nrow = 1)
    ```

-   Run linear hypothesis test:
    ```{r}
    #| label: lh-test3
    glht(model = aout_lc, linfct = combo_vec3) |>
      tidy(conf.int = TRUE)
    ```
-   It's the same $p$-value in the bottom right of this table:
    ```{r}
    pairwise.t.test(case0502$Percent, case0502$Judge, p.adjust.method = "none")
    ```

::: {.panel-tabset}
## Exercise
What would be the contrast for comparing Spock's judge to Judge A?

## Solution
```{r}
matrix(c(-1, 0, 0, 0, 0, 0, 1), nrow = 1)
```
:::

::: {.panel-tabset}
## Exercise
Is this a contrast?
$$
\gamma = -\mu_1 + -2 \mu_2 + \mu_3
$$

## Solution
No. The coefficients add up to -2, not 0.
:::

## Comparing Rates

-   Recall the diet restriction case study from Chapter 5. They gave mice different restricted diets and measured their life time.
    ```{r}
    #| message: false
    mice <- read_csv("https://dcgerard.github.io/stat_302/data/case0501.csv")
    ```

- What if researchers wanted to measure the amount of lifetime gained per kcal reduction in diet?

-   E.g., we have two mice, one on an 85 kcal/wk diet (`N/N85`) and the other on a 50 kcal/wk diet (`N/R50`). 

- The contrast of interest is
$$
\frac{\mu_{\text{N/R50}} - \mu_{\text{N/N85}}}{85 - 50}
$$
-   Let's verify the order of coefficient estimates:
    ```{r}
    aout_mice <- aov(Lifetime ~ Diet - 1, data = mice)
    coef(aout_mice)
    ```

-   Set up the contrast

    ```{r}
    combo_mice_1 <- matrix(c(0, -1/35, 0, 1/35, 0, 0), nrow = 1)
    ```

-   Get a confidence interval
    ```{r}
    glht(model = aout_mice, linfct = combo_mice_1) |>
      tidy(conf.int = TRUE)
    ```

- So for a 1 kcal reduction per week, we can expect on average an increase of 0.27 months of life (95% CI: 0.21 months to 0.34 months).

- If we want to compare the rate from 85 kcal/wk to 50 kcal/week with the rate from 50 kcal/wk to 40 kcal/wk, we set up the contrast:

  $$
  \frac{\mu_{\text{N/R50}} - \mu_{\text{N/N85}}}{85 - 50} - \frac{\mu_{\text{N/R40}} - \mu_{\text{N/R50}}}{50-40} = -\frac{1}{35}\mu_{\text{N/N85}} + \frac{9}{70}\mu_{\text{N/R50}} - \frac{1}{10}\mu_{\text{N/R40}}
  $$
-   This will test if the rates are the same for both reductions. We can set up this test in R with
    ```{r}
    coef(aout_mice)
    combo_mice_2 <- matrix(c(0, -1/35, -1/10, 9/70, 0, 0), nrow = 1)
    glht(model = aout_mice, linfct = combo_mice_2) |>
          tidy(conf.int = TRUE)
    ```

- We have no evidence that rates differ going from 85 kcal/wk to 50 kcal/wk and going from 50 kcal/wk to 40 kcal/wk ($p$ = 0.96).

# Linear trends

- Skip this section in the *Statistical Sleuth*. Nobody does it.
