---
title: "Chapter 6 Worksheet"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(car)
knitr::opts_chunk$set(echo       = TRUE,
                      eval       = FALSE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r}
#| label: load-data
#| eval: true
#| message: false
case0502 <- read_csv("https://dcgerard.github.io/stat_302/data/case0502.csv") |>
  mutate(Judge = factor(Judge))
sardine <- read_csv("https://dcgerard.github.io/stat_302/data/sardine.csv") |>
  mutate(location = factor(location,
                           labels = c("Alaska", "Brit. Columbia",
                                      "San Francisco", "Monterey",
                                      "San Pedro", "San Diego")))
```

# Linear Combinations (Spock Data)

**1.** In the Spock data, suppose researchers wanted — **before looking at the
data** — to test whether the three judges with the highest average percent-women
in the dataset (A, B, E) have a different mean from the three with the lowest
(C, D, F), ignoring Spock's judge entirely.

  a. Write out the coefficient vector for this contrast (using the factor level
     order A, B, C, D, E, F, Spock's). Does it qualify as a contrast?

  b. Use `glht()` to test this hypothesis. Report the $F$-statistic,
     $p$-value, and your conclusion.

```{r}

```

**2.** The hypothesis $H_0: \mu_A = \mu_B$ is a pairwise comparison — a
special case of a linear combination.

  a. What is the coefficient vector (in the order A, B, C, D, E, F, Spock's)?

  b. Verify using `glht()` that the $p$-value matches what
     `pairwise.t.test()` gives (with `p.adjust.method = "none"`).

```{r}

```

**3.** What is the estimated value of $\gamma$ for the contrast that compares
Spock's mean to the average of the other six judges? Interpret this number in
the context of the study.

**4.** Compute a 95% confidence interval for $\gamma$ (Spock's mean minus the
average of the other six means) using the formula
$\hat{\gamma} \pm t_{n-I}(0.975) \cdot SE(\hat\gamma)$.
What does this interval tell you about how different Spock's judge is?

```{r}

```

---

# Multiple Comparisons (Sardine Data)

These exercises use vertebrae counts from Pacific sardines sampled at six
locations along the Pacific coast. [@clarke1936variations]

**5.** Make side-by-side boxplots of `number` (vertebrae count) by `location`.
Do you think the equal-variance and normality assumptions are satisfied well
enough to proceed with ANOVA? Explain.

```{r}

```

**6.** Run the omnibus $F$-test with `aov()`. How many pairwise comparisons
would be needed to test all pairs of locations? How many spurious rejections
would you expect at $\alpha = 0.05$ if all location means were truly equal?

```{r}

```

**7.** Run all pairwise $t$-tests with **no** $p$-value adjustment. Which
pairs have a raw $p$-value below 0.05?

```{r}

```

**8.** Apply the **Tukey-Kramer** adjustment with `TukeyHSD()`. Which pairs
now show evidence of a difference? Produce the confidence-interval plot.
How do the results compare to the unadjusted tests?

```{r}

```

**9.** Apply the **Holm** adjustment using `pairwise.t.test()`. Compare these
adjusted $p$-values to the Tukey-Kramer values — are the conclusions the same?

```{r}

```

**10.** Suppose researchers had suspected *before collecting data* that sardines
from colder waters (Alaska and British Columbia) have more vertebrae than those
from warmer waters (San Francisco, Monterey, San Pedro, San Diego). This is a
single pre-planned contrast.

  a. Write the null hypothesis in terms of the six location means
     $\mu_1, \ldots, \mu_6$.

  b. Check the factor level order of `location`, then set up the coefficient
     vector.

```{r}

```

**11.** Test the planned contrast from Exercise 10 using `glht()`
(refit with `~ location - 1` first). Report the $F$-statistic, $p$-value,
and a plain-language conclusion. Because this is a single pre-planned
comparison, no multiplicity adjustment is needed.

```{r}

```

**12.** Explain in your own words why the Tukey-Kramer procedure gives smaller
adjusted $p$-values than Bonferroni for all-pairwise comparisons, while
Bonferroni is more general. In which situation would you prefer Bonferroni
over Tukey-Kramer?
