---
title: "Voltage Case Study"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 2.3, 
                      fig.width = 3.7, 
                      fig.align = "center")
```


## Objectives

- Analyze Voltage vs Breakdown Time Case Study

- Lack of Fit $F$-test

## Case Study: Voltage vs Breakdown Time, a Controlled Experiment

- Goal: study relationship between voltage and breakdown time of an electrical insulating fluid.

- The authors could control the voltage level of each trial.

```{r, echo = TRUE}
library(Sleuth3)
data("case0802")
head(case0802)
```


## Step 1: Make a plot

```{r, echo = TRUE, message=FALSE}
library(ggplot2)
qplot(Voltage, Time, data = case0802) + 
  geom_smooth(se = FALSE)
```


## Try an initial fit with a residual plot

```{r, echo=TRUE}
lmout <- lm(Time ~ Voltage, data = case0802)
resvec <- resid(lmout)
fitvec <- fitted(lmout)
qplot(fitvec, resvec) + geom_hline(yintercept = 0)
```

## Conclusion?

- As the mean increases, the variability increases.

- We see a curved relationship between $X$ and $Y$

- Clearly need a log-transformation

## Log-transformation and Re-plot

```{r, echo = TRUE, message=FALSE}
case0802$logTime <- log(case0802$Time)
qplot(Voltage, logTime, data = case0802) + 
  geom_smooth(se = FALSE)
```


## Log-transformation and Residual Plot

```{r, echo=TRUE}
lmout <- lm(logTime ~ Voltage, data = case0802)
resvec <- resid(lmout)
fitvec <- fitted(lmout)
qplot(fitvec, resvec) + geom_hline(yintercept = 0)
```

## Conclusion

- After the log-transformation, the data look pretty awesome.

## Formal test if there is a relationship

- There is clearly a relationship here, but you need to report $p$-values to get published, so ...

\small
```{r, echo = TRUE}
sumout <- summary(lmout)
coef(sumout)
```
\normalsize

## More interesting are coefficient estimates with confidence intervals

```{r, echo = TRUE}
cbind(coef(lmout), confint(lmout))
```

## Interpret on Original Scale

- A one kV increase results in a $\exp(-0.507) = 0.6$ multiplicative change in breakdown times. 

- 95% confidence of 
```{r, echo = TRUE}
exp(confint(lmout)[2, ])
```

- A one kV increase results in a 40% decrease in breakdown time, 95% confidence interval of 
```{r, echo = TRUE}
(1 - exp(confint(lmout)[2, ])) * 100
```

# Lack of Fit $F$-test

## We could have viewed this as an ANOVA problem

```{r, echo = TRUE}
case0802$VoltageFac <- as.factor(case0802$Voltage)
qplot(VoltageFac, logTime, data = case0802, geom = "boxplot") +
  xlab("Voltage")
```

## Which is better?

- If the linear model appears to fit fine, it is **always preferred**.

- You can interpolate with the linear model (not ANOVA).

- The linear model has easier interpretations.

- The linear model has fewer parameters

- We can formally test if the linear model does not fit using the $F$-testing strategy if we have replicates at given values of $X$

## Lack-of-fit $F$-test

- $H_0: E[Y_i] = \beta_0 + \beta_1X_i$ (mean is based on line)

- $H_A: E[Y_i] = \mu_j$ where $X_i = j$ (mean is based on group)

## Full Model Estimates

```{r, message=FALSE}
library(tidyverse)
case0802 %>%
  group_by(VoltageFac) %>%
  summarize(mean = mean(logTime)) %>%
  mutate(z = as.numeric(VoltageFac)) %>%
  mutate(zlower = z - 1/3, zupper = z + 1/3) ->
  meandf
case0802$jitterx <- jitter(as.numeric(case0802$VoltageFac))

meandf %>%
  select(VoltageFac, mean) %>%
  left_join(case0802, by = "VoltageFac") ->
  case0802

qplot(x = jitterx, y = logTime, data = case0802, xlab = "Voltage") +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_segment(data = meandf, mapping = aes(x = zlower, xend = zupper, y = mean, yend = mean), col = 2, lty = 2)
```

## Reduced Model Estimates

```{r, message=FALSE}
case0802$fitted <- fitted(lmout)
qplot(jitterx, logTime, data = case0802) +
  xlab("Voltage") +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_line(mapping = aes(x = (Voltage - 24)/2, y = fitted), lty = 2, col = 2) 
```

## Residuals Fnder Full

```{r}
qplot(x = jitterx, y = logTime, data = case0802, xlab = "Voltage") +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_segment(data = meandf, mapping = aes(x = zlower, xend = zupper, y = mean, yend = mean), col = 2, lty = 2) +
  geom_segment(mapping = aes(x = jitterx, xend = jitterx, y = logTime, yend = mean), col = "blue")
```

## Residuals Fnder Reduced

```{r}
qplot(jitterx, logTime, data = case0802) +
  xlab("Voltage") +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_line(mapping = aes(x = (Voltage - 24)/2, y = fitted), lty = 2, col = 2)  +
  geom_segment(mapping = aes(x = jitterx, xend = jitterx, y = logTime, yend = fitted), col = "blue")
```

## Lack-of-fit $F$-test

```{r}
rss_full <- sum((case0802$logTime - case0802$mean)^2)
rss_reduced <- sum((case0802$logTime - case0802$fitted)^2)
ESS <- rss_reduced - rss_full
df_full <- 69
df_reduced <- 74
df_extra <- df_reduced - df_full
fstat <- (ESS / df_extra) / (rss_full / df_full)
```

- $RSS_{full}$ =  `r rss_full`

- $df_{full} = n - I$ =  76 - 7 = 69\pause

- $RSS_{reduced}$ = `r rss_reduced`

- $df_{reduced} = n - 2$ =  76 - 2 = 74\pause

- $ESS = RSS_{reduced} - RSS_{full}$ =  `r ESS`

- $df_{extra} = df_{reduced} - df_{full}$ = 74 - 69 - 5\pause

- $F$-statistic = $\frac{ESS / df_{extra}}{RSS_{full} / df_{full}}$ = `r fstat`

## Compare to an $F_{5, 69}$

```{r}
x <- seq(0, 5, length = 100)
y <- df(x, 5, 69)

which_x <- x > fstat
polydf <- data.frame(polyx = c(min(x[which_x]), x[which_x], max(x[which_x]), min(x[which_x])),
                     polyy = c(0, y[which_x], 0, 0))
qplot(x, y, geom = "line") +
  geom_polygon(data = polydf, mapping = aes(x = polyx, y = polyy), 
               fill = "blue", alpha = 3/4) +
  xlab("F-stat") +
  ylab("Density")
```

## Calculate $p$-value

```{r, echo = TRUE}
pf(0.5024, df1 = 5, df2 = 69, lower.tail = FALSE)
```


## Lack of Fit in R

- Create a factor variable
```{r, echo = TRUE}
case0802$VoltageFac <- as.factor(case0802$Voltage)
```

- Fit both the ANOVA and regression models
```{r, echo = TRUE}
aout  <- aov(logTime ~ VoltageFac, data = case0802)
lmout <- lm(logTime ~ Voltage, data = case0802)
```


## Lack of Fit in R

- Use `anova()` to get ANOVA table
```{r, echo = TRUE}
anova(lmout, aout)
```







