---
title: "Linear Model Assumptions and Transformations"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r}
#| label: setup
#| include: false
set.seed(11)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 5,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(1)
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(broom)
library(lm.br)
library(quantreg)
library(sandwich)
library(lmtest)
```

# Learning Objectives

- Understand the assumptions of simple linear regression and their relative
  importance.
- Evaluate assumptions using scatterplots and residual plots.
- Choose and apply appropriate remedies when assumptions are violated.
- Interpret log-transformations of the explanatory and/or response variable.
- Use sandwich standard errors to handle non-constant variance.
- Perform a lack-of-fit $F$-test to assess whether a linear model is adequate.
- Chapter 8 in the book.

---

# Assumptions of Simple Linear Regression

## Assumptions in Decreasing Order of Importance

1. **Linearity** — Does the relationship look like a straight line?
2. **Independence** — Knowledge of one observation gives no information about
   another.
3. **Equal Variance** — The spread is the same for every value of $x$.
4. **Normality** — The distribution is not too skewed and there are no extreme
   outliers. (Only a concern when outliers are present and the sample size is
   small, thanks to the CLT.)

## Problems When Violated

1. **Linearity** — The regression line does not capture the actual relationship.
2. **Independence** — The regression line is unbiased, but standard errors are
   wrong.
3. **Equal Variance** — The regression line is unbiased, but standard errors are
   wrong.
4. **Normality** — Unstable results when outliers are present and the sample size
   is small.

## Assessment Tools: Scatterplots and Residual Plots

- Make a **scatterplot** of $x$ (horizontal) vs $y$ (vertical) to check for
  non-linearity, unequal variance, and normality violations.
- A **residual plot** (residuals on the $y$-axis, fitted values on the $x$-axis)
  is often clearer because the linear signal is removed.

---

# Demonstrating the Assumptions

## Dataset 1: Gold Standard

Generate some fake data:

```{r}
#| label: d1-gen
x <- rnorm(100, sd = 1)
y <- x + rnorm(100)
df <- tibble(x = x, y = y)
```

```{r}
#| label: d1-scatter
#| message: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
#| label: d1-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- The mean lies on a straight line.
- Residuals are centered at 0 for all $x$.
- Variance looks equal for all $x$.
- Everything looks ideal.

### Don't Get Fooled by the Distribution of $x$

The distribution of $x$ does not matter. The data below are **also** ideal:

```{r}
#| label: d1a-gen
x <- sample(1:3, size = 100, replace = TRUE)
y <- x + rnorm(100)
df <- tibble(x = x, y = y)
```

```{r}
#| label: d1a-scatter
#| message: false
#| warning: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE, method = lm)
```

```{r}
#| label: d1a-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

## Dataset 2: Curved Monotone Relationship, Equal Variances

Generate some fake data:

```{r}
#| label: d2-gen
x <- rexp(100)
x <- x - min(x) + 0.5
y <- log(x) * 20 + rnorm(100, sd = 4)
df <- tibble(x = x, y = y)
```

```{r}
#| label: d2-scatter
#| message: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
#| label: d2-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- Curved but always-increasing relationship between $x$ and $y$.
- Variance looks equal for all $x$.
- Residual plot has a parabolic shape — a $\log$ transformation of $x$ could
  help.

```{r}
#| label: d2-log-scatter
#| message: false
df |>
  mutate(x_log = log(x)) ->
  df
ggplot(df, aes(x = x_log, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(x = "log(x)")
```

```{r}
#| label: d2-log-resid
lmout <- lm(y ~ x_log, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

After the $\log$ transformation of $x$, the residual plot looks flat and
even — the linearity and equal-variance assumptions are now satisfied.

## Dataset 3: Curved Non-Monotone Relationship, Equal Variances

Generate some fake data:

```{r}
#| label: d3-gen
x <- rnorm(100)
y <- -x^2 + rnorm(100)
df <- tibble(x = x, y = y)
```

```{r}
#| label: d3-scatter
#| message: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
#| label: d3-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- Curved relationship that sometimes increases, sometimes decreases.
- Variance looks equal for all $x$.
- Residual plot has a parabolic form.

**Solution 1:** Add a quadratic term:
$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + \varepsilon_i$$

```{r}
#| label: d3-quad-fit
quad_lm <- lm(y ~ x + I(x^2), data = df)
df |>
  mutate(fitted = fitted(quad_lm)) ->
  df
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(x = x, y = fitted), color = "blue", linewidth = 1)
```

```{r}
#| label: d3-quad-resid
augment(quad_lm) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

**Solution 2:** Fit a broken-line (piecewise linear) model using the `lm.br`
package:
$$Y_i = \begin{cases} \beta_0 + \beta_1 X_i + \varepsilon_i & \text{if } X_i < C \\ \beta_0^* + \beta_1^* X_i + \varepsilon_i & \text{if } X_i > C \end{cases}$$

```{r}
#| label: d3-lmbr
#| message: false
lmbr_out <- lm.br(y ~ x, data = df)
df |>
  mutate(fitted = fitted(lmbr_out)) ->
  df
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(x = x, y = fitted), color = "blue", linewidth = 1)
```

```{r}
#| label: d3-lmbr-resid
tibble(.fitted = fitted(lmbr_out), .resid = resid(lmbr_out)) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

## Dataset 4: Curved Relationship, Variance Increases with $Y$

Generate some fake data:

```{r}
#| label: d4-gen
x <- rnorm(100)
y <- exp(x + rnorm(100, sd = 1/2))
df <- tibble(x = x, y = y)
```

```{r}
#| label: d4-scatter
#| message: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
#| label: d4-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- Curved relationship between $x$ and $y$.
- Variance increases as $y$ increases.
- Residual plot has a parabolic shape and shows increasing spread to the right.

**Solution:** Take a log-transformation of $y$.

```{r}
#| label: d4-log-scatter
#| message: false
df |>
  mutate(y_log = log(y)) ->
  df
ggplot(df, aes(x = x, y = y_log)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(y = "log(y)")
```

```{r}
#| label: d4-log-resid
lmout <- lm(y_log ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

After taking $\log(y)$, both the linearity and equal-variance problems are
resolved.

### Again, Don't Get Fooled by the Distribution of $x$

You should still look for increasing variance as $x$ increases, even when the
distribution of $x$ is discrete.

```{r}
#| label: d4a-gen
x <- sample(1:3, size = 100, replace = TRUE)
y <- exp(x + rnorm(100, sd = 1/2))
df <- tibble(x = x, y = y)
```

```{r}
#| label: d4a-scatter
#| message: false
#| warning: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
#| label: d4a-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- Curved relationship between $x$ and $y$.
- Variance increases as $y$ increases.
- Residual plot has a parabolic shape and shows increasing spread to the right.

**Solution:** Take a log-transformation of $y$.

```{r}
#| label: d4a-log-scatter
#| message: false
#| warning: false
df |>
  mutate(y_log = log(y)) ->
  df
ggplot(df, aes(x = x, y = y_log)) +
  geom_point() +
  geom_smooth(se = FALSE, method = lm) +
  labs(y = "log(y)")
```

```{r}
#| label: d4a-log-resid
lmout <- lm(y_log ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

After taking $\log(y)$, both the linearity and equal-variance problems are
resolved.

## Dataset 5: Linear Relationship, Equal Variances, Skewed Errors

Generate some fake data:

```{r}
#| label: d5-gen
x <- runif(200)
y <- 15 * x + rexp(200, 0.2)
df <- tibble(x = x, y = y)
```

```{r}
#| label: d5-scatter
#| message: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
#| label: d5-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- Straight-line relationship between $x$ and $y$.
- Variances are approximately equal for all $x$.
- The errors are right-skewed — visible in the residual plot.

**Solutions:**

1. Do nothing, but report the skewness (usually acceptable for large $n$).
2. Fit a **quantile regression** for the median:
   $$\text{Median}(Y_i \mid X_i) = \beta_0 + \beta_1 X_i$$

```{r}
#| label: d5-quantreg
#| message: false
qr_out <- rq(y ~ x, tau = 0.5, data = df)

df |>
  mutate(qr_fitted = fitted(qr_out)) ->
  df
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_line(aes(x = x, y = qr_fitted), color = "red", linewidth = 1)
```

The quantile regression line (red) is not very different from the OLS line
(blue) — skewness with equal variance has little practical effect on the
coefficient estimates.

## Dataset 6: Linear Relationship, Unequal Variances

Generate some fake data:

```{r}
#| label: d6-gen
x <- runif(100) * 10
y <- 0.85 * x + rnorm(100, sd = (x - 5)^2)
df <- tibble(x = x, y = y)
```

```{r}
#| label: d6-scatter
#| message: false
ggplot(df, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
#| label: d6-resid
lmout <- lm(y ~ x, data = df)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- Linear relationship between $x$ and $y$.
- Variance differs across values of $x$ — clearly visible in the residual plot.

**Solution:** Use **sandwich** standard errors.

```{r}
#| label: d6-sandwich
#| message: false
coeftest(lmout, vcov = sandwich) |> tidy()
```

Compare to the standard (non-robust) standard errors:

```{r}
#| label: d6-standard
tidy(lmout)
```

Confidence intervals with sandwich standard errors:

```{r}
#| label: d6-sandwich-ci
coefci(lmout, vcov = sandwich)
```

## Intuition Behind the Sandwich Estimator of Variance

Consider a simplified model with a zero intercept: $Y_i = \beta_1 X_i + \varepsilon_i$.

Using calculus:
$$\hat{\beta}_1 = \frac{\sum_{i=1}^n X_i Y_i}{\sum_{i=1}^n X_i^2}$$

So:
$$\text{Var}(\hat{\beta}_1) = \text{Var}\!\left(\frac{\sum_{i=1}^n X_i Y_i}{\sum_{i=1}^n X_i^2}\right) = \frac{\sum_{i=1}^n X_i^2\,\text{Var}(Y_i \mid X_i)}{\left(\sum_{i=1}^n X_i^2\right)^2}$$

- **Usual method:** Estimate $\text{Var}(Y_i \mid X_i)$ with $\hat{\sigma}^2$ — assumes the same variance for all $i$.
- **Sandwich method:** Estimate $\text{Var}(Y_i \mid X_i)$ with $(Y_i - \hat{\beta}_1 X_i)^2$ — allows a different variance at each $i$.

The sandwich estimator yields accurate standard errors for the coefficient
estimates as long as:

1. The linearity assumption holds, and
2. The sample size is large enough.

::: {.callout-warning}
Sandwich standard errors **cannot** be used for prediction intervals.
:::

---

# Interpreting Log Transformations

## No Log on $X$, No Log on $Y$

**Model:** $\mu(Y \mid X) = \beta_0 + \beta_1 X$

At $X + 1$:
$$\mu(Y \mid X+1) = \beta_0 + \beta_1(X+1) = \mu(Y \mid X) + \beta_1$$

- **Observational study:** $\beta_1$ is the mean difference in $Y$ between
  groups that differ by one unit in $X$.
- **Randomized experiment:** $\beta_1$ is the mean increase in $Y$ when $X$ is
  increased by one unit.

## No Log on $X$, Log on $Y$

**Model:** $\mu(\log(Y) \mid X) = \beta_0 + \beta_1 X$

If residuals are fairly symmetric, then
$\text{Median}(\log Y \mid X) \approx \mu(\log Y \mid X)$, so:

$$\text{Median}(Y \mid X) = \exp(\beta_0 + \beta_1 X)
= \exp(\beta_0)\exp(\beta_1 X)$$

At $X + 1$:
$$\text{Median}(Y \mid X+1) = \text{Median}(Y \mid X) \cdot \exp(\beta_1)$$

- **Observational study:** $\exp(\beta_1)$ is the ratio of medians when groups
  differ by one unit in $X$.
- **Randomized experiment:** $\exp(\beta_1)$ is the multiplicative change in median $Y$ when $X$ increases by one unit.

### Example: Voltage and Breakdown Time

Researchers applied different voltage levels to an insulating fluid and
measured the time until the fluid broke down.

```{r}
#| label: volt-ex-load
#| message: false
voltage <- read_csv("https://dcgerard.github.io/stat_302/data/case0802.csv")
```

The relationship is curved with unequal variance:

```{r}
#| label: volt-ex-scatter
#| message: false
ggplot(voltage, aes(x = Voltage, y = Time)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

A log-transformation of $Y$ works best:

```{r}
#| label: volt-ex-log-scatter
#| message: false
ggplot(voltage, aes(x = Voltage, y = Time)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_y_log10()
```

We fit the model with $\log(\text{Time})$ as the response:

```{r}
#| label: volt-ex-fit
lm(I(log(Time)) ~ Voltage, data = voltage) |>
  tidy(conf.int = TRUE)
```

- $\hat\beta_1 = -0.51$, with 95% CI $(-0.62, -0.39)$.
- $\exp(-0.51) = 0.60$, so a 1 kV increase in voltage is associated with a
  median breakdown time that is 0.60 times as large — a 40% decrease:
  $$
  \text{Median}(Y \mid X+1) - \text{Median}(Y \mid X)
  = (\exp(\beta_1) - 1)\,\text{Median}(Y \mid X) = -0.4\,\text{Median}(Y \mid X)
  $$
- The 95% CI for the multiplicative effect is
  $(\exp(-0.62), \exp(-0.39)) = (0.54, 0.68)$ — a 32% to 46% decrease.

## Log on $X$, No Log on $Y$

**Model:** $\mu(Y \mid X) = \beta_0 + \beta_1 \log(X)$

When $X$ is doubled:

$$\mu(Y \mid 2X) = \beta_0 + \beta_1\log(2X) = \mu(Y \mid X) + \beta_1\log(2)$$

- **Observational study:** $\beta_1\log(2)$ is the mean difference in $Y$ when one group's $X$ is twice the other's.
- **Randomized experiment:** $\beta_1\log(2)$ is the increase in mean $Y$ when $X$ is doubled.

A similar interpretation holds for $\beta_1\log(10)$ and a 10-fold increase in
$X$.

### Example: pH and Time After Slaughter

Researchers measured the pH of steer carcasses at various times after
slaughter.

```{r}
#| label: meat-load
#| message: false
meat <- read_csv("https://dcgerard.github.io/stat_302/data/case0702.csv")
```

The relationship between time since slaughter and pH looks curved:

```{r}
#| label: meat-scatter
#| message: false
ggplot(meat, aes(x = Time, y = pH)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

The plot looks better with a log-transformation of $X$:

```{r}
#| label: meat-log-scatter
#| message: false
ggplot(meat, aes(x = Time, y = pH)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_log10()
```

We fit a regression with $\log(\text{Time})$ as the explanatory variable:

```{r}
#| label: meat-fit
lm(pH ~ I(log(Time)), data = meat) |>
  tidy(conf.int = TRUE)
```

- $\hat{\beta}_1 = -0.726$.
- Doubling the time after slaughter is associated with a decrease in mean pH
  of $0.726\log(2) = 0.503$.
- 95% CI for this decrease: $(0.646\log(2),\,0.805\log(2)) = (0.448,\,0.558)$.

## Log on $X$, Log on $Y$

**Model:** $\mu(\log(Y) \mid X) = \beta_0 + \beta_1\log(X)$

Simplifying:
$$\text{Median}(Y \mid X) = \exp(\beta_0)\,X^{\beta_1}$$

When $X$ is doubled:
$$\text{Median}(Y \mid 2X) = \text{Median}(Y \mid X) \cdot 2^{\beta_1}$$

- **Observational study:** $2^{\beta_1}$ is the ratio of medians when one
  group's $X$ is twice the other's.
- **Randomized experiment:** $2^{\beta_1}$ is the multiplicative change in median $Y$ when $X$ is doubled.

A similar interpretation holds for $10^{\beta_1}$ and a 10-fold increase in
$X$.

### Example: Island Area and Number of Species

Researchers measured the area of islands in the West Indies and counted the
number of amphibian species on each island.

```{r}
#| label: species-load
#| message: false
species <- read_csv("https://dcgerard.github.io/stat_302/data/case0801.csv")
```

```{r}
#| label: species-scatter
#| message: false
ggplot(species, aes(x = Area, y = Species)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)
```

A log-log relationship works best:

```{r}
#| label: species-log-scatter
#| message: false
ggplot(species, aes(x = Area, y = Species)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_log10() +
  scale_y_log10()
```

We fit the linear model:

```{r}
#| label: species-fit
lm(I(log(Species)) ~ I(log(Area)), data = species) |>
  tidy(conf.int = TRUE)
```

- The estimated relationship, for $S$ the number of species and $A$ the island
  area, is
  $$
  S = e^{1.93}A^{0.25}
  $$
- Islands with twice the area have $2^{0.25} = 1.19$ times as many species —
  19% more.
- The 95% confidence interval is $(2^{0.22}, 2^{0.28}) = (1.16, 1.22)$ — 16%
  to 22% more species.

## Other Transformations

Other transformations (e.g., $1/Y$, $\sqrt{Y}$) may improve residual plots but
lack clean interpretations. They are appropriate when the goal is:

1. **Prediction**, or
2. Only detecting **whether** an association exists (not describing its size).

- $1/Y$ tends to fix more extreme non-constant variance.
- $\sqrt{Y}$ tends to fix less extreme non-constant variance.

---

# Case Study: Voltage vs. Breakdown Time

A controlled experiment studied the relationship between voltage and the
breakdown time of an electrical insulating fluid. The researchers could control
the voltage level of each trial.

```{r}
#| label: voltage-data
#| message: false
case0802 <- read_csv("https://dcgerard.github.io/stat_302/data/case0802.csv")
glimpse(case0802)
```

## Step 1: Scatterplot

```{r}
#| label: voltage-scatter
#| message: false
ggplot(case0802, aes(x = Voltage, y = Time)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

## Step 2: Residual Plot

```{r}
#| label: voltage-resid
lmout <- lm(Time ~ Voltage, data = case0802)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

- As the mean increases, variability increases.
- The relationship is clearly curved.
- A log-transformation of $Y$ is needed.

## Step 3: Log-Transform and Re-evaluate

```{r}
#| label: voltage-log-scatter
#| message: false
case0802 <- case0802 |>
  mutate(logTime = log(Time))

ggplot(case0802, aes(x = Voltage, y = logTime)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(y = "log(Time)")
```

```{r}
#| label: voltage-log-resid
lmout <- lm(logTime ~ Voltage, data = case0802)
augment(lmout) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Fitted values", y = "Residuals")
```

After the log-transformation, the linearity and equal-variance assumptions look
satisfied.

## Step 4: Inference

```{r}
#| label: voltage-tidy
tidy(lmout)
```

```{r}
#| label: voltage-confint
tidy(lmout, conf.int = TRUE)
```

## Interpretation on the Original Scale

A one kV increase in voltage results in a multiplicative change of
$\exp(-0.507) = 0.60$ in median breakdown time — a 40% decrease.

```{r}
#| label: voltage-interpret
tidy(lmout, conf.int = TRUE) |>
  filter(term == "Voltage") |>
  select(estimate, conf.low, conf.high) |>
  mutate(across(everything(), exp))
```

```{r}
#| label: voltage-pct-decrease
tidy(lmout, conf.int = TRUE) |>
  filter(term == "Voltage") |>
  transmute(
    pct_change      = (1 - exp(estimate)) * 100,
    pct_change_low  = (1 - exp(conf.high)) * 100,
    pct_change_high = (1 - exp(conf.low)) * 100
  )
```

---

# Lack-of-Fit $F$-Test

## We Could Have Viewed This as an ANOVA Problem

The voltage variable only takes seven distinct values, so we could have
treated it as categorical and used ANOVA:

```{r}
#| label: lof-boxplot
case0802 <- case0802 |>
  mutate(VoltageFac = as.factor(Voltage))

ggplot(case0802, aes(x = VoltageFac, y = logTime)) +
  geom_boxplot() +
  labs(x = "Voltage", y = "log(Time)")
```

## Which Is Better?

- If the linear model appears to fit fine, it is **always preferred**.
- You can interpolate with the linear model (not with ANOVA).
- The linear model has easier interpretations.
- The linear model has fewer parameters.
- We can formally test if the linear model does not fit using the $F$-testing
  strategy if we have replicates at given values of $X$.

## Lack-of-Fit $F$-Test

- $H_0: E[Y_i] = \beta_0 + \beta_1X_i$ (the means lie on a line).
- $H_A: E[Y_i] = \mu_j$ where $X_i = j$ (each group has its own mean).

Under the **full model**, each voltage group gets its own estimated mean
(dashed red segments):

```{r}
#| label: lof-full-setup
#| message: false
#| echo: false
lmout <- lm(logTime ~ Voltage, data = case0802)

meandf <- case0802 |>
  group_by(VoltageFac) |>
  summarize(mean = mean(logTime)) |>
  mutate(z = as.numeric(VoltageFac)) |>
  mutate(zlower = z - 1/3, zupper = z + 1/3)

case0802 <- case0802 |>
  mutate(jitterx = jitter(as.numeric(VoltageFac)),
         fitted  = fitted(lmout)) |>
  left_join(select(meandf, VoltageFac, mean), by = "VoltageFac")
```

```{r}
#| label: lof-full-plot
#| echo: false
ggplot(case0802, aes(x = jitterx, y = logTime)) +
  geom_point() +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_segment(data = meandf,
               mapping = aes(x = zlower, xend = zupper, y = mean, yend = mean),
               col = 2, lty = 2) +
  labs(x = "Voltage")
```

Under the **reduced model**, the group means must lie on a line:

```{r}
#| label: lof-reduced-plot
#| echo: false
ggplot(case0802, aes(x = jitterx, y = logTime)) +
  geom_point() +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_line(mapping = aes(x = (Voltage - 24) / 2, y = fitted), lty = 2, col = 2) +
  labs(x = "Voltage")
```

Residuals under the full model:

```{r}
#| label: lof-resid-full
#| echo: false
ggplot(case0802, aes(x = jitterx, y = logTime)) +
  geom_point() +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_segment(data = meandf,
               mapping = aes(x = zlower, xend = zupper, y = mean, yend = mean),
               col = 2, lty = 2) +
  geom_segment(mapping = aes(x = jitterx, xend = jitterx, y = logTime, yend = mean),
               col = "blue") +
  labs(x = "Voltage")
```

Residuals under the reduced model:

```{r}
#| label: lof-resid-reduced
#| echo: false
ggplot(case0802, aes(x = jitterx, y = logTime)) +
  geom_point() +
  scale_x_continuous(breaks = 1:7, labels = levels(case0802$VoltageFac)) +
  geom_line(mapping = aes(x = (Voltage - 24) / 2, y = fitted), lty = 2, col = 2) +
  geom_segment(mapping = aes(x = jitterx, xend = jitterx, y = logTime, yend = fitted),
               col = "blue") +
  labs(x = "Voltage")
```

We calculate the lack-of-fit $F$-statistic using the extra sum of squares:

```{r}
#| label: lof-calc
#| message: false
case0802 <- read_csv("https://dcgerard.github.io/stat_302/data/case0802.csv")
case0802 |>
  mutate(logTime = log(Time)) ->
  case0802
lmout <- lm(logTime ~ Voltage, data = case0802) ## estimates linear model
case0802 |>
  group_by(Voltage) |>
  mutate(mean = mean(logTime)) |> ## full model means
  ungroup() |>
  mutate(fitted = fitted(lmout)) ->
  case0802
rss_full    <- sum((case0802$logTime - case0802$mean)^2)
rss_reduced <- sum((case0802$logTime - case0802$fitted)^2)
ESS         <- rss_reduced - rss_full
df_full     <- 69
df_reduced  <- 74
df_extra    <- df_reduced - df_full
fstat       <- (ESS / df_extra) / (rss_full / df_full)
```

- $RSS_{full}$ = `r round(rss_full, 3)`
- $df_{full} = n - I$ = 76 - 7 = 69
- $RSS_{reduced}$ = `r round(rss_reduced, 3)`
- $df_{reduced} = n - 2$ = 76 - 2 = 74
- $ESS = RSS_{reduced} - RSS_{full}$ = `r round(ESS, 3)`
- $df_{extra} = df_{reduced} - df_{full}$ = 74 - 69 = 5
- $F$-statistic $= \frac{ESS / df_{extra}}{RSS_{full} / df_{full}}$ = `r round(fstat, 4)`

We compare the $F$-statistic to an $F_{5,69}$ distribution:

```{r}
#| label: lof-fdist
#| echo: false
x <- seq(0, 5, length = 100)
y <- df(x, 5, 69)

which_x <- x > fstat
polydf <- data.frame(
  polyx = c(min(x[which_x]), x[which_x], max(x[which_x]), min(x[which_x])),
  polyy = c(0, y[which_x], 0, 0)
)
ggplot(data.frame(x = x, y = y), aes(x = x, y = y)) +
  geom_line() +
  geom_polygon(data = polydf, aes(x = polyx, y = polyy), fill = "blue", alpha = 3/4) +
  labs(x = "F-stat", y = "Density")
```

We calculate the $p$-value as the area to the right of the $F$-statistic:

```{r}
#| label: lof-pvalue
pf(fstat, df1 = 5, df2 = 69, lower.tail = FALSE)
```

## Lack-of-Fit Test in R

Create a factor variable:

```{r}
#| label: lof-factor
case0802 <- case0802 |>
  mutate(VoltageFac = as.factor(Voltage))
```

Fit both the ANOVA and regression models:

```{r}
#| label: lof-models
aout  <- aov(logTime ~ VoltageFac, data = case0802)
lmout <- lm(logTime ~ Voltage,    data = case0802)
```

Use `anova()` to get the ANOVA table:

```{r}
#| label: lof-anova
anova(lmout, aout)
```
