---
title: "Interpreting Log Transformations"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = FALSE,
                      fig.height = 2.3, 
                      fig.width = 3.7, 
                      fig.align = "center")
```


```{r}
set.seed(1)
library(ggplot2)
```

## Objectives

- Interpret Log-transformations of either the explanatory of response variable.
- Ch 8 in the book.

# No log on $X$, no log on $Y$

## No log on $X$, no log on $Y$: Interpretation

- Model: $\mu(Y|X) = \beta_0 + \beta_1X$\pause

- So at $X+1$ the mean is
\begin{align*}
\mu(Y|X+1) &= \beta_0 + \beta_1(X + 1)\\ 
&= \beta_0 + \beta_1X + \beta_1\\
&= \mu(Y|X) + \beta_1
\end{align*}

## No log on $X$, no log on $Y$: Interpretation

- Observational study interpretation: $\beta_1$ is the mean difference in the $Y$'s when the $X$'s are only one unit apart.

- Randomized experiment interptation: $\beta_1$ is the increase in $Y$ when $X$ is increased by one unit.

# No log on $X$, log on $Y$

## No log on $X$, log on $Y$: Interpretation

- Model: $\mu(\log(Y)|X) = \beta_0 + \beta_1 X$\pause

- If *residuals* are fairly symmetric, this means
\begin{align*}
Median(\log(Y)|X) &= \beta_0 + \beta_1 X\\
\Rightarrow Median(Y|X) &= \exp(\beta_0 + \beta_1 X)\\
&= \exp(\beta_0) \exp(\beta_1 X)
\end{align*} \pause

- So at $X+1$ we have
\begin{align*}
Median(Y|X + 1) &= \exp(\beta_0) \exp(\beta_1 (X + 1))\\
&= \exp(\beta_0) \exp(\beta_1X + \beta_1)\\
&= \exp(\beta_0) \exp(\beta_1X) \exp(\beta_1)\\
&= Median(Y|X) \exp(\beta_1)\\
\end{align*}

## No log on $X$, log on $Y$: Interpretation

- Observational study interpretation: $\exp(\beta_1)$ is the ratio of medians of the $Y$'s when they differ only by one unit of the $X$'s.

- Randomized experiment interptation: $\exp(\beta_1)$ is the multiplicative change in $Y$ when $X$ is increased by one unit.


## No log on $X$, log on $Y$: Example

- Voltage and Breakdown: $\hat{\beta}_1 = -0.51$, 95% CI (-0.62, -0.39).

- "Increasing voltage of 1 kV results in a multiplicative change of $\exp(-0.51) = 0.6$""

- "Breakdown time at 28 kV is 60% that of 27 kV"

- 95% confidence interval of the multiplicative effect is $(\exp(-0.62), \exp(-0.39)) = (0.54, 0.68)$.

## No log on $X$, log on $Y$: Example

\begin{align*}
&Median(Y|X + 1)  - Median(Y|X) \\
&= Median(Y|X) \exp(\beta_1) - Median(Y|X)\\
&= (\exp(\beta_1) - 1)Median(Y|X)\\
&= (0.6 - 1)Median(Y|X)\\
&= -0.4Median(Y|X)
\end{align*}

- "Increasing kV by 1 decreases median breakdown time by 40%."

# log on $X$, no log on $Y$

## log on $X$, no log on $Y$: Interpretation

- Model $\mu(Y|X) = \beta_0 + \beta_1\log(X)$\pause

- Observe what happens to $X$ when we multiply it by 2.

\begin{align*}
\mu(Y|2X) &= \beta_0 + \beta_1\log(2X)\\
&= \beta_0 + \beta_1(\log(X) + \log(2))\\
&= \beta_0 + \beta_1\log(X) + \beta_1\log(2)\\
&= \mu(Y|X) + \beta_1\log(2)
\end{align*}

## log on $X$, no log on $Y$: Interpretation

- Observational study interpretation: $\beta_1\log(2)$ is the mean difference in the $Y$'s when the ratio of the $X$'s is 2.

- Randomized experiment interptation: $\beta_1\log(2)$ is the increase in $Y$ when $X$ is doubled.

- Similar interpretation for $\beta_1\log(10)$ and a 10-fold increase in $X$.

## log on $X$, no log on $Y$: Example

- pH and time in meat after slaughter

- $Y_i = \beta_0 + \beta_1 \log(X_i)$

- $\hat{\beta}_1 = -0.726$, 95% CI (-0.805, -0.646).

- If we double the time, then the $pH$ decreases by $0.726\log(2) = 0.503$

- 95% CI: $(-0.805\log(2), -0.646\log(2)) = (-0.558, -0.448)$.

# log on $X$, log on $Y$

## log on $X$, log on $Y$: Interpretation

- Combination of the past two interpretations\pause

- Simplifying the model:
\begin{align*}
Median(Y|X) &= \exp(\beta_0)\exp(\beta_1\log(X))\\
&= \exp(\beta_0)\exp(\log(X^{\beta_1}))\\
&= \exp(\beta_0)X^{\beta_1}\\
\end{align*}\pause

- See how $Median(Y|X)$ when you double $X$
\begin{align*}
Median(Y|2X) &= \exp(\beta_0)(2X)^{\beta_1}\\
&=\exp(\beta_0)X^{\beta_1}2^{\beta_1}\\
&=Median(Y|X)2^{\beta_1}
\end{align*}

## log on $X$, log on $Y$: Interpretation

- Observational study interpretation: $2^{\beta_1}$ is the ratio of medians of the $Y$'s when the ratio of the $X$'s is 2.

- Randomized experiment interptation: $2^{\beta_1}$ is the multiplicative change in $Y$ when $X$ is doubled.

- Similar interpretation for $10^{\beta_1}$ and a 10-fold increase in $X$.

## Other Transformations

- Might make residuals look better but don't have nice interpretations.

- If goal is 

    1. Prediction
    2. Just answering **if** there is an association and you don't care what it is.
    
    Then you can try other transformations (because interpretation does not matter).

- $1/Y$ tends to fix more extreme non-constant variance.

- $\sqrt{Y}$ tends to fix less extreme non-constant variance.
