---
title: "Linear Model Assumptions Worksheet"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
knitr::opts_chunk$set(echo       = TRUE,
                      eval       = FALSE,
                      fig.height = 3,
                      fig.width  = 4,
                      fig.align  = "center")
```

# Wine Consumption and Heart Disease

Data frame `ex0823` from the `Sleuth3` package contains the average wine
consumption rates (in liters per person) and number of ischemic heart disease
deaths (per 1,000 men aged 55 to 64 years old) for 18 industrial countries. Do
these data suggest that the heart disease death rate is associated with average
wine consumption? If so, how can that relationship be described? Do any
countries have substantially higher or lower death rates than others with
similar wine consumption rates? We will answer these questions.

Load the data with:

```{r}
#| label: load-data
#| eval: true
#| message: false
library(tidyverse)
ex0823 <- read_csv("https://dcgerard.github.io/stat_302/data/ex0823.csv")
glimpse(ex0823)
```

**1.** What is the explanatory variable? What is the response variable? What
are the observational units? Is this an observational study or a randomized
experiment?

**2.** Make a scatterplot of `Wine` (x-axis) vs `Mortality` (y-axis). Describe
the relationship — does it look linear? Comment on the direction and strength
of any trend you see.

```{r}

```

**3.** Based on the scatterplot, which of the four linear model assumptions
(linearity, independence, equal variance, normality) appear potentially
violated? Explain briefly.

**4.** Fit an initial simple linear regression of `Mortality` on `Wine` using
`lm()`. Then create a residual plot (fitted values on the $x$-axis, residuals
on the $y$-axis) using `augment()` from the `broom` package.

```{r}

```

**5.** Examine the residual plot. What patterns, if any, do you see? Which
assumption(s) does this suggest are violated? What transformation(s) might
address these issues?

**6.** Try the following transformations one at a time and evaluate each with
a scatterplot and residual plot:

  a. Log-transform `Mortality` only: fit `log(Mortality) ~ Wine`.
  b. Log-transform `Wine` only: fit `Mortality ~ log(Wine)`.
  c. Log-transform both variables: fit `log(Mortality) ~ log(Wine)`.

For each, note whether the residual plot looks better or worse than the
original. Which combination, if any, satisfies the linearity and equal-variance
assumptions?

```{r}

```

**7.** Using your final model from Question 6, report the estimated slope
coefficient. Then compute the appropriate back-transformed estimate and 95%
confidence interval. Write a one-sentence plain-language interpretation of the
slope on the original scale, in the context of this study.

```{r}

```

**8.** Is it appropriate to use causal language such as "drinking more wine
*causes* lower heart disease mortality"? Explain your reasoning based on the
study design.
