---
title: "Random Samples"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(echo = TRUE, fig.height = 2.3, fig.width = 3, fig.align = "center")
```

## Learning Objectives

- Gain intuitive understanding of CLT

## Birthweight Data

```{r}
library(ggplot2)
weight <- read.csv("../data/BIRTHWEIGHT.csv")$weight
qplot(x = weight, geom = "histogram", fill = I("white"), 
      color = I("black"), bins = 30)
```


## Get Sampling Distributions of Size 5

```{r}
xbarvec5 <- replicate(n = 10000,
                      expr = {
                        mean(sample(weight, size = 5, 
                                    replace = TRUE))
                      })
```


## Get Sampling Distributions of Size 10

```{r}
xbarvec10 <- replicate(n = 10000,
                       expr = {
                         mean(sample(weight, size = 10, 
                                     replace = TRUE))
                       })
```


## Get Sampling Distributions of Size 50

```{r}
xbarvec50 <- replicate(n = 10000,
                       expr = {
                         mean(sample(weight, size = 20, 
                                     replace = TRUE))
                       })
```

## Get Sampling Distributions of Size 100

```{r}
xbarvec100 <- replicate(n = 10000,
                        expr = {
                          mean(sample(weight, size = 100, 
                                      replace = TRUE))
                        })
```

## Get Sampling Distributions of Size 1000

```{r}
xbarvec1000 <- replicate(n = 10000,
                        expr = {
                          mean(sample(weight, size = 1000,
                                      replace = TRUE))
                        })
```

## Variances Decrease over Time

```{r, echo = FALSE}
xbardf <- data.frame(xbar = c(xbarvec5, xbarvec10, xbarvec50, xbarvec100, xbarvec1000),
                     n    = c(rep(5, length(xbarvec5)), rep(10, length(xbarvec10)), 
                              rep(50, length(xbarvec50)), rep(100, length(xbarvec100)),
                              rep(1000, length(xbarvec1000))))
ggplot(data = xbardf, mapping = aes(x = as.factor(n), y = xbar)) +
  geom_boxplot() +
  xlab("n")
```


## Not Very Normal (n = 5)

```{r, echo = FALSE}
qplot(x = xbarvec5, y = ..density.., geom = "histogram", 
      fill = I("white"), color = I("black"), bins = 30) +
  geom_density()
```


## More Normal (n = 10)

```{r, echo = FALSE}
qplot(x = xbarvec10, y = ..density.., geom = "histogram", 
      fill = I("white"), color = I("black"), bins = 30) +
  geom_density()
```

## More Normal (n = 50)

```{r, echo = FALSE}
qplot(x = xbarvec50, y = ..density.., geom = "histogram", 
      fill = I("white"), color = I("black"), bins = 30) +
  geom_density()
```

## Very Normal (n = 100)

```{r, echo = FALSE}
qplot(x = xbarvec100, y = ..density.., geom = "histogram", 
      fill = I("white"), color = I("black"), bins = 30) +
  geom_density()
```


##  Oh-my-god-can-it-be-any-more Normal (n = 1000)

```{r, echo = FALSE}
qplot(x = xbarvec1000, y = ..density.., geom = "histogram", 
      fill = I("white"), color = I("black"), bins = 30) +
  geom_density()
```



## Not Very Normal (n = 5)

```{r, echo = FALSE}
qplot(sample = xbarvec5, geom = "qq") +
  geom_qq_line()
```


## More Normal (n = 10)

```{r, echo = FALSE}
qplot(sample = xbarvec10, geom = "qq") +
  geom_qq_line()
```



## More Normal (n = 50)

```{r, echo = FALSE}
qplot(sample = xbarvec50, geom = "qq") +
  geom_qq_line()
```

## More Normal (n = 100)

```{r, echo = FALSE}
qplot(sample = xbarvec100, geom = "qq") +
  geom_qq_line()
```


##  Oh-my-god-can-it-be-any-more Normal (n = 1000)


```{r, echo = FALSE}
qplot(sample = xbarvec1000, geom = "qq") +
  geom_qq_line()
```



