---
title: "Assumptions of t-Tools"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r setup}
#| include: false
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- List the assumptions of the two-sample $t$-test in order of importance.
- Detect and address violations of independence.
- Detect and address violations of the equal-variance assumption.
- Detect and address non-normality (skew and outliers).
- Apply log transformations and interpret results on the original scale.

# Assumptions of 2-Sample $t$-Tools

- Assumptions in decreasing order of importance:
  1. Independence between observational units.
  2. Equal variance (if assuming $\sigma_1^2 = \sigma_2^2$).
  3. Normality (no severe skew or outliers).

- Each assumption can be checked, and solutions exist for each violation.

# 1. Independence

- **Definition**: Knowing the value of one observation does not give you
  any information about the value of another observation.

- **Example of violation — cluster effects**:
  - Measuring gene expression levels in two groups (disease vs. control).
  - Samples collected by two different technicians.
  - Technician 1 tended to produce samples with systematically higher
    levels.
  - Observations within a technician are correlated.

- **Example of violation — serial effects**:
  - Measuring expression levels on the *same* individuals over time.
  - Observations on the same person are correlated across time.

![Serial effects: measurements within each individual (dashed lines) are correlated across time, violating independence.](02_fig/fig_serial_effects.svg){fig-align="center" width="85%"}

## To Detect

1. Think carefully about how the data were collected.
   - Were different responses measured on the same subject?
   - Were data collected in groups?
   - Were groups treated differently in ways unrelated to treatment?
2. Residual plots (covered in Chapter 8).

## Issues When Independence Is Violated

- Recall $\bar{X} - \bar{Y} \sim N(\mu_1 - \mu_2,\; \sigma_1^2/n_1 + \sigma_2^2/n_2)$.
- If samples are not independent, the variance formula above is *too small*.
- You have less information than you think.
- Consequently: p-values are too small and confidence intervals are too narrow.

- **Extreme example**:
  - You have 4 people. Measure each one 50 times.
  - Group 1: Person 1 always scores 0; Person 2 always scores 1.
  - Group 2: Person 3 always scores 2; Person 4 always scores 3.
  - Nominal $n = 100$ per group, but effective sample size is 2 per group.
  - Correct variance estimate: $\sigma^2 \approx 0.25 / 2 = 0.125$.
  - Incorrectly assumed variance: $\sigma^2 \approx 0.25 / 100 = 0.0025$.

## Solutions

- ANOVA with blocking (Chapters 12–13) for cluster effects.
- Longitudinal analysis (Chapter 18) for serial effects.

# 2. Unequal Variances

- Sometimes people assume $\sigma_1^2 = \sigma_2^2$ (the *pooled*
  variance $t$-test), because it generalizes more easily to more complex
  methods.

- Under the equal-variance assumption:
  $$\bar{X} - \bar{Y} \;\sim\; N\!\left(\mu_1 - \mu_2,\; \sigma^2\!\left(\frac{1}{n_1} + \frac{1}{n_2}\right)\right)$$
  Estimate $\sigma^2$ with the pooled variance:
  $$s_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}$$

## To Detect

- Make side-by-side boxplots; only worry if the spread looks very different.

```{r unequal-var-example}
#| echo: false
library(tidyverse)
set.seed(42)
tibble(
  group = rep(c("Group 1", "Group 2"), each = 50),
  y     = c(rnorm(50, mean = 5, sd = 1), rnorm(50, mean = 5, sd = 4))
) |>
  ggplot(aes(x = group, y = y)) +
  geom_boxplot(coef = Inf) +
  labs(x = NULL, y = "Response", title = "Unequal spread across groups")
```

## Issues

- If $\sigma_1^2 \neq \sigma_2^2$ but we assume they are equal, the
  variance estimate of $\bar{X} - \bar{Y}$ is wrong.
- This is mainly a problem when $n_1$ is very different from $n_2$.

## Solution

- Assume unequal variances — this is the default in `t.test()` in R,
  which uses the Satterthwaite approximation for degrees of freedom.
- Alternatively, transform the data.

# 3. Normality (Skew)

- The $t$-test assumes that $X_1, \ldots, X_{n_1}$ and
  $Y_1, \ldots, Y_{n_2}$ are approximately normal, or that the sample
  size is "large enough" (Central Limit Theorem).
- More skew $\Rightarrow$ larger sample size needed.

## Issues

- Confidence intervals do not achieve 95% coverage.
- p-values do not have the correct null interpretation.

## To Detect

- Make quantile-quantile (QQ) plots.

## Solutions

- Apply a transformation (see Section 5 below).
- Perform a permutation test.

# 4. Normality (Outliers)

## To Detect

- Histograms, boxplots, QQ-plots.

## Issues

- Results are unstable — removing a single point can change conclusions.

## Solution

- Run the analysis with and without the outlier.
- If answers are the same, do not remove it.
- If answers differ, either:
  1. Use a robust analysis (Chapter 4).
  2. Report both analyses.

# 5. Transformations

- Many transformations are possible: $\sqrt{y}$, $\arcsin(\sqrt{y})$,
  $\log(y)$.
- **Log is usually the only one used in practice.**

- **When to use log**:
  1. Values are positive.
  2. Larger mean $\Rightarrow$ larger variance.
  3. Data are skewed.
  - Log makes such data symmetric and stabilizes variance.

```{r rainfall-plots}
#| message: false
library(Sleuth3)
library(patchwork)

data("case0301")
p_raw <- ggplot(case0301, aes(x = Treatment, y = Rainfall)) +
  geom_boxplot() +
  labs(title = "Raw scale")

p_log <- ggplot(case0301, aes(x = Treatment, y = Rainfall)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Log scale")

p_raw + p_log
```

## Interpreting Log Transformations

- Let $W_i = \log(Y_i)$ and $Z_j = \log(X_j)$ in the two-sample model:
  $$W_i = \mu + \varepsilon_i \qquad Z_j = \mu + \delta + \xi_j$$

- Then:
  $$\delta = \text{Average}(Z) - \text{Average}(W)
           = \text{Average}(\log X) - \text{Average}(\log Y)$$

- We want to interpret $\delta$ on the original $X, Y$ scale.

- **Important caveat**: $\log(\text{Average}(X)) \neq \text{Average}(\log X)$.
  - Example: $X = 2, 4$.
  - $\log_2 X = 1, 2 \Rightarrow \text{Average}(\log_2 X) = 1.5$.
  - $\text{Average}(X) = 3 \Rightarrow \log_2(3) \approx 1.58$.
  - So $E[X] \neq e^\delta \cdot E[Y]$ in general.

- **What is true**: if $\log(X)$ and $\log(Y)$ are both symmetric,
  $$e^\delta = \frac{\text{Median}(X)}{\text{Median}(Y)}$$
  - $e^\delta$ is the **multiplicative difference in medians**.

- **Procedure**:
  1. Log-transform the data.
  2. Run `t.test()` on the log-transformed values to get $\hat{\delta}$
     and a CI $(\hat{\delta}_L, \hat{\delta}_U)$.
  3. Exponentiate: $e^{\hat{\delta}}$ estimates the ratio of medians;
     $(e^{\hat{\delta}_L}, e^{\hat{\delta}_U})$ is the CI for the ratio.

```{r rainfall-log-ttest}
#| message: false
library(broom)

case0301 |>
  mutate(logRainfall = log(Rainfall)) |>
  t.test(logRainfall ~ Treatment, data = _) |>
  tidy() ->
  tout

tout |> select(estimate, p.value, conf.low, conf.high)

## Estimated ratio of median rainfall (Seeded / Unseeded):
exp(tout$estimate)

## 95% CI for the ratio of medians:
exp(c(lower = tout$conf.low, upper = tout$conf.high))
```

- Estimated median rainfall in the seeded group is about
  `r round(exp(tout$estimate), 1)` times that of the unseeded group,
  with a 95% confidence interval of
  (`r round(exp(tout$conf.low), 1)`, `r round(exp(tout$conf.high), 1)`).
