---
title: "Review: Logarithms and Exponentials"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
theme_set(theme_bw(base_size = 12))
knitr::opts_chunk$set(
  echo       = FALSE,
  fig.height = 3,
  fig.width  = 5,
  fig.align  = "center"
)
```

# Why Logarithms?

In statistics, we sometimes transform data before analysis. The **log
transformation** is the most common transformation used with $t$-tools because
it can simultaneously:

- Reduce right skew, making the data more symmetric.
- Stabilize variance across groups.
- Allow us to interpret results as *ratios* rather than *differences*.

This review covers the mathematics behind the logarithm and exponential
functions so that you can apply and interpret log transformations confidently.

---

# Euler's Number $e$

$$e = 2.71828\ldots$$

$e$ is an irrational constant (like $\pi$) that appears throughout mathematics,
probability, and statistics. It is called **Euler's number**.

Integer powers of $e$ work just like any other base:

$$e^1 = e \approx 2.718 \qquad e^2 = e \cdot e \approx 7.389 \qquad e^3 = e \cdot e \cdot e \approx 20.086$$

Non-integer powers are also defined. For example:

$$e^{2.5} = e^{5/2} = \left(\sqrt{e}\right)^5 \approx 12.182$$

In R, compute powers of $e$ with `exp()`:

```{r}
#| echo: true
exp(1)   # e^1
exp(2)   # e^2
exp(2.5) # e^2.5
```

---

# The Natural Logarithm

The **natural logarithm** $\log(x)$ answers the question:

> *To what power must I raise $e$ to get $x$?*

Formally, $\log(x) = a$ means $e^a = x$.

::: {.callout-tip}
In statistics, "$\log$" always means the **natural logarithm** (base $e$),
not $\log_{10}$. In R, `log(x)` is the natural log.
:::

Some examples:

| Expression | Value | Because |
|:-----------|------:|:--------|
| $\log(e)$ | $1$ | $e^1 = e$ |
| $\log(e^3)$ | $3$ | $e^3 = e^3$ |
| $\log(e^{2.5})$ | $2.5$ | $e^{2.5} = e^{2.5}$ |
| $\log(1)$ | $0$ | $e^0 = 1$ |
| $\exp(0)$ | $1$ | same as above |

**Domain**: $\log(x)$ is only defined for $x > 0$. This is why we can only
log-transform positive data.

```{r}
#| echo: true
log(exp(3))   # log and exp are inverses: gives 3
exp(log(5))   # same idea: gives 5
log(1)        # log(1) = 0
```

## Visualizing $\exp$ and $\log$

The functions $y = e^x$ and $y = \log(x)$ are **inverses** of each other —
they are mirror images across the line $y = x$:

```{r}
#| label: exp-log-plot
x_exp <- seq(-2, 2.5, length.out = 300)
x_log <- seq(0.01, 12, length.out = 300)
df_exp <- data.frame(x = x_exp, y = exp(x_exp), fn = "exp(x)")
df_log <- data.frame(x = x_log, y = log(x_log), fn = "log(x)")
df_id  <- data.frame(x = seq(-2, 12, length.out = 300),
                     y = seq(-2, 12, length.out = 300), fn = "y = x")
df_all <- rbind(df_exp, df_log, df_id)

ggplot(df_all, aes(x = x, y = y, color = fn, linetype = fn)) +
  geom_line(linewidth = 0.8) +
  coord_cartesian(xlim = c(-2.5, 10), ylim = c(-2.5, 10)) +
  scale_color_manual(values = c("steelblue", "tomato", "gray50")) +
  scale_linetype_manual(values = c("solid", "solid", "dashed")) +
  labs(x = "x", y = "y", color = NULL, linetype = NULL) +
  theme(legend.position = "bottom")
```

Notice:

- $e^x$ is always **positive** and grows rapidly.
- $\log(x)$ is only defined for $x > 0$ and grows slowly.
- The log function **stretches** small values and **compresses** large values —
  exactly what we want when data are right-skewed.

---

# Key Properties

These properties follow directly from the definition and are used constantly
when working with log transformations.

## Log properties

$$\log(ab) = \log(a) + \log(b)$$
$$\log\!\left(\frac{a}{b}\right) = \log(a) - \log(b)$$
$$\log(a^r) = r\,\log(a)$$

## Exp properties

$$\exp(a + b) = \exp(a)\,\exp(b)$$
$$\exp(a - b) = \frac{\exp(a)}{\exp(b)}$$

## Inverse properties

$$\log(\exp(a)) = a \qquad \exp(\log(a)) = a$$

These last two are the most important: **log and exp undo each other**.

## Worked examples

::: {.panel-tabset}
### Example 1
Simplify $\log(3 \cdot 7)$.

### Solution
$$\log(3 \cdot 7) = \log(3) + \log(7) \approx 1.099 + 1.946 = 3.045$$

Check: $\log(21) \approx 3.045$. $\checkmark$
:::

::: {.panel-tabset}
### Example 2
Simplify $\log(20/4)$.

### Solution
$$\log(20/4) = \log(20) - \log(4) \approx 2.996 - 1.386 = 1.609$$

Check: $\log(5) \approx 1.609$. $\checkmark$
:::

::: {.panel-tabset}
### Example 3
Simplify $\exp(\log(7) - \log(3))$.

### Solution
$$\exp(\log(7) - \log(3)) = \exp\!\left(\log\!\left(\frac{7}{3}\right)\right) = \frac{7}{3} \approx 2.333$$

The key step: $\exp(\log(\cdot)) = \cdot$.
:::

::: {.panel-tabset}
### Example 4
If $\log(x) - \log(y) = 1.4$, what is $x/y$?

### Solution
$$\log(x) - \log(y) = \log\!\left(\frac{x}{y}\right) = 1.4$$

$$\frac{x}{y} = e^{1.4} \approx 4.055$$

So $x$ is about 4 times larger than $y$.
:::

---

# Other Bases

Logarithms can be defined for any base $b > 0$.

$$\log_b(x) = a \quad \text{means} \quad b^a = x$$

The **base-2 logarithm** $\log_2(x)$ counts how many times you must multiply
2 by itself to get $x$. For example:

$$2 \times 2 \times 2 = 8 \quad \Rightarrow \quad \log_2(8) = 3$$

Equivalently, it counts how many times you must divide $x$ by 2 to reach 1:

$$8 \xrightarrow{\div 2} 4 \xrightarrow{\div 2} 2 \xrightarrow{\div 2} 1 \quad \Rightarrow \quad \log_2(8) = 3$$

In R, use `log(x, base = 2)` or `log2(x)` for base-2, and `log10(x)` for
base-10. **In this course, we always use the natural log** `log(x)`.

---

# Connection to Log Transformations in Statistics

Here is why the log properties matter for interpreting the two-sample $t$-test
on log-transformed data.

Suppose we log-transform two groups:
$$W_i = \log(Y_i) \qquad Z_j = \log(X_j)$$

The $t$-test estimates the difference in means on the log scale:
$$\hat{\delta} = \overline{Z} - \overline{W}$$

If $\log(X)$ and $\log(Y)$ are both symmetric, then
$\text{Average}(\log X) = \log(\text{Median}(X))$, so:

$$\hat{\delta} \approx \log(\text{Median}(X)) - \log(\text{Median}(Y))
  = \log\!\left(\frac{\text{Median}(X)}{\text{Median}(Y)}\right)$$

Exponentiating both sides:
$$e^{\hat{\delta}} = \frac{\text{Median}(X)}{\text{Median}(Y)}$$

**So $e^{\hat{\delta}}$ estimates the ratio of medians** — a multiplicative
comparison on the original scale.

::: {.panel-tabset}
### Example 5
A $t$-test on log-transformed data gives $\hat{\delta} = 0.916$. Interpret
this on the original scale.

### Solution
$$e^{0.916} \approx 2.5$$

Group $X$ has a median approximately **2.5 times larger** than group $Y$.
:::

::: {.panel-tabset}
### Example 6
A 95% CI for $\delta$ is $(0.2,\; 0.9)$. Give a 95% CI on the original
scale.

### Solution
Exponentiate the endpoints:
$$\left(e^{0.2},\; e^{0.9}\right) \approx (1.22,\; 2.46)$$

We are 95% confident that the median of group $X$ is between **1.22 and 2.46
times** the median of group $Y$.
:::

---

# Exercises

**1.** Without a calculator, evaluate the following. Explain your reasoning.

  a. $\log(e^5)$
  b. $\exp(\log(12))$
  c. $\log(1)$
  d. $e^0$

**2.** Use the log properties to simplify each expression to a single log or a
number. Then verify your answer in R with `log()`.

  a. $\log(6) + \log(2)$
  b. $\log(50) - \log(5)$
  c. $3\,\log(4)$
  d. $\log(e^2 \cdot e^3)$

**3.** Solve for $x$.

  a. $\log(x) = 3$
  b. $e^x = 15$
  c. $\log(x) - \log(4) = 2$
  d. $\exp(2x - 1) = 10$

**4.** Use R to fill in the table below. What pattern do you notice as $x$
increases? What about as $x$ decreases toward 0?

```{r}
#| echo: true
#| eval: false
x_vals <- c(0.01, 0.1, 1, 2, 5, 10, 100)
log(x_vals)
```

**5.** Consider two groups with medians 12 and 4 on the original scale.

  a. Compute the log of each median.
  b. Compute the difference of the log-medians (group 1 minus group 2).
  c. Exponentiate the result from (b). What does this number represent?
  d. Verify algebraically that $e^{\log(12) - \log(4)} = 12/4 = 3$.

**6.** A two-sample $t$-test on log-transformed data yields the following 95%
CI for $\delta$ (log-scale difference, Treatment minus Control): $(0.40,\; 1.25)$.

  a. Exponentiate both endpoints. What does this CI say about the medians on
     the original scale?
  b. Does the interval suggest a statistically significant difference? How can
     you tell without knowing the $p$-value?
  c. A classmate reports the result as "the treatment increases the mean by a
     factor of $e^{0.40}$ to $e^{1.25}$." What is wrong with this statement?

**7.** The log transformation is appropriate when three conditions hold. For
each dataset described below, decide whether a log transformation is likely
appropriate. Justify your answer.

  a. Annual household incomes in the United States (skewed right; range:
     \$0–\$10 million).
  b. IQ scores (approximately symmetric; range: 50–150).
  c. Bacterial colony counts in petri dishes (values from 1 to 10,000;
     larger counts tend to have higher variability).
  d. Temperatures (in Celsius) recorded daily in Chicago (range: $-20$ to
     $35$°C).
