---
title: "Checking For Normality"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 2.7, 
                      fig.width = 4, 
                      fig.align = "center")
```

# Checking for normality

## Some distributions look approximately normal
```{r, fig.height=2, fig.width=3}
library(Sleuth3)
library(ggplot2)
data("case0201")
beak78 <- case0201$Depth[case0201$Year == 1978]
qplot(beak78, bins = 15, color = I("black"), fill = I("white"))
```



## Clearly not all distributions are normal

```{r, fig.height=2, fig.width=3}
data("case0301") ## Rainfall data
rainun <- case0301$Rainfall[case0301$Treatment == "Unseeded"]
qplot(rainun, bins = 10, color = I("black"), 
      fill = I("white"), xlab = "rainfall")
```

## It's sometimes important to check if normality is a valid approximation.

1. Idea: Is the 68-95-99.7 rule approximately correct for a given dataset?
2. More generally, do the percentiles (quantiles) of the data match with the percentiles (quantiles) of the theoretical normal distribution?
3. Compare the $p$th percentile (quantile) of the data and the $p$th percentile (quantile) of a $N(\bar{x}, s^2)$ distribution. If they are pretty close, then normality is a good approximation.


## Look at percentiles (quantiles)

```{r}
mu    <- mean(beak78)
sigma <- sd(beak78)
qnorm(p = 0.2, mean = mu, sd = sigma)
quantile(x = beak78, probs = 0.2)
```

That matches almost exactly, what about other percentiles (quantiles)?

## More quantiles

```{r}
qnorm(p = 0.7, mean = mu, sd = sigma)
quantile(x = beak78, probs = 0.7)
```

## More quantiles

```{r}
qnorm(p = 0.9, mean = mu, sd = sigma)
quantile(x = beak78, probs = 0.9)
```
These are all pretty close!


## Quantile-quantile plot

1. Plots the observed quantiles against the quantiles of a $N(\bar{x}, s^2)$ density.
2. If the points lie close to a line, then the normal approximation is approximately correct.
3. Can just plot the observed quantiles against $N(0, 1)$ and look for a straight line (more on why later).


## QQplot
```{r}
qplot(sample = beak78, geom = "qq") +
    geom_qq_line()
```

## But what does a "good" qqplot look like?
```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
dftot <- data_frame(x = sort(beak78, decreasing = FALSE))
dftot$sim <- 0
for (index in 1:5) {
  x <- rnorm(length(beak78), mean = mu, sd = sigma)
  df_temp <- data_frame(x = sort(x, decreasing = FALSE))
  df_temp$sim <- index
  dftot <- bind_rows(dftot, df_temp)
}
dftot$sim <- as.factor(dftot$sim)
ggplot(data = dftot, mapping = aes(sample = x)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~sim) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"))
```
Top left is real data, rest are simulated from $N(\bar{x}, s^2)$ --- maybe a little non-normal?


## Problem: Skewed right
```{r, echo=FALSE, fig.width=2, fig.height=1.5}
x <- rgamma(200, 2, 2)
qplot(x, geom = "histogram", xlab = "", bins = 20)
qplot(sample = x, geom = "qq") +
    geom_qq_line()
```

## Problem: Skewed left
```{r, echo=FALSE, fig.width=2, fig.height=1.5}
op <- par(mfrow = c(1,2))
qplot(-x, xlab = "", bins = 20)
qplot(sample = -x, geom = "qq") +
    geom_qq_line()
```


## Problem: Outliers
```{r, echo=FALSE, fig.width=2, fig.height=1.5}
x <- c(rnorm(200), 4.5)
qplot(x, xlab = "", bins = 20)
qplot(sample = x, geom = "qq") +
    geom_qq_line()
```

## Problem: Heavy tails
```{r, echo=FALSE, fig.width=2, fig.height=1.5}
x <- rt(200, 3)
qplot(x, xlab = "", bins = 20)
qplot(sample = x, geom = "qq") +
    geom_qq_line()
```


## Problem: Light tails
```{r, echo=FALSE, fig.width=2, fig.height=1.5}
x <- runif(200)
qplot(x, xlab = "", bins = 20)
qplot(sample = x, geom = "qq") +
    geom_qq_line()
```


## The rainfall data

```{r}
qplot(sample = rainun, geom = "qq") +
  geom_qq_line()
```

