---
title: "Rainfall Example"
author: "David Gerard"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "metropolis"
    slide_level: 2
urlcolor: "blue"
---

```{r setup, include=FALSE}
ggplot2::theme_set(
  ggplot2::theme_bw()
  )
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 2.5, 
                      fig.width = 4, 
                      fig.align = "center")
```

## Objectives

Here, we work through the rainfall analysis

## Load in Data
```{r}
library(Sleuth3)
library(ggplot2)
data("case0301")
```


## Rainfall EDA
```{r}
qplot(x = Treatment, y = Rainfall, 
      data = case0301, geom = "boxplot")
```

## Rainfall EDA
```{r}
qplot(x = Rainfall, facets = . ~ Treatment,
      data = case0301, geom = "histogram", bins = 10)
```

## Rainfall EDA
```{r}
qplot(sample = Rainfall, facets = . ~ Treatment,
      data = case0301, geom = "qq") +
  geom_qq_line()
```

## Apply Transformation

```{r}
case0301$logRainfall <- log(case0301$Rainfall)
```

## Rain EDA

```{r}
qplot(x = Treatment, y = logRainfall, 
      data = case0301, geom = "boxplot")
```

## Rainfall EDA
```{r}
qplot(x = logRainfall, facets = . ~ Treatment,
      data = case0301, geom = "histogram", bins = 10)
```

## Rainfall EDA
```{r}
qplot(sample = logRainfall, facets = . ~ Treatment,
      data = case0301, geom = "qq") +
  geom_qq_line()
```

# Interpretation

## Posit a Model

- $Z_i$ = rainfall on unseeded days.
- $Y_i$ = **log** rainfall on unseeded days.
- $Z_i^*$ = rainfall on seeded days.
- $Y_i^*$ = **log** rainfall on seeded days.
- $Y_i^* = Y_i + \delta$.
- $Z_i^* = e^{\delta}Z_i$.
- $e^{\delta}$ is the multiplicative effect of seeding on rainfall.
- $e^{\delta} = 2$ means rainfall is twice as large on seeded days.
- $e^{\delta} = 3$ means rainfall is three times as large on seeded days.

## Posit Hypotheses

- $H_0: \delta = 0$
- $H_a: \delta \neq 0$

## Run t-test

```{r}
tout <- t.test(logRainfall ~ Treatment, data = case0301)
tout
```

## Estimate and Confidence Intervals on Original Scale

```{r}
exp(tout$estimate[1] - tout$estimate[2])
exp(tout$conf.int)
```


## Conclusion

- We estimate that that seeding results in a 3.1 factor increase in rainfall ($p$-value 0.01, 95\% confidence interval of 1.3 to 7.7).
- Note the causal language because this is a randomized experiment. I will deduct many points if you use causal language in an observational study.
