---
title: "Chapter 3 Worksheet"
author: "David Gerard"
date: today
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Case Study 3.1.2: Effects of Agent Orange on Troops in Vietnam

Many Vietnam veterans are concerned that their health may have been affected by
exposure to Agent Orange, a herbicide sprayed in South Vietnam between 1962 and
1970. The particularly worrisome component of Agent Orange is a dioxin called
TCDD, which in high doses is known to be associated with certain cancers.
Studies have shown that high levels of this dioxin can be detected 20 or more
years after heavy exposure. Consequently, researchers from the Centers for
Disease Control compared dioxin levels in Vietnam veterans to those in veterans
who did not serve in Vietnam.

The 646 Vietnam veterans in the study were a sample of U.S. Army combat
personnel who served in Vietnam during 1967 and 1968, in areas most heavily
treated with Agent Orange. The 97 non-Vietnam veterans entered the Army between
1965 and 1971 and served only in the United States or Germany. Neither sample
was randomly selected.

Blood samples from each veteran were analyzed for the presence of dioxin.

```{r}
#| eval: true
#| message: false
library(tidyverse)
case0302 <- read_csv("https://dcgerard.github.io/stat_302/data/case0302.csv")
glimpse(case0302)
```

**1.** From the description, do you think there are any problems with
independence between observations? Explain your reasoning.

**2.** Make side-by-side boxplots of `Dioxin` by `Veteran` group. Do the
variances look approximately equal?

```{r}

```

**3.** Based on the same plot, how does the symmetry assumption look?

**4.** Are there any major outliers? Reference the plot from Question 2.

**5.** Address any issues identified above and run a two-sample $t$-test.
Follow these steps:

  a. Posit a model. Let $Y_i$ denote the dioxin level for non-Vietnam veteran
     $i$ and $X_j$ denote the dioxin level for Vietnam veteran $j$. Write the
     location-shift model and define all symbols.

  b. State the null and alternative hypotheses in symbols.

  c. Run the $t$-test on the full dataset and report the $p$-value.

```{r}

```

  d. Identify any extreme outliers and remove them, then re-run the $t$-test.

```{r}

```

  e. Report your conclusions in plain language. Are the two analyses
     consistent? What do you conclude about dioxin levels in Vietnam vs.
     non-Vietnam veterans?

---

# QQ Plots

```{r}
#| eval: true
#| message: false
library(tidyverse)
case0301 <- read_csv("https://dcgerard.github.io/stat_302/data/case0301.csv")
case0201 <- read_csv("https://dcgerard.github.io/stat_302/data/case0201.csv")
```

**6.** Make a QQ plot of `Rainfall` for the `Unseeded` group in `case0301`.

```{r}

```

**7.** Based on the QQ plot from Question 6, what type of departure from
normality do you observe (right skew, left skew, heavy tails, light tails, or
outliers)? Describe how the shape of the plotted points supports your
conclusion.

**8.** Make a QQ plot of beak `Depth` for finches measured in 1978 from
`case0201`.

```{r}

```

**9.** Does the 1978 beak depth data appear approximately normal? What
features of the QQ plot support your conclusion?

---

# Log Transformations — Rainfall Data

Researchers randomly assigned days to receive cloud seeding (`Seeded`) or no
seeding (`Unseeded`) and recorded the resulting rainfall (in acre-feet).

**10.** The notes list three conditions that suggest a log transformation is
appropriate. State all three and check whether each holds for the rainfall
data. (Use plots to support your answer.)

```{r}

```

**11.** Create a new variable `logRainfall = log(Rainfall)`. Make side-by-side
boxplots and QQ plots of `logRainfall` by `Treatment`.

```{r}

```

**12.** Does the log transformation improve normality and stabilize the
variance? What do you observe compared to Question 10?

**13.** Posit a model for the log-scale analysis. Let $Y_i = \log(\text{Rainfall}_i)$
for unseeded days and $Y_j^* = \log(\text{Rainfall}_j^*)$ for seeded days.
Write the location-shift model and state the null and alternative hypotheses.

**14.** Run the $t$-test on the log-transformed data. Report the estimated
$\hat{\delta}$, its 95% CI, and the $p$-value.

```{r}

```

**15.** Exponentiate $\hat{\delta}$ and the endpoints of the 95% CI to
interpret the results on the original scale. Write a complete conclusion,
including units and a statement about causality (is a causal claim warranted
here?).

---

# Conceptual Questions

**16.** The chapter lists three assumptions of the two-sample $t$-test in
decreasing order of importance. List them. Then explain *why* independence is
the most critical — what specifically goes wrong with the $p$-value and
confidence interval when independence is violated?

**17.** Explain the difference between a "conservative" and "liberal"
statistical method in terms of (a) confidence intervals and (b) $p$-values.
If a $t$-test is applied to severely right-skewed data with a small sample
size, which direction of error would you expect? Why is this hard to predict
in general?

**18.** In the Agent Orange study, neither sample was randomly selected.

  a. Does the lack of random sampling affect whether we can *generalize* the
     results to all Vietnam veterans? Explain.

  b. Does it affect whether we can draw *causal* conclusions? Explain.

**19.** The Central Limit Theorem (CLT) guarantees that $\bar{X}$ is
approximately normal for large $n$, regardless of the population distribution.
Describe a scenario where you would *not* want to rely on the CLT to justify a
$t$-test, even with $n = 50$. What would you do instead?

**20.** An outlier is detected in a two-sample dataset. A student removes it,
runs the $t$-test only on the cleaned data, and reports those results.

  a. What is the problem with this approach?

  b. What is the recommended procedure?

**21.** Suppose you log-transform data and run a $t$-test. The 95% CI for
$\delta$ (the difference in log-medians, Group A minus Group B) is
$(0.2,\; 1.1)$. Exponentiate the endpoints. What does this interval say about
the ratio of medians (Group A to Group B) on the original scale? Is there
strong evidence that Group A has higher values than Group B?
