---
title: "Non-parameteric Alternatives to $t$-tools"
author: "David Gerard"
date: today
bibliography: "../bib.bib"
---

```{r setup}
#| include: false
set.seed(6)
knitr::opts_chunk$set(echo       = TRUE,
                      fig.height = 3,
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(broom)
```

# Learning Objectives

- Recognize which non-parameteric method to use in one- and two-sample scenarios.
- Understand the mechanims of these non-parameteric methods.
- Implement these non-parameteric methods in R.

# Assumptions of $t$-Tools

- Assumptions in decreasing order of importance:
  1. Independence between observational units.
  2. Equal variance (if assuming $\sigma_1^2 = \sigma_2^2$).
  3. Normality (no severe skew or outliers).

- Solutions when violated:
1. Independence: Use multiple linear regression for cluster effects, or use serial correlation adjustment methods (Chapter 15).
  2. Equal variance: Use Welch's $t$-test (which I think you should always use anyway).
  3. Normality: This chapter.
     - Use a Wilcoxon rank sum test for an 2-sample $t$-test alternative.
     - Use a Wilcoxon signed-rank test for a paired $t$-test alternative.
     
# Case Study

- Traditional mathematical graphics have text separate from the graph. Some researchers think this makes learning harder by creating excess "cognitive load".

- They assigned 14 students to learn a new mathematical subject using a traditional graph like this ("conventional") and a different 14 students to learn using a graph that incorporates together text and the graph ("modified").

-   They then had the students complete math problems using the techniques they learned from the graphics. They measured how long it took each student to complete them.

    ```{r}
    #| message: false
    graph <- read_csv("https://dcgerard.github.io/stat_302/data/case0402.csv")
    glimpse(graph)
    ```
-   These data are highly skewed and non-normal
    ```{r}
    ggplot(graph, aes(x = Treatment, y = Time)) +
      geom_boxplot() +
      geom_jitter(width = 0.1)
    ```

- A more worrying issue is that there are 5 students in the conventional group that did not finish the math problems in the allotted time. 

- Their values are listed as "300", but all we know is that their true values are greater than 300 seconds. 

- We call such values "censored".

- $t$-methods do not work well with very skewed data, and they cannot work at all with censored data. We need a new approach.

# The Wilcoxon rank-sum test






